---
title: "Plastic responses in height at the population level"
author: "Juliette Archambeau"
date: last-modified
format: 
  html:
    toc: true
    toc-depth: 4
    code-fold: true
    page-layout: full
    number-sections: true
embed-resources: true
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

<style type="text/css">
body {
   font-size: 15px;
}
code.r{
  font-size: 11px;
}
pre {
  font-size: 11px
}

table {
  font-size: 11px
}
</style>

```{r setup, include=F}
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))
knitr::opts_chunk$set(cache=F)
options(width = 300)
library(knitr)
library(kableExtra)
library(tidyverse)
library(here)
library(magrittr)
library(corrplot)
library(readxl)
library(rstan)
library(tidybayes)
library(bayesplot)
library(ggplot2)
library(parallel)
library(tictoc)
library(latex2exp)
library(brms)
library(Reacnorm)
library(matrixStats)
library(posterior)
library(missMDA)
library(ggrepel)
library(cmdstanr)
library(cowplot)
library(ggExtra)

rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

theme_set(theme_bw(base_size = 20))

# my own function for building tables in reports
source(here("scripts/functions/kable_mydf.R"))
```

# Introduction

The goal of this report is to determine whether the populations have different plastic responses for height.

In @sec-ReacNormVarPart, we applied the approach developed by Villemereuil & Chevin (2025) to assess whether the heritability of the reaction norm ($h_{RN}^2$) primarily stems from the environment-blind heritability of the trait ($h^2$, based on breeding values averaged across environments) or from the heritability of plasticity ($h_I^2$, based on breeding values by environment interaction). We implemented this approach at the species level, rather than at the population level, because it requires a large amount of data. We implemented the "animal model" version of the Villemereuil and Chevin approach, i.e. the version that uses a genomic relationship matrix (GRM) to estimate additive genetic effects. These models take a very long time to run because they require many iterations for the Markov chains to converge and careful parameter tuning.


In @sec-FamReactionNormEstim, we run a standard reaction-norm model by estimating family-level reaction-norms and we then extract two variables of interest: the median reaction-norm slope across families within a population (quantifies the average change in height between sites for that population), and the median variance of family-specific slopes within a population (wich measures the within-population variation in responses to the site contrast). These two parameters summarize, respectively, the population-level environmental response and the heterogeneity among families in that response.

In @sec-MeanTraitValues, we simply quickly extract the mean trait values (sign-inverted and scaled PC1 and scaled PC2) of each population in each field site to compare their values to those of the parameters of interest.

In @sec-RDPI, we quantified phenotypic plasticity using the relative distance plasticity index (RDPI; Valladares et al., 2006), which measures the mean relative phenotypic distance among individuals of the same population grown in different environments. We calculated RDPI in two ways: (i) at the individual level (i.e., the common practice), where all pairwise phenotypic distances between individuals of a given population and from different common gardens are considered, and (ii) using population means per site, which summarizes differences in average population responses across sites.

In @sec-IndexRelationships, we estimated the relationships among all these indice using a principal component analysis (PCA) and a correlation matrix.

In @sec-CorrWithGenomicOffset, we estimated the relationships between our phenology indexes with the predicted genomic offsets. We used genomic offset predictions from the RCP scenario 2.6, the prediction time period 2030-2050 and the average across the four climatic models (ensemble members).


# Heigh data

## loading the data

We load the phenotypic data from the common gardens and filter for either the common garden FS combined with the NG nursery, or the common garden FE and the nursery NE.

```{r LoadHeightData}
data <- readRDS(file=here("data/ScotsPine/formatted_CG_data.rds")) %>% # 1848 obs
  filter(FieldSite %in% c("FE","FS")) %>% # 1344 obs
  filter(!(FieldSite=="FE" & Nursery =="NG")) %>% # 1340 obs
  dplyr::select(-c(starts_with("BT"), starts_with("BD")))
```

## Removing families in only one field site

We remove families that are found in only one field site.

```{r RemovingFamiliesInOneSiteOnly}
# Count how many FieldSites each family occurs in
family_sites <- data %>%
  group_by(PopulationCode, Family) %>%
  summarise(n_sites = n_distinct(FieldSite), .groups = "drop")

# Families that occur in only one FieldSite
single_site_families <- family_sites %>%
  filter(n_sites == 1)

single_site_families %>% kable_mydf

# Remove families present in only one FieldSite
data <- data %>%
  group_by(PopulationCode, Family) %>%
  filter(n_distinct(FieldSite) > 1) %>%
  ungroup()
```


## Removing missing values

We first identify the number of missing values for each trait:

```{r CountNAs}
subdata <- data %>% dplyr::select(FieldSite,Block,FieldCode,PopulationCode,Family,Nursery,starts_with("HA")) 
  
print(sapply(subdata, function(x) sum(is.na(x))))
```

We are going to use height data from 2020. We remove rows with NAs for height in 2020.

We also rescale height with a min–max transformation to obtain strictly positive values on a [0,1] scale

```{r removingNAs}
subdata <- data %>% 
  select(FieldSite,Block,FieldCode,PopulationCode,Family,Nursery,HA20) %>% 
  drop_na() %>% 
  mutate(HA20sc = (HA20 - min(HA20)) / (max(HA20) - min(HA20)))
```


The final dataset contains `r length(unique(data$FieldCode))` individuals, `r length(unique(data$PopulationCode))` populations, `r length(unique(data$Family))` families. 

```{r}
subdata %>%  
  ggplot(aes(x=HA20)) +  
  geom_histogram(colour="blue",fill="lightblue",bins = 100) +
  #geom_density(alpha=.2,fill="pink") +
  xlab("Height in 2020") + ylab("") +
  theme_bw() 
```


# Variance partitioning with GRM {#sec-ReacNormVarPart}

To redo.

# Family-level reaction norms{#sec-FamReactionNormEstim}

## Running the model

We use a Bayesian hierarchical model to determine whether `HA20sc` differ between the two field sites, and how much this difference varies among populations and families.

`0 + FieldSite` in the model formula means that the model estimates a separate mean for each site rather than choosing one site as a reference. So:

  - `FieldSiteFE` = average `HA20sc` in field site FE
  - `FieldSiteFS` = average `HA20sc` in field site FS
  
The hierarchical structure (`0 + FieldSite | PopulationCode/Family`) then allows each population and each family within population to have its own value at each site.

```{r RunTheSpeciesLevelModelWithoutGRMNested, eval=F}
n_chains <- 4  # number of independant chains (MCMC)
n_iter <- 4000 # number of iterations
n_warm <- 1500 # number of iterations in the warm-up phase (which are then discarded)
n_thin <- 1    # thinning interval
n_seed <- 404

form_1 <- brmsformula(HA20sc ~ 0 + FieldSite + (0 + FieldSite | PopulationCode/Family))

mod <- brm(formula = form_1,
           data = subdata,
           #save_pars = save_pars(group = FALSE),
           chains = n_chains,
           cores = n_chains,
           seed = n_seed,
           iter = n_iter,
           backend = "cmdstanr",
           warmup = n_warm)

saveRDS(mod, file = here(paste0("outputs/ScotsPine/StanModels/HA20sc_TwoCGs_MixedModelWithoutGRM.rds")))
```


## Model summary

```{r MixedModelSummaryHA20, fig.height=8, fig.width=10, eval=T}
mod_HA20sc <- readRDS(file = here(paste0("outputs/ScotsPine/StanModels/HA20sc_TwoCGs_MixedModelWithoutGRM.rds")))
summary(mod_HA20sc)
#plot(mod_HA20sc, variable = c("^b_", "sigma"), regex = TRUE)
```

**Summary interpretation**

`HA20sc` was substantially higher in FS than in FE, with estimated means of 0.65 (95% credible interval: 0.63–0.67) and 0.39 (0.37–0.41), respectively. 


In the summary, the `sd(FieldSiteFE)` and `sd(FieldSiteFS)` parameters describe how much populations or families vary around the overall site mean. 

There was variation among populations, with a standard deviation of approximately 0.04 in both environments. This means that there is some population-level variation, but it is much smaller than the difference between sites (site difference ~ 0.26 units while population variation ~ 0.04 units around the overall site mean).

Population effects were strongly correlated between environments (r = 0.76), indicating that populations tended to maintain relatively similar `HA20sc` values across sites (`cor(FieldSiteFE,FieldSiteFS)` = 0.76 in the summary). 
Family-level variation was smaller in FE (SD = 0.02) than in FS (SD = 0.06), although the estimate for FE was relatively uncertain. The correlation of 0.56 indicates that families with relatively high `HA20sc` in one environment tend to be somewhat higher in the other environment, but the relationship is considerably weaker than at the population level.

And we have to be cautious about interpreting the FE family SD too strongly because its credible interval is 0.00 to 0.04. So the model is quite uncertain about the magnitude of family variation in FE.

sigma ~ 12 is the residual standard deviation = variation that remains within families, after accounting for field site, population and family. So there is still considerable individual-level variation around the predicted family/site values. The residual SD of 0.12 is considerably larger than the population-level SDs (0.04) and family-level SD in FE (0.02), and larger than the family-level SD in FS (0.06). This indicates that a substantial amount of variation occurs among observations within families, rather than being explained by population or family differences.


`Rhat` assesses whether the four MCMC chains have converged to the same posterior distribution. Here `Rhat` values close to 1, so there is no obvious indication of convergence problems.

`Bulk_ESS` and `Tail_ESS` are measures of how much independent information you effectively obtained from the MCMC sampling. Here we have generally high ESS values, so the posterior estimates are based on a good amount of effective sampling.

The relatively low ESS for the family correlation (`Bulk_ESS` = 277, `Tail_ESS` = 231) is worth noting, but it does not necessarily invalidate the model. Combined with `Rhat` = 1.01, it means that this particular parameter is estimated less precisely than most others.

Overall the model diagnostics indicated good convergence.

## Interval plots of family-level reaction norms

The graphs below show the family-level reaction norms across the two field sites. For each family, the reaction norm is quantified as the difference in predicted trait values between FS and FE (FS − FE). For each family, the point represents the median posterior estimate of this difference, while the horizontal error bar represents its 95% credible interval.

Families are ordered along the y-axis according to their median posterior difference. The dashed vertical line at zero indicates no difference in predicted trait values between the two field sites. Families with values greater than zero have higher predicted trait values in FS than in FE, whereas families with values below zero have higher predicted trait values in FE than in FS. Credible intervals that do not overlap zero provide strong posterior evidence that the predicted trait value differs between the two environments for that family.


```{r IntervalPlotFamilyLevelReactionNormNested, fig.height=15, fig.width=8}

## Interval plots of family-level reaction norms
##
## For each family, the reaction norm is quantified as the difference
## in predicted trait values between the two field sites:
## FS - FE.
##
## The plot shows the posterior median of this difference for each family,
## together with its 95% credible interval. Families are ordered according
## to their median posterior difference.
##
## Positive values indicate higher predicted trait values in FS,
## whereas negative values indicate higher predicted trait values in FE.
## The vertical line at zero represents no difference between environments.

interval_plot_family_reaction_norms_nested <- function(mod, y_label){

  family_diff <- mod %>%
    
    # Extract posterior draws of the family-level effects for each field site.
    # PopFam identifies the nested PopulationCode:Family combination.
    spread_draws(`r_PopulationCode:Family`[PopFam, FieldSite]) %>% 
    
    # Separate the nested PopulationCode:Family identifier into
    # population and family components.
    mutate(
      PopulationCode = sub("_.*", "", PopFam),
      Family         = sub(".*_", "", PopFam)
    ) %>%
    
    # Rename the extracted family-level effect for clarity.
    dplyr::rename(slope = `r_PopulationCode:Family`) %>%
    
    # Put the two field-site effects into separate columns,
    # allowing the FS - FE contrast to be calculated for each
    # posterior draw.
    pivot_wider(names_from = FieldSite, values_from = slope) %>%
    
    # Calculate the family-specific difference in predicted trait
    # values between the two environments for every posterior draw.
    mutate(
      slope_FS_FE = FieldSiteFS - FieldSiteFE
    ) %>%
    
    # Summarise the posterior distribution of the FS - FE contrast
    # for each family.
    group_by(PopulationCode, Family) %>%
    summarise(
      median_slope = median(slope_FS_FE),
      lower_slope  = quantile(slope_FS_FE, 0.025),
      upper_slope  = quantile(slope_FS_FE, 0.975),
      .groups = "drop"
    ) %>%
    
    # Order families from the most negative to the most positive
    # posterior median reaction norm.
    arrange(median_slope) %>%
    mutate(Family = factor(Family, levels = Family))

  ggplot(family_diff, aes(x = median_slope, y = Family)) +
    
    # Posterior median of the family-specific FS - FE difference.
    geom_point() +
    
    # 95% credible interval around the posterior median.
    geom_errorbar(aes(xmin = lower_slope, xmax = upper_slope), width = 0.2) +
    
    # Zero represents no difference between FS and FE.
    geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
    theme_bw() +
    labs(
      y = "",
      x = "Family-level reaction norms (FS − FE) with 95% credible intervals",
      title = y_label
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}


interval_plot_family_reaction_norms_nested(mod = mod_HA20sc, y_label = "Height in 2020 (HA20sc)")
```

## Reaction-norm plots

### Observed relationships

We compute and visualize the **observed** (non-modeled) family-level reaction norms across the two field sites (FE and FS). More precisely, we directly use the **raw data** to calculate the median height per family and site, and we then generate two plots.

The first plot displays the family-level median reaction norms of all families together. The background points represent individual tree observations. The colored lines connect the median trait values per family across the two field sites. Colors correspond to populations. This plot provides an overall view of the range of family-level differentiation in reaction norms.

The second plot shows the same information but faceted by population, allowing clearer visualization of within-population variation in median slopes. Each panel corresponds to one population. Background points represent individual trees. Blue lines connect family median values across sites.


```{r ObservedReactionNormsPlot, fig.height=10, fig.width=10}
observed_family_reaction_norms <- function(mod, subdata, y_var, y_label) {
  
  # Compute family medians per FieldSite
  family_means <- subdata %>%
    group_by(Family, FieldSite) %>%
    summarise(median_y = median(.data[[y_var]], na.rm = TRUE), .groups = "drop") %>%
    mutate(PopulationCode = substr(as.character(Family), 1, 2)) 
  
  # reshape wide to compute slopes
  family_slopes <- family_means %>%
    pivot_wider(names_from = FieldSite, values_from = median_y) %>%
    mutate(slope = FS - FE) %>% # family-level slope = difference between the two sites
    select(Family, PopulationCode, slope)
  
  sum_pop <- family_slopes %>%
    group_by(PopulationCode) %>%
    summarise(
      obs_median_pop_slope = median(slope, na.rm = TRUE), # median of the family slopes
      obs_var_family_slope = var(slope, na.rm = TRUE),
      .groups = "drop")
  
  # Plots
  p_all_reaction_norms <- 
    ggplot() +
    # Points = individual trees, colored by family
    geom_point(data = subdata,
               aes(x = FieldSite, y = .data[[y_var]], colour = PopulationCode),
               position = position_jitter(width = 0.1, height = 0), alpha = 0.5) +
    # Lines = family median across sites
    geom_line(data = family_means,
              aes(x = FieldSite, y = median_y, group = Family, colour = PopulationCode),
              linewidth = 1) +
    # Aesthetic tuning
    theme_minimal(base_size = 14) +
    #theme(legend.position = "none") +
    labs(x = "Field site", y = y_label) +
    scale_colour_viridis_d(option = "turbo", name = "Populations")  # nice categorical palette
  
  
  
  p_reaction_norms_per_pop <- ggplot() +
  # raw data points
  geom_jitter(data = subdata,
              aes(x = FieldSite, y = .data[[y_var]]),
              width = 0.1, alpha = 0.4, show.legend = FALSE) +
  
  # median/mean predictions per family
  geom_line(data = family_means,
            aes(x = FieldSite, y = median_y, group = Family),
            linewidth = 0.8, color = "blue", alpha = 0.3) +

  # facet by population
  facet_wrap(~PopulationCode, scales = "fixed") +
  
  theme_bw() +
  labs(
    x = "Field Site",
    y = y_label,
    title = "Family-level observed reaction norms (median values)"
  ) +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12))
  
  
  return(list(sum_pop = sum_pop,
              p_all_reaction_norms = p_all_reaction_norms,
              p_reaction_norms_per_pop = p_reaction_norms_per_pop))
}

# Run the function for each trait
obs_HA20sc <- observed_family_reaction_norms(mod_HA20sc, subdata, "HA20sc", y_label = "Height in 2020 (HA20sc)")


# Join into one table
sum_pop <- obs_HA20sc$sum_pop %>%
  rename(obs_median_pop_slope_HA20sc = obs_median_pop_slope,
         obs_var_family_slope_HA20sc = obs_var_family_slope)

# Show the plots
obs_HA20sc$p_all_reaction_norms
obs_HA20sc$p_reaction_norms_per_pop
```

For each population, two empirical descriptors were extracted from the raw phenotypic data. First, the *observed median population slope* (`obs_median_pop_slope`) corresponds to the median of the family-level reaction norm slopes, where each family slope is calculated as the difference between its median phenotype at the two field sites (FE – FS). This metric reflects the typical magnitude and direction of population-level plasticity. Second, the *observed variance in family-level slopes* (`obs_var_family_slope`) is the variance of these family-specific slopes within each population and quantifies the heterogeneity of plastic responses among families.

### Fitted relationships

#### Family-level reaction-norms

In the plots below, family-level reaction norms were obtained by generating posterior expected values from the fitted *brms* models using the function `add_epred_draws()` from *tidybayes*. For each combination of *FieldSite × Population × Family*, we constructed a prediction grid and extracted posterior draws of the expected response (`.epred`), specifying `re_formula = ~(0 + FieldSite | PopulationCode/Family)` so that predictions include the fixed *FieldSite* effect, the population-level random slopes, and the family-level random slopes. The function returns the posterior distribution of the linear predictor after marginalising over residual variance.

The first plot show the family-level reaction norms of all families together. Background points correspond to the observed height values, colored by population.

The second plot is faceted by population, allowing visualization of within-population variation in family responses. Background points correspond to the observed height values. The red line represents the global median response across all families, providing a reference for the overall trend.


```{r ReactionNormPlotFamilyLevelNested, fig.width=10, fig.height=10}
family_reaction_norms <- function(mod, subdata, y_var, y_label){
  
  valid_combos <- subdata %>% distinct(PopulationCode, Family)
  newdata <- crossing(FieldSite = unique(subdata$FieldSite), valid_combos)
  
  # Posterior fitted values including BOTH random-effect levels
  preds <- add_epred_draws(
    mod,
    newdata = newdata,
    re_formula = ~(0 + FieldSite | PopulationCode/Family))
  
  # Summaries per family × fieldsite
  family_summary <- preds %>%
    group_by(PopulationCode, Family, FieldSite) %>%
    summarise(
      median_pred       = median(.epred),
      mean_pred         = mean(.epred),
      lower_quant_pred  = quantile(.epred, 0.025),
      upper_quant_pred  = quantile(.epred, 0.975),
      sd_pred           = sd(.epred),
      lower_sd_pred     = mean_pred - sd_pred,
      upper_sd_pred     = mean_pred + sd_pred,
      .groups = "drop"
    )
  
  # Global summary across all families
  global_summary <- preds %>%
    group_by(FieldSite) %>%
    summarise(
      median_all       = median(.epred),
      mean_all         = mean(.epred),
      lower_quant_all  = quantile(.epred, 0.025),
      upper_quant_all  = quantile(.epred, 0.975),
      sd_all           = sd(.epred),
      lower_sd_all     = mean_all - sd_all,
      upper_sd_all     = mean_all + sd_all,
      .groups = "drop"
    )
  
  ## ------------------------------
  ## Plot 1 — All families together
  ## ------------------------------
  
  plot_all_fams <- ggplot() +
    geom_jitter(
      data = subdata,
      aes(x = FieldSite, y = .data[[y_var]], color = PopulationCode),
      width = 0.1, alpha = 0.4, show.legend = FALSE
    ) +
    geom_ribbon(
      data = family_summary,
      aes(x = FieldSite, ymin = lower_quant_pred, ymax = upper_quant_pred, group = Family, fill = PopulationCode),
      alpha = 0.05, show.legend = FALSE
    ) +
    geom_line(
      data = family_summary,
      aes(x = FieldSite, y = median_pred, group = Family, color = PopulationCode),
      linewidth = 0.8
    ) +
    theme_bw() +
    labs(x = "Field Site", y = y_label,
         title = "Family-level reaction norms (median ± 95% CI)") +
    theme(axis.text = element_text(size = 12))
  
  
  ## ----------------------------------
  ## Plot 2 — Faceted by population
  ## ----------------------------------
  
  plot_facet_median_with_CI <- ggplot() +
    geom_jitter(
      data = subdata,
      aes(x = FieldSite, y = .data[[y_var]]),
      width = 0.1, alpha = 0.4
    ) +
    geom_ribbon(
      data = family_summary,
      aes(x = FieldSite, ymin = lower_quant_pred, ymax = upper_quant_pred, group = Family),
      alpha = 0.05, fill = "blue"
    ) +
    geom_line(
      data = family_summary,
      aes(x = FieldSite, y = median_pred, group = Family),
      linewidth = 0.8, color = "blue", alpha = 0.3
    ) +
    geom_ribbon(
      data = global_summary,
      aes(x = FieldSite, ymin = lower_quant_all, ymax = upper_quant_all),
      alpha = 0.2, fill = "red"
    ) +
    geom_line(
      data = global_summary,
      aes(x = FieldSite, y = median_all, group = 1),
      color = "red", linewidth = 1.2
    ) +
    facet_wrap(~PopulationCode, scales = "fixed") +
    theme_bw() +
    labs(
      x = NULL, y = y_label,
      title = "Family-level reaction norms by population (median ± 95% CI)"
    ) +
    theme(axis.text = element_text(size = 12))
  
  
  ## -----------------------------------------
  ## Plot 3 — Mean ± SD version (optional)
  ## -----------------------------------------
  
  plot_facet_mean_with_SD <- ggplot() +
    geom_jitter(
      data = subdata,
      aes(x = FieldSite, y = .data[[y_var]]),
      width = 0.1, alpha = 0.4
    ) +
    geom_ribbon(
      data = family_summary,
      aes(x = FieldSite, ymin = lower_sd_pred, ymax = upper_sd_pred, group = Family),
      alpha = 0.05, fill = "blue"
    ) +
    geom_line(
      data = family_summary,
      aes(x = FieldSite, y = mean_pred, group = Family),
      linewidth = 0.8, color = "blue", alpha = 0.3
    ) +
    geom_line(
      data = global_summary,
      aes(x = FieldSite, y = mean_all, group = 1),
      color = "red", linewidth = 1.2
    ) +
    facet_wrap(~PopulationCode, scales = "fixed") +
    theme_bw() +
    labs(
      x = NULL, y = y_label,
      title = "Family-level fitted reaction norms (mean ± SD)"
    ) +
    theme(axis.text = element_text(size = 12))
  
  
  list(
    family_summary = family_summary,
    plot_all_fams = plot_all_fams,
    plot_facet_median_with_CI = plot_facet_median_with_CI,
    plot_facet_mean_with_SD = plot_facet_mean_with_SD
  )
}


family_reaction_norms_HA20sc <- family_reaction_norms(mod=mod_HA20sc, subdata, y_var="HA20sc", y_label = "Height in 2020 (HA20sc)")

family_reaction_norms_HA20sc$plot_all_fams

family_reaction_norms_HA20sc$plot_facet_median_with_CI
```

#### Population-level reaction-norms


```{r ReactionNormPlotPopulationLevelNested, fig.width=8, fig.height=8}
population_reaction_norms <- function(mod, subdata, y_var, y_label) {
  
  # Unique population × field site combinations
  valid_pops <- subdata %>% distinct(PopulationCode)
  newdata <- crossing(FieldSite = unique(subdata$FieldSite), valid_pops)
  
  # Posterior fitted values including only population-level random effects
  preds <- add_epred_draws(
    mod,
    newdata = newdata,
    re_formula = ~(0 + FieldSite | PopulationCode)
  )
  
  # Summarise posterior per population × field site
  pop_summary <- preds %>%
    group_by(PopulationCode, FieldSite) %>%
    summarise(
      median_pred      = median(.epred),
      mean_pred        = mean(.epred),
      lower_quant_pred = quantile(.epred, 0.025),
      upper_quant_pred = quantile(.epred, 0.975),
      sd_pred          = sd(.epred),
      lower_sd_pred    = mean_pred - sd_pred,
      upper_sd_pred    = mean_pred + sd_pred,
      .groups = "drop"
    )
  
  ## ------------------------------
  ## Plot 1 — All populations together
  ## ------------------------------
  plot_all_pops <- ggplot() +
    geom_jitter(
      data = subdata,
      aes(x = FieldSite, y = .data[[y_var]], color = PopulationCode),
      width = 0.1, alpha = 0.4
    ) +
    geom_ribbon(
      data = pop_summary,
      aes(x = FieldSite, ymin = lower_quant_pred, ymax = upper_quant_pred, group = PopulationCode, fill = PopulationCode),
      alpha = 0.1, show.legend = FALSE
    ) +
    geom_line(
      data = pop_summary,
      aes(x = FieldSite, y = median_pred, group = PopulationCode, color = PopulationCode),
      linewidth = 1
    ) +
    theme_bw() +
    labs(
      x = "Field Site",
      y = y_label,
      title = "Population-level reaction norms (median ± 95% CI)"
    ) +
    theme(axis.text = element_text(size = 12))
  
  ## ------------------------------
  ## Plot 2 — Faceted by population
  ## ------------------------------
  plot_facet_median_with_CI <- ggplot() +
    geom_jitter(
      data = subdata,
      aes(x = FieldSite, y = .data[[y_var]]),
      width = 0.1, alpha = 0.4
    ) +
    geom_ribbon(
      data = pop_summary,
      aes(x = FieldSite, ymin = lower_quant_pred, ymax = upper_quant_pred),
      alpha = 0.2, fill = "blue"
    ) +
    geom_line(
      data = pop_summary,
      aes(x = FieldSite, y = median_pred, group = 1),
      color = "blue", linewidth = 1
    ) +
    facet_wrap(~PopulationCode, scales = "fixed") +
    theme_bw() +
    labs(
      x = "Field Site",
      y = y_label,
      title = "Population-level reaction norms by population (median ± 95% CI)"
    ) +
    theme(axis.text = element_text(size = 12))
  
  list(
    pop_summary = pop_summary,
    plot_all_pops = plot_all_pops
  )
}

# ------------------------------
# Apply function to HA20sc
# ------------------------------
pop_reaction_norms_HA20sc <- population_reaction_norms(mod = mod_HA20sc, subdata, y_var = "HA20sc", y_label = "Height in 2020 (HA20sc)")

# Display plots
pop_reaction_norms_HA20sc$plot_all_pops
```


#### Fitted parameters of interest

For each response variable (only `HA20sc` here), we used posterior expected values (`epred`) to compute, for every *Population × Family*, the difference in height between the two field sites (FS – FE). This difference represents the reaction-norm slope.

From these slopes, two population-level parameters were extracted:

  - `fitted_median_pop_slope`: The median reaction-norm slope across families within a population. Quantifies the average change in height between sites for that population.
  
  - `fitted_var_family_slope`: The median variance of family-specific slopes within a population. Measures the within-population variation in responses to the site contrast.

These two parameters summarize, respectively, the population-level environmental response and the heterogeneity among families in that response.

```{r ExtractParamsOfInterest}
extract_slopes_epred <- function(model, subdata) {
  
  # Get all population/family combinations
  newdata <- subdata %>% dplyr::select(FieldSite, PopulationCode, Family) %>% distinct

  # Posterior predictions including both population and family effects
  preds <- add_epred_draws(model, newdata = newdata, re_formula = ~(0 + FieldSite | PopulationCode/Family)) %>% 
    ungroup %>% 
    select(FieldSite, PopulationCode, Family, .draw, .epred)
  
  # Compute slope for each draw: FS - FE per family
  slopes <- preds %>%
    pivot_wider(
      names_from = FieldSite,
      values_from = .epred
    ) %>%
    mutate(
      fam_slope = FS -FE
    )
  
  # Summarize per draw & population
  slopes_by_draw <- slopes %>%
    group_by(.draw, PopulationCode) %>%
    arrange(PopulationCode) %>% 
    summarise(
      pop_slope = median(fam_slope),
      var_family_slope = if(n() > 1) var(fam_slope) else 0, # zero if only one family
      .groups = "drop"
    )
  
  # Summarize across draws
  summary_pop <- slopes_by_draw %>%
    group_by(PopulationCode) %>%
    summarise(
      fitted_median_pop_slope = median(pop_slope),
      fitted_var_family_slope = median(var_family_slope),
      .groups = "drop"
    )
  
  summary_pop
}

# Apply to both models
summary_HA20sc <- extract_slopes_epred(mod_HA20sc, subdata)

# Join into one table
sum_pop <- summary_HA20sc %>%
  rename(fitted_median_pop_slope_HA20sc = fitted_median_pop_slope,
         fitted_var_family_slope_HA20sc = fitted_var_family_slope) %>%
  left_join(sum_pop, by="PopulationCode")

sum_pop %>%  kable_mydf()
```


# Mean trait values{#sec-MeanTraitValues}

We extract the mean trait values of each population in each common garden.

```{r ExtractMeanTraitValue}
# Function to compute mean phenotype per population and field site
compute_means_per_site <- function(subdata, trait) {
  subdata %>%
    group_by(PopulationCode, FieldSite) %>%
    summarise(mean_y = mean(.data[[trait]], na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(
      names_from = FieldSite,
      values_from = mean_y,
      names_prefix = paste0("mean_", trait, "_")
    )
}

# Compute means for both traits
mean_HA20sc <- compute_means_per_site(subdata, "HA20sc")

# Merge with your existing summary table
sum_pop <- sum_pop %>%
  left_join(mean_HA20sc, by = "PopulationCode")
```

# RDPI{#sec-RDPI}


In this section, we calculate the relative distance plasticity index (RDPI) from @valladaresQuantitativeEstimationPhenotypic2006. In this paper, the authors propose to quantify phenotypic plasticity based on phenotypic distances among individuals of a given species (in our study, population) exposed to different environments (in our study, planted in different common gardens), which is summarized in the RDPI index that allows for statistical comparisons of phenotypic plasticity between species (in our study, populations). 

For a given trait, the relative distance plasticity index $RDPI_p$ for a given population $p$ can be calculated as follows:

$$RDPI_p = \frac{\sum_{n=1}^N{\frac{x_{ij}-x_{i'j'}}{x_{ij}+x_{i'j'}}}}{N}$$

where $N$ is the number of pairwise comparisons among individuals of the same population but from different common gardens. $x_{ij}$ and $x_{i'j'}$ are the phenotypic values of individual $j$ and $j'$ in common gardens $i$ and $i’$. As such, the phenotypic plasticity of a population for a given trait corresponds to the sum of the relative phenotypic distances for all pairs of individuals of the population grown in different common gardens. 

We also calculate the RDPI at the population level. In this case, we first compute the mean trait value of each population in each common garden, and then calculate RDPI based on these population means. The population-mean RDPI therefore summarizes overall population-level plasticity, integrating the average phenotypic shift of the population across environments while ignoring within-population variation among individuals.


```{r CalcRDPI}
## ---- RDPI computation for HA20sc  ----

compute_RDPI <- function(subdata, trait) {
  
  # Rename target trait to generic 'y'
  subdata <- subdata %>% mutate(y = .data[[trait]])
  
  # Identify populations and field sites
  pop_codes <- unique(subdata$PopulationCode) %>% sort()
  site_codes <- unique(subdata$FieldSite) %>% sort()
  site_comb <- combn(site_codes, 2)
  calc_rel_dist <- function(x) abs(x[1] - x[2]) / (x[1] + x[2])

  # Loop over populations
  res_list <- lapply(pop_codes, function(pop) {
    subpop <- subdata %>% filter(PopulationCode == pop)

    # --- Individual-based RDPI ---
    rel_dist_ind <- lapply(1:ncol(site_comb), function(nb_comb) {
      subsite1 <- subpop %>% filter(FieldSite == site_comb[1, nb_comb])
      subsite2 <- subpop %>% filter(FieldSite == site_comb[2, nb_comb])
      
      if (nrow(subsite1) == 0 | nrow(subsite2) == 0) return(NULL)
      
      lapply(1:nrow(subsite1), function(x) {
        rbind(subsite1[x, "y"], subsite2[, "y"]) %>%
          pull() %>%
          combn(2, calc_rel_dist) %>%
          .[1:nrow(subsite2)] %>%
          as.numeric()
      }) %>% unlist()
    }) %>% unlist()

    N_ind <- length(rel_dist_ind)
    ind_RDPI <- if (N_ind == 0) NA_real_ else sum(rel_dist_ind) / N_ind

    # --- Population-mean RDPI ---
    mean_tab <- subpop %>%
      drop_na(y) %>%
      group_by(FieldSite) %>%
      summarise(pop_mean = mean(y, na.rm = TRUE), .groups = "drop")

    rel_dist_pop_means <- lapply(1:ncol(site_comb), function(nb_comb) {
      s1 <- site_comb[1, nb_comb]
      s2 <- site_comb[2, nb_comb]
      mean1 <- mean_tab %>% filter(FieldSite == s1) %>% pull(pop_mean)
      mean2 <- mean_tab %>% filter(FieldSite == s2) %>% pull(pop_mean)
      if (length(mean1) == 0 || length(mean2) == 0) return(NA_real_)
      denom <- mean1 + mean2
      if (denom == 0) return(NA_real_) else return(abs(mean1 - mean2) / denom)
    }) %>% unlist()

    pop_mean_RDPI <- if (all(is.na(rel_dist_pop_means))) NA_real_ else mean(rel_dist_pop_means, na.rm = TRUE)

    tibble(PopulationCode = pop,
           ind_RDPI = ind_RDPI,
           pop_mean_RDPI = pop_mean_RDPI)
  })

  df_rdpi <- bind_rows(res_list) %>%
    mutate(across(c(ind_RDPI, pop_mean_RDPI), as.numeric))

  # Rename columns with trait suffix
  colnames(df_rdpi)[2:3] <- paste0(colnames(df_rdpi)[2:3], "_", trait)
  
  return(df_rdpi)
}

# ---- Apply function to both traits ----
df_rdpi_HA20sc <- compute_RDPI(subdata, "HA20sc")

# ---- Merge all results ----
sum_pop <- sum_pop %>%
  left_join(df_rdpi_HA20sc, by = "PopulationCode") 

# Final summary
sum_pop %>% kable_mydf()
```

# Relationships among all the estimated indexes{#sec-IndexRelationships}

Description for each variable:

  - `PopulationCode`: Identifier of the population.

  - `fitted_median_pop_slope_HA20sc`: Median reaction-norm slope of a population predicted by the model, computed from posterior draws and averaged across its families. Summarizing the predicted population-level difference in height between field sites.

  - `fitted_var_family_slope_HA20sc`: Model-based variance in family-level slopes within a population, quantifying predicted among-family heterogeneity in reaction norms (i.e., within-population variation in responses).
  
  - `obs_median_pop_slope_HA20sc`: Observed median reaction-norm slope (FS – FE) of a population, calculated directly from the empirical family slopes. Summarizing the observed population-level difference in height between the two field sites.
  
  - `obs_var_family_slope_HA20sc`: observed variance in family-level slopes within a population; an empirical measure of how differently families respond across sites.

  - `mean_FE_HA20sc`: Mean scaled-height value of the population in the FE site.

  - `mean_FS_HA20sc`: Mean scaled-height value of the population in the FS site.

  - `pop_mean_RDPI_HA20sc`: RDPI computed from population means, summarizing average phenotypic divergence across sites.
  
  - `ind_RDPI_HA20sc`: Individual-based RDPI, quantifying relative phenotypic change across sites among individuals within the population.


```{r RenameMeanTraitValues}
# Rename columns so FE/FS comes first
sum_pop <- sum_pop %>%
  rename(
    mean_FE_HA20sc = mean_HA20sc_FE,
    mean_FS_HA20sc = mean_HA20sc_FS)

sum_pop %>% kable_mydf
```

## PCA

```{r PCACalculatedIndexes, fig.height=8, fig.width=8}
# keep only populations with complete data for PCA (or use na.omit)
pca_df <- sum_pop %>% drop_na()

# run PCA (center & scale)
pca_res <- prcomp(pca_df %>% select(-PopulationCode), center = TRUE, scale. = TRUE)

# scores (PC coordinates for populations)
scores <- as.data.frame(pca_res$x)
scores$PopulationCode <- pca_df$PopulationCode

# loadings (variable contributions)
loadings <- as.data.frame(pca_res$rotation)
loadings$var <- rownames(loadings)

# get % variance explained for axis labels
eig <- (pca_res$sdev)^2
pct_var <- round(100 * eig / sum(eig), 1)

# scale loadings to make arrows visible on the same plotting scale as scores
# choose a multiplier so arrows reach ~30-50% of the score cloud
mult <- min(
  (max(scores$PC1) - min(scores$PC1)) / (max(loadings$PC1) - min(loadings$PC1)),
  (max(scores$PC2) - min(scores$PC2)) / (max(loadings$PC2) - min(loadings$PC2))
) * 0.6

loadings_plot <- loadings %>%
  mutate(arrow_x = PC1 * mult,
         arrow_y = PC2 * mult)

# create a color palette with one color per population
n_pops <- nrow(scores)
pal <- scales::hue_pal()(n_pops) # distinct hues; change palette if you prefer

# Plot
ggplot() +
  # points for populations
  geom_point(data = scores, aes(x = PC1, y = PC2, color = PopulationCode), size = 3, alpha = 0.9) +
  # population labels
  geom_text_repel(data = scores, aes(x = PC1, y = PC2, label = PopulationCode, color = PopulationCode),
                  size = 3.5, show.legend = FALSE, max.overlaps = 100) +
  # arrows for variable loadings (from origin to scaled loading)
  geom_segment(data = loadings_plot,
               aes(x = 0, y = 0, xend = arrow_x, yend = arrow_y),
               arrow = arrow(length = unit(0.25, "cm")), colour = "black", linewidth = 0.6) +
  # variable labels at arrow tips
  geom_text_repel(data = loadings_plot,
                  aes(x = arrow_x, y = arrow_y, label = var),
                  color = "black", size = 4, segment.size = 0.3) +
  scale_color_manual(values = pal) +
  labs(x = paste0("PC1 (", pct_var[1], "%)"),
       y = paste0("PC2 (", pct_var[2], "%)"),
       title = "",
       color = "Population") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none") # legend suppressed because each pop has its own color label

```

## Correlations 

```{r PCAonCalculatedPhenoIndexes, fig.height=10, fig.width=10}
sum_pop %>% 
  dplyr::select(-PopulationCode) %>% 
  cor() %>% 
  corrplot(method = 'number',
           type = 'lower', diag = FALSE,
           title=paste0(""),
           mar=c(0,0,2,0),
           number.cex=0.9,tl.cex=0.9)
```


## Box plots

```{r BoxPlots, fig.height=5, fig.width=10}
# Pivot to long format
sum_pop_long <- sum_pop %>%
  pivot_longer(
    cols = -PopulationCode,
    names_to = c("variable"),
    names_pattern = "(.*_FE|.*_FS|.*)_HA",
    values_to = "value"
  )

# --- ensure correct variable grouping ---
sum_pop_long$variable <- factor(sum_pop_long$variable,
                                levels = c("fitted_median_pop_slope",
                                           "fitted_var_family_slope",
                                           "obs_median_pop_slope",
                                           "obs_var_family_slope",
                                           "mean_FS",
                                           "mean_FE",
                                           "ind_RDPI",
                                           "pop_mean_RDPI"))


# Split long table into two sets of variables
pop_traits <- c("mean_FS","mean_FE")
pop_slopes <- c("ind_RDPI", "obs_median_pop_slope", "fitted_median_pop_slope", "pop_mean_RDPI")
pop_vars <- c("obs_var_family_slope", "fitted_var_family_slope")

df_traits <- sum_pop_long %>% filter(variable %in% pop_traits)
df_slopes <- sum_pop_long %>% filter(variable %in% pop_slopes)
df_vars <- sum_pop_long %>% filter(variable %in% pop_vars)

# Traits
ggplot(df_traits, aes(x = variable, y = value)) +
  geom_boxplot(alpha = 0.7, width = 0.6, outlier.shape = NA, color = "black", fill="lightblue") +
  geom_jitter(color = "#1f77b4", width = 0.15, alpha = 0.5, size = 2, show.legend = FALSE) +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.minor = element_blank()) +
  labs(x = "", y = "Index value", title = "Population mean trait values in each field site")

ggplot(df_slopes, aes(x = variable, y = value)) +
  geom_boxplot(alpha = 0.7, width = 0.6, outlier.shape = NA, color = "black",  fill="lightblue") +
  geom_jitter(color = "#1f77b4", width = 0.15, alpha = 0.5, size = 2, show.legend = FALSE) +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.minor = element_blank()) +
  labs(x = "", y = "Index value", title = "Population median plastic responses")

ggplot(df_vars, aes(x = variable, y = value)) +
  geom_boxplot(alpha = 0.7, width = 0.6, outlier.shape = NA, color = "black",  fill="lightblue") +
  geom_jitter(color = "#1f77b4", width = 0.15, alpha = 0.5, size = 2, show.legend = FALSE) +
  theme_bw(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "", y = "Value", title = "Fitted and observed variance of the family-level slopes")
```


## Scatter plots among indexes

Some scatter plots.

```{r ScatterPlotsAmongIndexes, fig.height=7, fig.width=9}
pop_n <- subdata %>% dplyr::count(PopulationCode, name = "n_pop")
df_diag <- sum_pop %>%  left_join(pop_n, by = "PopulationCode")

# Variables to compare
vars <- c(
  "fitted_median_pop_slope_HA20sc",
  "obs_median_pop_slope_HA20sc",
  "fitted_var_family_slope_HA20sc",
  "obs_var_family_slope_HA20sc",
  "pop_mean_RDPI_HA20sc",
  "ind_RDPI_HA20sc"
)

# All pairwise combinations
pairs <- t(combn(vars, 2)) %>% as.data.frame()
colnames(pairs) <- c("xvar", "yvar")


# Function to build one plot
make_pairplot <- function(xvar, yvar, df) {
  
  ggplot(df, aes(x = .data[[xvar]], y = .data[[yvar]])) +
    
    geom_point(
      aes(size = n_pop),
      col="blue",
      alpha = 0.7
    ) +
    
    geom_text(
      aes(label = PopulationCode),
      vjust = 1,hjust = -0.5,
      size = 3
    ) +
    
    geom_smooth(
      method = "lm",
      se = FALSE,
      formula = y ~ x,
      linewidth = 1
    ) +
    
    scale_size_continuous(name = "Sample size") +
    
    labs(
      x = xvar,
      y = yvar,
      title = paste(yvar, "vs", xvar)
    ) +
    theme_bw(base_size = 14)
}

# Build all plots
pairplots <- pmap(
  pairs,
  ~ make_pairplot(..1, ..2, df_diag)
)

# Name them
names(pairplots) <- paste0(pairs$yvar, "_vs_", pairs$xvar)

# Display some plots
pairplots$fitted_var_family_slope_HA20sc_vs_fitted_median_pop_slope_HA20sc
pairplots$obs_var_family_slope_HA20sc_vs_obs_median_pop_slope_HA20sc
pairplots$obs_var_family_slope_HA20sc_vs_fitted_var_family_slope_HA20sc
pairplots$obs_median_pop_slope_HA20sc_vs_fitted_median_pop_slope_HA20sc
pairplots$ind_RDPI_HA20sc_vs_pop_mean_RDPI_HA20sc
pairplots$ind_RDPI_HA20sc_vs_fitted_var_family_slope_HA20sc
pairplots$ind_RDPI_HA20sc_vs_fitted_median_pop_slope_HA20sc 
```


# Correlations with genomic offset predictions{#sec-CorrWithGenomicOffset}

```{r DownloadGOpred}
df_go_clim_dist <- readRDS(here("outputs/ScotsPine/df_gen_off_pred_clim_dist.rds"))

df_go_wide <- df_go_clim_dist %>%
  select(PopulationCode, rcp_scenario, pred_timeslice, method, ensemble_member, var_code,val) %>%
  mutate(var_name_comb = paste(rcp_scenario, pred_timeslice, method, ensemble_member, var_code,sep = "_") %>%  gsub("[ .-]", "", .)) %>%
  dplyr::select(PopulationCode, var_name_comb,val) %>% 
  pivot_wider(names_from = var_name_comb, values_from = val)


df_go_wide <- df_go_clim_dist %>%
  dplyr::filter(var_code == "all_snps" & rcp_scenario == "RCP2.6" & pred_timeslice == "2030-2050" & ensemble_member == "Average across EMs") %>% 
  select(PopulationCode, method, val) %>%
  mutate(var_name_comb = paste(method, sep = "_") %>%  gsub("[ .-]", "", .)) %>%
  dplyr::select(PopulationCode, var_name_comb, val) %>% 
  pivot_wider(names_from = var_name_comb, values_from = val)
```

## PCA

```{r PCAwithGO, fig.height=8, fig.width=8}
# keep only populations with complete data for PCA (or use na.omit)
pca_df <- sum_pop %>% left_join(df_go_wide, by="PopulationCode")

# run PCA (center & scale)
pca_res <- prcomp(pca_df %>% select(-PopulationCode), center = TRUE, scale. = TRUE)

# scores (PC coordinates for populations)
scores <- as.data.frame(pca_res$x)
scores$PopulationCode <- pca_df$PopulationCode

# loadings (variable contributions)
loadings <- as.data.frame(pca_res$rotation)
loadings$var <- rownames(loadings)

# get % variance explained for axis labels
eig <- (pca_res$sdev)^2
pct_var <- round(100 * eig / sum(eig), 1)

# scale loadings to make arrows visible on the same plotting scale as scores
# choose a multiplier so arrows reach ~30-50% of the score cloud
mult <- min(
  (max(scores$PC1) - min(scores$PC1)) / (max(loadings$PC1) - min(loadings$PC1)),
  (max(scores$PC2) - min(scores$PC2)) / (max(loadings$PC2) - min(loadings$PC2))
) * 0.6

loadings_plot <- loadings %>%
  mutate(arrow_x = PC1 * mult,
         arrow_y = PC2 * mult)

# create a color palette with one color per population
n_pops <- nrow(scores)
pal <- scales::hue_pal()(n_pops) # distinct hues; change palette if you prefer

# Plot
ggplot() +
  # points for populations
  geom_point(data = scores, aes(x = PC1, y = PC2, color = PopulationCode), size = 3, alpha = 0.9) +
  # population labels
  geom_text_repel(data = scores, aes(x = PC1, y = PC2, label = PopulationCode, color = PopulationCode),
                  size = 3.5, show.legend = FALSE, max.overlaps = Inf) +
  # arrows for variable loadings (from origin to scaled loading)
  geom_segment(data = loadings_plot,
               aes(x = 0, y = 0, xend = arrow_x, yend = arrow_y),
               arrow = arrow(length = unit(0.25, "cm")), colour = "black", linewidth = 0.6) +
  # variable labels at arrow tips
  geom_text_repel(data = loadings_plot,
                  aes(x = arrow_x, y = arrow_y, label = var),
                  color = "black", size = 4, segment.size = 0.3, max.overlaps = Inf) +
  scale_color_manual(values = pal) +
  labs(x = paste0("PC1 (", pct_var[1], "%)"),
       y = paste0("PC2 (", pct_var[2], "%)"),
       title = "",
       color = "Population") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none") # legend suppressed because each pop has its own color label

```
 
 
## Correlations

```{r CorrelationsWithGO, fig.height=12, fig.width=12}
sum_pop %>% left_join(df_go_wide, by="PopulationCode") %>% 
  dplyr::select(-PopulationCode) %>% 
  cor() %>% 
  corrplot(method = 'number',
           type = 'lower', diag = FALSE,
           title=paste0(""),
           mar=c(0,0,2,0),
           number.cex=0.9,tl.cex=0.9)
```

## Scatter plots with RDA predictions

```{r PlasticityIndexesVsRDApredictionsScatterPlots, fig.height=7, fig.width=9}
df_diag <- df_diag %>% left_join(df_go_wide[,c("PopulationCode", "RDA")],by = "PopulationCode")

# Build a tibble of pairs: each xvar paired with RDA
pairs <- tibble(
  xvar = vars,
  yvar = "RDA"
)

# Function to build one plot
make_RDA_plot <- function(xvar, yvar, df) {
  
  ggplot(df, aes(x = .data[[xvar]], y = .data[[yvar]])) +
    
  geom_point(
    aes(size = n_pop),
    col="blue",
    alpha = 0.7
  ) +
    
    geom_text(
      aes(label = PopulationCode),
      vjust = 1,hjust = -0.5,
      size = 3
    ) +
    
    geom_smooth(
      method = "lm",
      formula = y ~ x,
      se = FALSE,
      linewidth = 1
    ) +
    scale_size_continuous(name = "Sample size") +
    
    labs(
      x = xvar,
      y = yvar,
      title = paste(yvar, "vs", xvar)
    ) +
    
    theme_bw(base_size = 14)
}

# Build all plots
RDA_plots <- pmap(
  pairs,
  ~ make_RDA_plot(..1, ..2, df_diag)
)

# Name the plots
names(RDA_plots) <- paste0("RDA_vs_", vars)

# Example: display one plot
for(p in RDA_plots) print(p)
```


