---
title: "Estimating plasticity at the population level"
subtitle: "Villemereuil & Chevin method & RDPI of Valladeres"
author: "Juliette Archambeau"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  html:
    toc: true
    toc-depth: 4
    code-fold: true
    page-layout: full
    number-sections: true
embed-resources: true
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

<style type="text/css">
body {
   font-size: 15px;
}
code.r{
  font-size: 11px;
}
pre {
  font-size: 11px
}

table {
  font-size: 11px
}
</style>

```{r setup, include=F}
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))
knitr::opts_chunk$set(cache=F)
options(width = 300)
library(knitr)      # CRAN v1.26
library(kableExtra) # CRAN v1.1.0
library(tidyverse)
library(here)
library(magrittr)
library(corrplot)
library(readxl)
library(rstan)
library(tidybayes)
library(bayesplot)
library(ggplot2)
library(parallel)
library(tictoc)
library(latex2exp)
library(brms)
library(Reacnorm)
library(matrixStats)
library(posterior)

rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

theme_set(theme_bw(base_size = 20))

# my own function for building tables in reports
source(here("scripts/functions/kable_mydf.R"))
source(here("scripts/functions/make_spatialpoints_map.R"))
```

# Introduction

The goal of this report is to determine whether the populations have different plastic responses. For what, we use different approaches presented below.

The focus trait of the present report is (except in @sec-RDPI and @sec-RawDiff which are applied to many traits):

```{r FocusTrait}
trait <- "HA20"
```

To account for the nursery effects, we estimate plasticity to environment variation originating from the combined effect of the site and the nursery. For that, in the present report, we only use the two common gardens in which the trees come from a unique nursery, i.e. FS (in which all trees come from NG) and FE (in which all trees come from NE, except four trees that come from NG and that we remove to estimate plasticity).


In @sec-SPmodelGRM, we partitioned the variance of the trait of interest at the species level, i.e., with all populations combined. For that, we follow the methodology in [de Villemereuil & Chevin (2025)](https://doi.org/10.24072/pcjournal.515) to partition the phenotypic variance of the traits of interest. Below are sections extracted from the paper:

$$V_P =  V_{Plas} + V_{Gen} + V_{Res}$$

$V_P$ is the total phenotypic variance of the trait.
$V_{Plas}$ is the phenotypic variance arising from changes in the mean reaction norm across environments.
$V_{Gen}$ is the total genetic variance in the trait across environments.
$V_{Res}$ is the residual variance, not explained by the reaction norm.

$$V_{Gen} = V_{Add} + V_{nonAdd}$$

$V_{Add}$ is the total additive genetic variance in the trait across environments.
$V_{nonAdd}$ is the non additive genetic variance in the trait.

$$V_{Add} =  V_A + V_{AxE}$$

$V_A$ is the environment-blind additive genetic variance of the trait, i.e., based on the mean breeding values across environments.
$V_{AxE}$ is the additive genetic variance arising from plasticity, i.e., variance of the mean-centered breeding values.

From this, it is possible to derive unitless quantities of interest.

$P_{RN}^2 = V_{Plas}/V_P$ is the proportion of the phenotypic variance arising from average plastic responses to environments.

Variance-standardised additive genetic variances are heritabilities:

$h_{RN}^2=\frac{V_{Add}}{V_P}=\frac{V_A}{V_P}+\frac{V_{AxE}}{V_P}=h^2+h^2_I$

In other words, the heritability of the trait when fully accounting for its reaction norm ($h_{RN}^2$) is equal to the environment-blind heritability of the trait ($h^2$, based on the breeding values averaged across environments) plus the heritability from plasticity ($h_I^2$, based on the breeding values by environment interaction).

$T_{RN}^2=\frac{V_{Plas}+V_{Gen}}{V_P}=P_{RN}^2+h_{RN}^2$

$T_{RN}^2$ is the proportion of the phenotypic variance explained by the (possibly plastic and genetically variable) reaction norm, and thus our ability to predict the individual phenotype from the genotype and the environment.

To apply this methodology, we followed the tutorials associated with the `Reacnorm` R package: <https://github.com/devillemereuil/Reacnorm>.

In @sec-PopmodelGRM, we attempted to partition the variance of the trait of interest for each population individually. However, it appeared that the sample sizes were likely too small to allow the models to converge.

In @sec-SPmodelPopVaryingSlopes, we estimated population varying slopes (i.e., random slopes) to quantify the mean change in the trait of interest across two different sites (Yair and Glensaugh) for each population. This approach (if I have understood correctly) would be similar to estimate $V_{Plas}$ for each population. While this estimate would not indicate the genetic variation in reaction norms within populations, it would allow for comparisons of the mean plastic responses across sites among populations.

In @sec-RDPI, we calculate the relative distance plasticity index (RDPI) from @valladaresQuantitativeEstimationPhenotypic2006 In this paper, the authors propose to quantify phenotypic plasticity based on phenotypic distances among individuals of a given species (in our study, population) exposed to different environments (in our study, planted in different common gardens), which is summarized in the RDPI index that allows for statistical comparisons of phenotypic plasticity between species (in our study, populations). 

For a given trait, the relative distance plasticity index $RDPI_p$ for a given population $p$ can be calculated as follows:

$$RDPI_p = \frac{\sum_{n=1}^N{\frac{x_{ij}-x_{i'j'}}{x_{ij}+x_{i'j'}}}}{N}$$

where $N$ is the number of pairwise comparisons among individuals of the same population but from different common gardens. $x_{ij}$ and $x_{i'j'}$ are the phenotypic values of individual $j$ and $j'$ in common gardens $i$ and $i’$. As such, the phenotypic plasticity of a population for a given trait corresponds to the sum of the relative phenotypic distances for all pairs of individuals of the population grown in different common gardens. We calculate the RDPI for height measurements, the duration of bud burst and the number of days to reach of stage 4, 5 or 6 of budburst.

<span style="color: orange;">**Comment:**</span> Note that RDPI may also be calculated at the family level (instead of the population level as above) and then be averaged for each population. However, as some families in FS are not the as in FE and FW, I thought that calculating the RDPI indexes directly at the population level would allow to consider all families (not having to remove some). But this is point I will be happy to discuss! 

In @sec-RawDiff, we calculate the difference in mean trait values between the two studied sites for each population.

Last, we compared the correlations among population slopes obtained in @sec-SPmodelPopVaryingSlopes, RDPI calculated in @sec-RDPI and the raw differences calculated in @sec-RawDiff.



# The data

We load the phenotypic data from the common gardens.

```{r LoadPhenoCGData}
data <- readRDS(file=here("data/ScotsPine/formatted_CG_data.rds"))
```

The dataset contains `r length(unique(data$FieldCode))` individuals, `r length(unique(data$PopulationCode))` populations, `r length(unique(data$Family))` families. Here are the first 10 rows:

```{r}
data[1:10,] %>% kable_mydf()
```

We load the genomic relationship matrix.

```{r LoadGRM}
GRM <- readRDS(file=here("data/ScotsPine/GenomicData/data_update_Fev2025/formatted_data/Gmat_VanRaden.rds"))
GRM[1:10,1:10]
```

Matrix dimensions:

```{r}
dim(GRM)
```


# Species level model with GRM{#sec-SPmodelGRM}

We filtered the phenotypic data to retain only the trait of interest and measurements from the Glensaugh (FE) and Yair (FS) sites. Additionally, we excluded the four trees measured in FE that originated from the NG nursery.

The remaining trees all originate either from the FE site and the NE nursery or the FS site and the NG nursery.

```{r FilterDataAndFormatGRM}
subdata <- data %>% 
  dplyr::select(FieldSite,Block,FieldCode,PopulationCode,Family,Nursery,all_of(trait)) %>% 
  drop_na() %>% # 1848 to 1765 values
  filter(FieldSite %in% c("FE","FS")) %>% # 1765 to 1311
  filter(!(FieldSite=="FE" & Nursery =="NG")) %>% # 1311 to 1307
  filter((FieldCode%in%rownames(GRM))) %>%  # 1307 to 1293
  dplyr::rename(y=any_of(trait)) %>% 
  arrange(FieldCode)

# filter the grm with the remainng trees
GRM <-  GRM[as.character(subdata$FieldCode), as.character(subdata$FieldCode), drop = FALSE]
# dim(GRM)
# GRM[1:10,1:10]
# 
# all.equal(as.character(subdata$FieldCode),colnames(GRM))
# all.equal(as.character(subdata$FieldCode),rownames(GRM))
```


```{r RunTheSpeciesLevelModelWithGRM, eval=F}
n_chains <- 4 # number of chains (MCMC)
n_iter <- 6000 # number of iterations
n_warm <- 3000 # number of iterations in the warm-up phase
n_thin <- 1 # thinning interval
n_seed <- 404

form <- brmsformula(y ~ 0 + FieldSite + (0 + FieldSite | gr(FieldCode, cov = GRM)))

mod <- brm(formula = form,
           data = subdata,
           data2 = list(GRM=GRM),
           save_pars = save_pars(group = FALSE),
           chains = n_chains,
           cores = n_chains,
           seed = n_seed,
           iter = n_iter,
           warmup = n_warm)

saveRDS(mod, file = here(paste0("outputs/ScotsPine/StanModels/",trait,"_TwoCG_SpeciesLevelModel_PartVarVillemereuil.rds")))

# Messages d'avis :
# 1: There were 4 chains where the estimated Bayesian Fraction of Missing Information was low. See
# https://mc-stan.org/misc/warnings.html#bfmi-low 
# 2: Examine the pairs() plot to diagnose sampling problems
#  
# 3: The largest R-hat is 1.06, indicating chains have not mixed.
# Running the chains for more iterations may help. See
# https://mc-stan.org/misc/warnings.html#r-hat 
# 4: Bulk Effective Samples Size (ESS) is too low, indicating posterior means and medians may be unreliable.
# Running the chains for more iterations may help. See
# https://mc-stan.org/misc/warnings.html#bulk-ess 
# 5: Tail Effective Samples Size (ESS) is too low, indicating posterior variances and tail quantiles may be unreliable.
# Running the chains for more iterations may help. See
# https://mc-stan.org/misc/warnings.html#tail-ess 
```

Model summary:
  
```{r InspectTheSpeciesLevelModelWithGRM, fig.height=8, fig.width=8}
mod <- readRDS(here(paste0("outputs/ScotsPine/StanModels/",trait,"_TwoCG_SpeciesLevelModel_PartVarVillemereuil.rds")))
summary(mod)
plot(mod, variable=c("b_FieldSiteFE", "b_FieldSiteFS","sigma"))
plot(mod, variable=c("cor_FieldCode__FieldSiteFE__FieldSiteFS", "sd_FieldCode__FieldSiteFE","sd_FieldCode__FieldSiteFS"))
```

```{r PartVarSpeciesLevelModelWithGRM, fig.height=5, fig.width=5, warning=F}
# Getting the mean uncertainty on the fixed-effect parameters
var_uncert <- vcov(mod) |> # extracts the variance-covariance matrix of fixed effects.
  diag() |> # diagonal values represent the variances of each fixed effect parameter.
  mean()  # Calculate mean uncertainty across parameters

# Computing V_plas (variance of the average reaction norm)
post_theta_cs <- fixef(mod, summary = FALSE) |> # extracts posterior samples for fixed effects
  as_draws_df() # formats them as a posterior draws object.

# Select fixed effects corresponding to the sites and compute their row-wise variances.
var_plas_cs <- post_theta_cs |> 
  select(starts_with("FieldSite")) |>
  as.matrix() |> 
  rowVars()

# Correcting V_plas for uncertainty
post_cs <- data.frame(V_Plas = var_plas_cs - var_uncert) |>  # subtract the mean parameter uncertainty (var_uncert) from the raw V_plas values.
  cbind(select(post_theta_cs, starts_with("."))) |>  # combine with MCMC metadata (e.g., chain, iteration, draw) to create a valid posterior object.
  as_draws_df()

# Getting the G-matrix (variance-covariance of random effects)
post_cs[["G"]] <- VarCorr(mod, summary = FALSE)[["FieldCode"]][["cov"]] |> # extracts variance-covariance matrices of random effects
  apply(1, \(mat_) { mat_ }, simplify = FALSE) # apply() to structure the posterior samples into a list of matrices for further processing

# Getting the residual variance
# Extract posterior samples for residual variance
post_cs[["V_R"]] <- VarCorr(mod, summary = FALSE)[["residual__"]][["sd"]][, 1]^2

# Thinning the posterior draws for computational efficiency
# Thin the posterior draws to reduce computation time while maintaining sufficient information
post_cs <- thin_draws(post_cs, thin = length(var_plas_cs) / 1000)

# Save MCMC metadata (iteration, chain, etc.) for further use
post_cs_info <- select(post_cs, starts_with("."))

# Summarising V_Plas (posterior estimates and diagnostics)
# Display summary statistics for V_Plas (mean, median, credible intervals).
summarise_draws(subset_draws(post_cs, variable = "V_Plas"))  %>% kable_mydf()

# Computing genetic variance decomposition
# Apply function to compute genetic variances from the G-matrix
post_gen_cs <- 
  map(post_cs[["G"]], rn_cs_gen, .progress = TRUE) |> 
  bind_rows() |> 
  select(where(\(col_) { abs(mean(col_)) > 10^-5 })) |>  # Remove negligible components
  cbind(post_cs_info) |> 
  as_draws_df()

# Summarise genetic variance components
summarise_draws(post_gen_cs) %>% kable_mydf()

# Compute total phenotypic variance and variance-standardised estimates
post_var_tpc_cs <- bind_draws(post_cs, post_gen_cs) |> 
  subset_draws(variable = c("V_Plas", "V_Add", "V_A", "V_AxE", "V_R")) |> 
  mutate_variables(V_Tot = V_Plas + V_Add + V_R)

# Compute standardised variance estimates (P2, H2_RN, H2, H2_I, T2)
post_std_tpc_cs <- post_var_tpc_cs |> 
  transmute(
    P2_RN = V_Plas / V_Tot,       # Proportion of phenotypic variance due to plasticity
    H2_RN = V_Add / V_Tot,        # Heritability of reaction norm
    H2 = V_A / V_Tot,             # Heritability of environment-blind genetic variance
    H2_I = V_AxE / V_Tot,         # Heritability of genetic variance due to plasticity
    T2_RN = (V_Plas + V_Add) / V_Tot # Total variance explained by genetic and plasticity components
  ) |> 
  cbind(post_cs_info) |> 
  as_draws_df()

# Summarising standardised variance estimates
summarise_draws(post_std_tpc_cs) %>% kable_mydf()
```

MCMC traces of the parameters of interest:

```{r MCMCtracesSpeciesLevelModelWithGRM, fig.height=6, fig.width=10}
mcmc_trace(post_std_tpc_cs)
```

Posterior distributions of the parameters of interest:

```{r MCMCAreasSpeciesLevelModelWithGRM, fig.height=5, fig.width=5}
mcmc_areas(post_std_tpc_cs, prob = 0.95, area_method = "scaled height")
```

<span style="color: orange;">Main take-home message:</span> At the species level, the heritability of the reaction norm ($h_{RN}^2$) mainly comes from the environment-blind heritability of the trait ($h^2$, based on the breeding values averaged across environments) and not from the heritability from plasticity ($h_I^2$, based on the breeding values by environment interaction).


# Population-level models with GRM{#sec-PopmodelGRM}

We apply the same model separately to each population.

```{r RunPopulationSpecificModelsWithGRM, eval=F}
pop_names <- subdata$PopulationCode %>% unique() %>% sort # sorted by alphabetical number

lapply(pop_names, function(pop_i) {
  
  subdata_pop <- subdata %>% 
    filter(PopulationCode %in% pop_i)
  
  GRM_pop <-  GRM[as.character(subdata_pop$FieldCode), as.character(subdata_pop$FieldCode), drop = FALSE]
  
  form <- brmsformula(y ~ 0 + FieldSite + (0 + FieldSite | gr(FieldCode, cov = GRM_pop)))
  
  mod <- brm(formula = form,
             data = subdata_pop,
             data2 = list(GRM_pop=GRM_pop),
             save_pars = save_pars(group = FALSE),
             chains = n_chains,
             cores = n_chains,
             seed = n_seed,
             iter = n_iter,
             warmup = n_warm,
             control = list(adapt_delta = 0.9))
  
  saveRDS(mod, file = here(paste0("outputs/ScotsPine/StanModels/PopulationLevelModels/",trait,"_TwoCG_PopLevelModel_PartVarVillemereuil_pop",pop_i,".rds")))
  })
```

Let's look at the model for the AB population.

```{r VarPartPopLevelModelABPopulation, fig.width=8, fig.height=8, warning=F}
pop_i <- "AB"

mod <- readRDS(here(paste0("outputs/ScotsPine/StanModels/PopulationLevelModels/",trait,"_TwoCG_PopLevelModel_PartVarVillemereuil_pop",pop_i,".rds")))
  
summary(mod)
plot(mod, nvariables = 3)

# Getting the mean uncertainty on the fixed-effect parameters
var_uncert <- vcov(mod) |> # extracts the variance-covariance matrix of fixed effects
  diag() |> # diagonal values represent the variances of each fixed effect parameter
  mean()  # Calculate mean uncertainty across parameters
  
# Computing V_plas (variance of the average reaction norm)
post_theta_cs <- fixef(mod, summary = FALSE) |> # extracts posterior samples for fixed effects
  as_draws_df() # formats them as a posterior draws object.
  
# Select fixed effects corresponding to the field or environment and compute their row-wise variances.
var_plas_cs <- post_theta_cs |> 
  select(starts_with("FieldSite")) |>  # Match "FieldSite" to your model's fixed-effect prefix
  as.matrix() |> 
  rowVars()
  
# Correcting V_plas for uncertainty
post_cs <- data.frame(V_Plas = var_plas_cs - var_uncert) |>  # subtract the mean parameter uncertainty (var_uncert) from the raw V_plas values.
  cbind(select(post_theta_cs, starts_with("."))) |>  # combine with MCMC metadata (e.g., chain, iteration, draw) to create a valid posterior object.
  as_draws_df()
  
# Getting the G-matrix (variance-covariance of random effects)
post_cs[["G"]] <- VarCorr(mod, summary = FALSE)[["FieldCode"]][["cov"]] |> # extracts variance-covariance matrices of random effects
  apply(1, \(mat_) { mat_ }, simplify = FALSE) # apply() to structure the posterior samples into a list of matrices for further processing
  
# Getting the residual variance
# Extract posterior samples for residual variance
post_cs[["V_R"]] <- VarCorr(mod, summary = FALSE)[["residual__"]][["sd"]][, 1]^2

# Thinning the posterior draws for computational efficiency
# Thin the posterior draws to reduce computation time while maintaining sufficient information
post_cs <- thin_draws(post_cs, thin = length(var_plas_cs) / 1000)

# Save MCMC metadata (iteration, chain, etc.) for further use
post_cs_info <- select(post_cs, starts_with("."))

# Summarising V_Plas (posterior estimates and diagnostics)
# Display summary statistics for V_Plas (mean, median, credible intervals).
summarise_draws(subset_draws(post_cs, variable = "V_Plas")) %>% kable_mydf()

# Computing genetic variance decomposition
# Apply function to compute genetic variances from the G-matrix
post_gen_cs <- 
  map(post_cs[["G"]], rn_cs_gen, .progress = TRUE) |> 
  bind_rows() |> 
  select(where(\(col_) { abs(mean(col_)) > 10^-5 })) |>  # Remove negligible components
  cbind(post_cs_info) |> 
  as_draws_df()

# Summarise genetic variance components
summarise_draws(post_gen_cs) %>% kable_mydf()

# Compute total phenotypic variance and variance-standardised estimates
post_var_tpc_cs <- bind_draws(post_cs, post_gen_cs) |> 
  subset_draws(variable = c("V_Plas", "V_Add", "V_A", "V_AxE", "V_R")) |> 
  mutate_variables(V_Tot = V_Plas + V_Add + V_R)

# Compute standardised variance estimates (P2, H2_RN, H2, H2_I, T2)
post_std_tpc_cs <- post_var_tpc_cs |> 
  transmute(
    P2 = V_Plas / V_Tot,          # Proportion of phenotypic variance due to plasticity
    H2_RN = V_Add / V_Tot,        # Heritability of reaction norm
    H2 = V_A / V_Tot,             # Heritability of environment-blind genetic variance
    H2_I = V_AxE / V_Tot,         # Heritability of genetic variance due to plasticity
    T2 = (V_Plas + V_Add) / V_Tot # Total variance explained by genetic and plasticity components
    ) |> 
  cbind(post_cs_info) |> 
  as_draws_df()

# Summarising standardised variance estimates
summarise_draws(post_std_tpc_cs) %>% kable_mydf()

# Plotting MCMC diagnostics and posterior distributions
mcmc_trace(post_std_tpc_cs)
mcmc_areas(post_std_tpc_cs, prob = 0.95, area_method = "scaled height")
```


MCMC traces of the parameters of interest:

```{r MCMCtracesPopLevelModelABPopulation, fig.height=6, fig.width=10}
mcmc_trace(post_std_tpc_cs)
```

Posterior distributions of the parameters of interest:

```{r MCMCAreasPopLevelModelABPopulation, fig.height=5, fig.width=5}
mcmc_areas(post_std_tpc_cs, prob = 0.95, area_method = "scaled height")
```


<span style="color: orange;">Main take-home message:</span> We do not have enough data to estimate those parameters of interest at the population level.



# Species level model with population varying slopes{#sec-SPmodelPopVaryingSlopes}

```{r SubsetDataSpeciesLevelModelRandomSlopes}
subdata <- data %>% 
  dplyr::select(FieldSite,Block,FieldCode,PopulationCode,Family,Nursery,all_of(trait)) %>% 
  drop_na() %>% # 1848 to 1765 values
  filter(FieldSite %in% c("FE","FS")) %>% # 1765 to 1311
  filter(!(FieldSite=="FE" & Nursery =="NG")) %>% # 1311 to 1307
  dplyr::rename(y=any_of(trait)) %>% 
  arrange(FieldCode) %>% 
  mutate(FieldSite_num = as.numeric(as_factor(FieldSite)))
```

```{r RunSpeciesLevelModelRandomSlopes, eval=F}
# Model options
n_chains <- 4 # number of chains (MCMC)
n_iter <- 8000 # number of iterations
n_warm <- 4000 # number of iterations in the warm-up phase
n_thin <- 1 # thinning interval
n_seed <- 404

form <- brmsformula(y ~ 1 + FieldSite_num + (1 + FieldSite_num | PopulationCode))

mod <- brm(formula = form,
           data = subdata,
           chains = n_chains,
           cores = n_chains,
           seed = n_seed,
           iter = n_iter,
           warmup = n_warm,
           control = list(adapt_delta = 0.95))

# no warnings!

saveRDS(mod, file = here(paste0("outputs/ScotsPine/StanModels/",trait,"_TwoCG_SpeciesLevelModel_PopulationVaryingSlopes.rds")))
```

Summary of the model:

```{r LoadSpeciesLevelModelRandomSlopes}
mod <- readRDS(file = here(paste0("outputs/ScotsPine/StanModels/",trait,"_TwoCG_SpeciesLevelModel_PopulationVaryingSlopes.rds"))) 

summary(mod)
```

Interval plots of the population varying slopes (i.e., random slopes).

```{r IntervalPlotsPopulationSpecificSlopesSpeciesLevelModelRandomSlopes, fig.height=10, fig.width=8}
# Extract posterior samples for population-level slopes
population_slopes <- as_draws_df(mod) %>%
  spread_draws(r_PopulationCode[PopulationCode, FieldSite_num]) %>% 
  filter(FieldSite_num=="FieldSite_num")

population_slopes %>%
  ggplot(aes(x = r_PopulationCode, y = PopulationCode)) +
  stat_halfeye(
    .width = c(0.80, 0.95),   # Specify intervals
    fill = "skyblue",
    point_interval = "median_qi",
    color = "blue") +
  theme_bw() +
  labs(title = "MCMC areas of population varying slopes (median, 80% and 95% credible intervals)",
       x = "Slope",
       y = "Population")
```

Global slope of the model and its 95% credible interval.
```{r PlotGlobalSlope, fig.height=5, fig.width=5}
mod_coeff <- mod %>% 
  broom.mixed::tidyMCMC(
    robust = T, # median and mean absolute deviation are used to compute point estimates and uncertainty
    ess = F, 
    rhat = F, 
    conf.int = T,
    conf.level = 0.95)

global_intercept <- mod_coeff$estimate[["b_Intercept"]]
global_slope <- mod_coeff$estimate[["b_FieldSite_num"]]
global_slope_low95 <- mod_coeff[mod_coeff$term=="b_FieldSite_num","conf.low"] %>% pull()
global_slope_high95 <- mod_coeff[mod_coeff$term=="b_FieldSite_num","conf.high"] %>% pull()

ggplot(data = subdata) +
  geom_point(aes(x = FieldSite_num, y = y), color = "gray") +
  scale_x_continuous(breaks = c(1, 2), 
                     limits = c(0.8, 2.2),
                     labels = c("Glensaugh (FE)", "Yair (FS)")) +  # Set x-axis limits and breaks
    geom_ribbon(aes(x = FieldSite_num, 
                    ymin = global_intercept + global_slope_low95 * FieldSite_num, 
                    ymax = global_intercept + global_slope_high95 * FieldSite_num),  
                inherit.aes = FALSE, 
                fill = "blue", 
                alpha = 0.2) + 
  geom_abline(slope = global_slope, 
              intercept = global_intercept, 
              color = "blue", 
              linetype = "dashed", 
              linewidth = 1) +
  theme_bw() +
  labs(title = "Global slope (median and 95% intervals)",
       x = "",
       y = "Tree height (mm)")
```

Now we look the estimated intercept and slope of each population. The credible intervals reflect uncertainty in the estimation of the slope only.

```{r PlotPopRandomSlopes,fig.height=5, fig.width=5}
pop_names <- subdata$PopulationCode %>% unique %>% sort

lapply(pop_names, function(pop_i){
  
  term_intercept <- paste0("r_PopulationCode[",pop_i,",Intercept]")
  term_slope <- paste0("r_PopulationCode[",pop_i,",FieldSite_num]")
  pop_intercept <- mod_coeff[mod_coeff$term==term_intercept,"estimate"] %>% pull
  pop_slope <- mod_coeff[mod_coeff$term==term_slope,"estimate"] %>% pull
  pop_slope_low95 <- mod_coeff[mod_coeff$term==term_slope,"conf.low"] %>% pull
  pop_slope_high95 <- mod_coeff[mod_coeff$term==term_slope,"conf.high"] %>% pull
  
ggplot(data = subdata) +
  geom_point(aes(x = FieldSite_num, y = y), color="gray") +
  scale_x_continuous(breaks = c(1, 2), 
                     limits = c(0.5, 2.5),
                     labels = c("Glensaugh (FE)", "Yair (FS)")) +  # Set x-axis limits and breaks
  geom_ribbon(aes(x = FieldSite_num, 
                  ymin = global_intercept + global_slope_low95 * FieldSite_num, 
                  ymax = global_intercept + global_slope_high95 * FieldSite_num),  
              inherit.aes = FALSE, 
              fill = "blue", 
              alpha = 0.2) + 
  geom_ribbon(aes(x = FieldSite_num, 
                  ymin = (global_intercept + pop_intercept) + (global_slope + pop_slope_low95) * FieldSite_num, 
                  ymax = (global_intercept + pop_intercept) + (global_slope + pop_slope_high95) * FieldSite_num),  
              inherit.aes = FALSE, 
              fill = "red", 
              alpha = 0.2) + 
  geom_point(data=subdata %>% filter(subdata$PopulationCode==pop_i), aes(x=FieldSite_num, y=y), color="red", alpha=0.8) +
  geom_abline(slope = global_slope, 
              intercept = global_intercept, 
              color = "blue", 
              linetype = "dashed", 
              linewidth = 1) +
  geom_abline(slope=(global_slope + pop_slope) , intercept=(global_intercept + pop_intercept), color="red", alpha=0.8, linewidth=0.5) +
  scale_color_manual(values=c("red",rep("gray",20))) +
  theme_bw() +
  labs(title = paste0("Population ",pop_i),
       x = "",
       y = "Tree height (mm)")
})
```

We extract the slopes for comparison.

```{r extractMedianSlopes}
median_slope_df <- mod_coeff %>% 
  filter(stringr::str_detect(term, ",FieldSite_num]$")) %>% 
  mutate(Population = str_extract(term, "\\[([A-Z]{2}),")) %>%
  mutate(Population = str_remove_all(Population, "\\[|,")) %>% 
  dplyr::select(Population, estimate) %>% 
  dplyr::rename(median_slope = estimate)
```


<span style="color: orange;">Main take-home message:</span> Very low variation in the slope across sites among populations, and the differences in slope are not statistically maeningful.




# RDPI{#sec-RDPI}


For that, the first option is to use the two common gardens in which the trees come from a unique nursery, i.e. FS (in which all trees come from NG) and FE (in which all trees come from NE, except four trees that come from NG and that we remove to calculate the RDPI index).

So we calculate the RDPI indexes capturing plasticity to environment variation between FS and FE.

```{r CalculateRDPISubsettingTwoCG, eval=F}
rdpi_data <- data %>% filter(!(FieldSite=="FE" & Nursery =="NG")) # we remove the four trees in FE that come from NG
pop_codes <- unique(rdpi_data$PopulationCode) %>% sort() # population codes
site_codes <- c("FE","FS") # field site codes
site_comb <- combn(site_codes,2)
calc_rel_dist <- function(x){abs(x[1]-x[2])/(x[1]+x[2])}

list_traits <- list(heights = rdpi_data %>% dplyr::select(contains("HA"), -contains("13")) %>% colnames(), # height measurements (except 2013)
                    duration_budburst = rdpi_data %>% dplyr::select(contains("BD")) %>% colnames(), # duration of bud burst
                    timing_budburst = rdpi_data %>% dplyr::select(contains("BT")) %>% colnames()) # time taken to reach budburst stages

lapply(list_traits, function(traits) {
  
  lapply(traits, function(trait){

  lapply(pop_codes, function(pop){
  
  subpop <- rdpi_data %>%
    drop_na(any_of(trait)) %>% 
    dplyr::filter(PopulationCode %in% pop)
  
  n_pop <- nrow(subpop) # nb of individuals in the population

 rel_dist <- lapply(1:dim(site_comb)[[2]], function(nb_comb){
 
    subsite1 <- subpop %>% dplyr::filter(FieldSite %in% site_comb[1,nb_comb])
    subsite2 <- subpop %>% dplyr::filter(FieldSite %in% site_comb[2,nb_comb])
    
    
    lapply(1:nrow(subsite1), function(x) {
      
      
      rbind(subsite1[x,trait],subsite2[,trait]) %>% 
        pull() %>% 
        combn(2, calc_rel_dist) %>% 
        .[1:nrow(subsite2)] %>% 
        as.numeric()
      
      }) %>% unlist()
    
    }) %>% unlist()
 
 N <- length(rel_dist)
 
sum(rel_dist)/N
 
 
 }) %>% 
    setNames(pop_codes) %>% 
    as_tibble() %>% 
    pivot_longer(everything(), names_to = "PopulationCode", values_to = "RDPI") %>% 
    mutate(Trait=trait)
  
  }) %>% bind_rows()

  
}) %>% saveRDS(file=here("outputs/ScotsPine/PhenotypicPlasticity/RDPI_based_on_FE_FS.rds"))
```


```{r CorrelationPlotsRDPIAcrossClassOfTraits, fig.height=7,fig.width=7, results='hide'}
list_titles <- list(list("heights","height"),
                    list("duration_budburst","the duration of budburst"),
                    list("timing_budburst","the time taken to reach budburst stages"))

lapply(list_titles, function(x){
  
  readRDS(file=here("outputs/ScotsPine/PhenotypicPlasticity/RDPI_based_on_FE_FS.rds"))[[x[[1]]]] %>% 
  pivot_wider(values_from = RDPI, names_from = Trait) %>% 
  dplyr::select(-PopulationCode) %>% 
  cor() %>% 
  corrplot(method = 'number',
           type = 'lower', diag = FALSE,
           title=paste0("Correlation among RDPI for ",x[[2]]),
           mar=c(0,0,2,0),
           number.cex=0.9,tl.cex=0.9)
  })
```


Let's look at the correlation with the population varying slopes of the model above (median estimates).
 
```{r CorrelationRDPIvsPopVaryingSlopes}
df <- readRDS(file=here("outputs/ScotsPine/PhenotypicPlasticity/RDPI_based_on_FE_FS.rds"))$heights %>% 
  filter(Trait == trait) %>% 
  dplyr::rename(Population = PopulationCode) %>% 
  inner_join(median_slope_df, by="Population")

cor(df$RDPI, df$median_slope)
```


# Raw difference betwen population means{#sec-RawDiff}

```{r CalculateRawDiffAcrossClassOfTraits}
diff_data <- data %>% 
  filter(FieldSite %in% c("FE","FS")) %>% # 1765 to 1311
  filter(!(FieldSite=="FE" & Nursery =="NG")) # we remove the four trees in FE that come from NG
pop_codes <- unique(diff_data$PopulationCode) %>% sort() # population codes

list_traits <- list(heights = diff_data %>% dplyr::select(contains("HA"), -contains("13")) %>% colnames(), # height measurements (except 2013)
                    duration_budburst = diff_data %>% dplyr::select(contains("BD")) %>% colnames(), # duration of bud burst
                    timing_budburst = diff_data %>% dplyr::select(contains("BT")) %>% colnames()) # time taken to reach budburst stages

lapply(list_traits, function(traits) {
  
  lapply(traits, function(trait){

  lapply(pop_codes, function(pop){
  
  raw_diff <- diff_data %>%
    drop_na(any_of(trait)) %>% 
    dplyr::filter(PopulationCode %in% pop) %>% 
    arrange(FieldSite) %>% 
    group_by(FieldSite) %>% 
    summarise(mean_trait = mean(!!sym(trait), na.rm = TRUE)) %>% 
    summarise(difference = diff(mean_trait)) %>%
    pull(difference)
    
  tibble(PopulationCode = pop,
         Raw_Diff = raw_diff,
         Trait = trait)
  
  }) %>% bind_rows()
    
  }) %>% bind_rows()

  
}) %>% saveRDS(file=here("outputs/ScotsPine/PhenotypicPlasticity/Raw_Diff_based_on_FE_FS.rds"))
```


```{r CorrelationPlotsRawDiffAcrossClassOfTraits, fig.height=7,fig.width=7, results='hide'}
list_titles <- list(list("heights","height"),
                    list("duration_budburst","the duration of budburst"),
                    list("timing_budburst","the time taken to reach budburst stages"))

lapply(list_titles, function(x){
  
  readRDS(file=here("outputs/ScotsPine/PhenotypicPlasticity/Raw_Diff_based_on_FE_FS.rds"))[[x[[1]]]] %>% 
  pivot_wider(values_from = Raw_Diff, names_from = Trait) %>% 
  dplyr::select(-PopulationCode) %>% 
  cor() %>% 
  corrplot(method = 'number',
           type = 'lower', diag = FALSE,
           title=paste0("Correlation among raw differences for ",x[[2]]),
           mar=c(0,0,2,0),
           number.cex=0.9,tl.cex=0.9)
  })
```

# Correlations 

Correlation among population slopes, RDPI and raw differences


 
```{r CorrelationRDPIvsPopVaryingSlopesvsRawDiff}
df <- readRDS(file=here("outputs/ScotsPine/PhenotypicPlasticity/RDPI_based_on_FE_FS.rds"))$heights %>% 
  filter(Trait == trait) %>% 
  dplyr::rename(Population = PopulationCode) %>% 
  inner_join(median_slope_df, by="Population")

readRDS(file=here("outputs/ScotsPine/PhenotypicPlasticity/Raw_Diff_based_on_FE_FS.rds"))$heights %>% 
  filter(Trait == trait) %>% 
  dplyr::rename(Population = PopulationCode) %>% 
  inner_join(df, by=c("Population","Trait")) %>% 
  dplyr::select(-Population,-Trait) %>% 
  cor() %>% 
  corrplot(method = 'number',
           type = 'lower', diag = FALSE,
           title=paste0(""),
           mar=c(0,0,2,0),
           number.cex=0.9,tl.cex=0.9)
```
