---
title: "Plastic responses in disease susceptibility at the population level"
author: "Juliette Archambeau"
date: last-modified
format: 
  html:
    toc: true
    toc-depth: 4
    code-fold: true
    page-layout: full
    number-sections: true
embed-resources: true
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

<style type="text/css">
body {
   font-size: 15px;
}
code.r{
  font-size: 11px;
}
pre {
  font-size: 11px
}

table {
  font-size: 11px
}
</style>

```{r setup, include=F}
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))
knitr::opts_chunk$set(cache=F)
options(width = 300)
library(knitr)
library(kableExtra)
library(tidyverse)
library(here)
library(magrittr)
library(corrplot)
library(readxl)
library(rstan)
library(tidybayes)
library(bayesplot)
library(ggplot2)
library(parallel)
library(tictoc)
library(latex2exp)
library(brms)
library(Reacnorm)
library(matrixStats)
library(posterior)
library(missMDA)
library(ggrepel)
library(cmdstanr)
library(cowplot)

rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

theme_set(theme_bw(base_size = 20))

# my own function for building tables in reports
source(here("scripts/functions/kable_mydf.R"))
```

# Introduction

The goal of this report is to determine whether the populations have different plastic responses for disease susceptibility. We have four disease-related variables: `DNB16`and `DNB22`, which are the percentage of needles with lesions measured in 2016 and 2022, and`NR16`and `NR22`, which are the number of years needle retention, also measured in 2016 and 2022.

In @sec-SynthDiseaseIndex, we performed a PCA on the four disease-related variables and extracted the first two principal components. We scaled them and multiplied PC1 by –1 so that positive values of both components correspond to higher disease susceptibility. These two variables, referred to as `DiseaseIndex_PC1` and `DiseaseIndex_PC2`, are then used as synthetic traits capturing overall disease susceptibility.

To account for nursery effects (i.e., the fact that trees planted in some common gardens originated from different nurseries), we estimated plasticity to environmental variation arising from the combined effects of site and nursery. Consequently, we restricted the analysis to the two common gardens where all trees originated from a single nursery: Yair (FS), where all trees came from NG, and Glensaugh (FE), where all trees came from NE. In Glensaugh, four trees originated from NG; these were excluded from the dataset used to estimate plasticity.

In section @sec-ReacNormVarPart, we applied the approach developed by Villemereuil & Chevin (2025) to assess whether the heritability of the reaction norm ($h_{RN}^2$) primarily stems from the environment-blind heritability of the trait ($h^2$, based on breeding values averaged across environments) or from the heritability of plasticity ($h_I^2$, based on breeding values by environment interaction). We implemented this approach at the species level, rather than at the population level, because it requires a large amount of data. We implemented the "animal model" version of the Villemereuil and Chevin approach, i.e. the version that uses a genomic relationship matrix (GRM) to estimate additive genetic effects. These models take a very long time to run because they require many iterations for the Markov chains to converge and careful parameter tuning. Some warnings still occurred for `DiseaseIndex_PC1`, but I do not believe these are concerning. I will try to re-run this model with further parameter tuning (it may take about two days to complete), in the hope that the warnings will disappear. 
<!-- Because these models were very difficult to fit using the GRM, I also implemented the Villemereuil and Chevin (2025) approach without the genomic relationship matrix in section @sec-SPmodelWithoutGRM. In this case, additive genetic effects cannot still be estimated, only broad-sense heritabilities are accessible. The results differ slightly between the two approaches, as expected, but the main patterns remain consistent, so I believe they can be interpreted with confidence. -->

In @sec-FamReactionNormEstim, we run a standard reaction-norm model by estimating family-level reaction-norms and we then extract two variables of interest: the median reaction-norm slope across families within a population (quantifies the average change in disease susceptibility between sites for that population), and the median variance of family-specific slopes within a population (wich measures the within-population variation in responses to the site contrast). These two parameters summarize, respectively, the population-level environmental response and the heterogeneity among families in that response.

In @sec-MeanTraitValues, we simply quickly extract the mean trait values (sign-inverted and scaled PC1 and scaled PC2) of each population in each field site to compare their values to those of the parameters of interest.

In @sec-RDPI, we quantified phenotypic plasticity using the relative distance plasticity index (RDPI; Valladares et al., 2006), which measures the mean relative phenotypic distance among individuals of the same population grown in different environments. We calculated RDPI in two ways: (i) at the individual level (i.e., the common practice), where all pairwise phenotypic distances between individuals of a given population and from different common gardens are considered, and (ii) using population means per site, which summarizes differences in average population responses across sites.

In @sec-IndexRelationships, we estimated the relationships among all these indice using a principal component analysis (PCA) and a correlation matrix.

In @sec-CorrWithGenomicOffset, we estimated the relationships between our phenology indexes with the predicted genomic offsets. We used genomic offset predictions from the RCP scenario 2.6, the prediction time period 2030-2050 and the average across the four climatic models (ensemble members).


# Indexes capturing disease susceptibility{#sec-SynthDiseaseIndex}

## The data

We load the disease data and show the first 10 rows of the dataset.

```{r LoadDiseaseData}
disease_data <- read.delim(here("data/ScotsPine/Disease_raw_Juliette.txt")) %>% 
  dplyr::rename(FieldCode=id,
                FieldSite=Site)
disease_data[1:10,] %>% kable_mydf()
```


We merge this data with the other phenotypic data from the common gardens filter for either the common garden FS combined with the NG nursery, or the common garden FE and the nursery NE.

```{r LoadPhenoCGData}
data <- readRDS(file=here("data/ScotsPine/formatted_CG_data.rds")) %>% # 1848 obs
  full_join(disease_data,by=c("FieldSite","FieldCode")) %>% 
  filter(FieldSite %in% c("FE","FS")) %>% # 1344 obs
  filter(!(FieldSite=="FE" & Nursery =="NG")) # 1340 obs
```

The dataset contains `r length(unique(data$FieldCode))` individuals, `r length(unique(data$PopulationCode))` populations, `r length(unique(data$Family))` families. Here are the first 10 rows:

```{r VizData, fig.width=8, fig.height=8}
data[1:10,] %>% kable_mydf()

# Correlation matrix plot for phenotypic variables
data %>%
  dplyr::select(HA20, contains("16"), contains("22"), contains("DNB"), contains("NR"), -DI16, -DA16, -HI16) %>%
  cor(use = "pairwise.complete.obs") %>%
  corrplot::corrplot(
    method = "color",
    type = "lower",
    order = "hclust",
    diag=F,
    addCoef.col = "black",
    tl.col = "black",
    tl.srt = 45,
    col = colorRampPalette(c("#E46726", "white","#6D9EC1"))(200),
    title = "Correlation among phenotypic variables",
    mar = c(0,0,2,0)
  )

data %>% 
  dplyr::select(contains("NR"),contains("DNB")) %>% 
  pivot_longer(everything(),names_to="variable",values_drop_na = TRUE) %>% 
  ggplot(aes(x=value)) +  
  geom_histogram(aes(y=after_stat(density)), colour="blue",fill="white",bins = 100) +
  #geom_density(alpha=.2,fill="pink") +
  xlab("") +
  facet_wrap(~variable,scales="free") + 
  theme_bw() 
```

A weak correlation ($\rho = 0.4$) was observed between `NR16` and `NR22`, indicating that needle retention was somewhat consistent across the two years, 2016 and 2022. This was not the case for `DNB16` and `DNB22` ($\rho = 0.05$). Interestingly, `NR22`, and even more so `NR16`, were slightly positively correlated with height. $\rho = -0.45$ between `NR16` and `DNB16`, so trees that had higher percentage of needles with lesions had also less needle retention, which is what we might expect. This was not the case in 2022: very low correlation between `NR22` and `DNB22` ($\rho = 0.15$; unexpected direction). This highlights the importance of accounting for both variables, as they capture different symptoms of the disease. A slightly negative correlation was found between `NR16` and both budburst duration (`BD16`) and stage 6 of budburst (`BT16_6`), suggesting that trees that completed budburst earlier were slightly more likely to retain their needles.

## PCA

We first identify the number of missing values for each trait:

```{r CountNAs}
subdata <- data %>%
  dplyr::select(FieldSite,Block,FieldCode,PopulationCode,Family,Nursery,matches("NR|DNB"))
  
print(sapply(subdata, function(x) sum(is.na(x))))
```

We then check whether the missing values correspond to the same individuals:

```{r PrintRowsWithNAs}
subdata_with_na <- subdata %>%
  filter(if_any(everything(), is.na))

subdata_with_na %>% kable_mydf()
```

Most missing values occur in the same individuals. We therefore remove these individuals before further analyses.


```{r removingNAs}
subdata <- subdata %>% drop_na()
```

We perform a PCA on the dataset without missing values.
The first plot shows individual scores colored by field site.
The second plot shows population centroids for each field site.

```{r PCAwithoutNAs, fig.height=6, fig.width=8}
# select variables of interest
pca_data <- subdata %>% 
  column_to_rownames("FieldCode") %>%
  select(matches("NR|DNB"))

# PCA
pca_noNA <- prcomp(pca_data, scale. = TRUE)
print(summary(pca_noNA))
  
# decide PC1 and PC2 column name
pc1_name <- "DiseaseIndex_PC1"
pc2_name <- "DiseaseIndex_PC2"
  
# Add PC1 and PC2 as new variables
subdata[[pc1_name]] <- pca_noNA$x[, 1]
subdata[[pc2_name]] <- pca_noNA$x[, 2]
  
# Extract PCA scores (individuals) and loadings (variables)
scores_df <- as.data.frame(pca_noNA$x) %>%
  mutate(FieldCode = as.double(rownames(pca_noNA$x))) %>%
  left_join(subdata, ., by = "FieldCode")
  
loadings_df <- as.data.frame(pca_noNA$rotation) %>%
  rownames_to_column(var = "Trait")
  
# Scale loadings for plotting as arrows
scale_factor <- min(
  (max(scores_df$PC1) - min(scores_df$PC1)) / (max(loadings_df$PC1) - min(loadings_df$PC1)),
  (max(scores_df$PC2) - min(scores_df$PC2)) / (max(loadings_df$PC2) - min(loadings_df$PC2))
  ) * 0.6
  
loadings_df <- loadings_df %>%
  mutate(arrow_x = PC1 * scale_factor,
         arrow_y = PC2 * scale_factor)
  
# Plot: individuals as points, coloured by FieldSite, arrows for variables
p <- ggplot() +
  geom_point(data = scores_df, aes(x = PC1, y = PC2, color = FieldSite), alpha = 0.7, size = 2) +
  geom_segment(data = loadings_df,
               aes(x = 0, y = 0, xend = arrow_x, yend = arrow_y),
               arrow = arrow(length = unit(0.3, "cm")), 
               colour = "black", 
               linewidth = 0.6) +
  geom_text_repel(data = loadings_df,
                  aes(x = arrow_x, y = arrow_y, label = Trait),
                  size = 3.5,
                  max.overlaps = 20) +
  labs(x = paste0("PC1 (", round(100 * summary(pca_noNA)$importance[2,1], 1), "%)"),
       y = paste0("PC2 (", round(100 * summary(pca_noNA)$importance[2,2], 1), "%)"),
       color = "Site") +
  theme_classic(base_size = 14) +
  theme(legend.position = "right")
  
print(p)
  
centroids <- scores_df %>%
  group_by(PopulationCode, FieldSite) %>%
  summarise(PC1 = mean(PC1, na.rm = TRUE),
            PC2 = mean(PC2, na.rm = TRUE),
            n = n(), 
            .groups = "drop")
  
p_centroids <- ggplot() +
  geom_point(data = scores_df, aes(x = PC1, y = PC2), color = "grey80", alpha = 0.4, size = 1) +
  geom_point(data = centroids, aes(x = PC1, y = PC2, color = FieldSite), size = 3) +
  geom_text_repel(data = centroids, aes(x = PC1, y = PC2, label = PopulationCode), size = 3, max.overlaps = Inf) +
  geom_segment(data = loadings_df,
               aes(x = 0, y = 0, xend = arrow_x, yend = arrow_y),
               arrow = arrow(length = unit(0.3, "cm")), colour = "black", linewidth = 0.6) +
  geom_text_repel(data = loadings_df, aes(x = arrow_x, y = arrow_y, label = Trait), size = 3.5) +
  labs(x = paste0("PC1 (", round(100 * summary(pca_noNA)$importance[2,1], 1), "%)"),
       y = paste0("PC2 (", round(100 * summary(pca_noNA)$importance[2,2], 1), "%)"),
       title = "PCA with population centroids") +
  theme_minimal(base_size = 13)
  
print(p_centroids)
```


`DiseaseIndex_PC1` and `DiseaseIndex_PC2` are already almost scaled as:
  
  - mean(`DiseaseIndex_PC1`) = `r mean(subdata$DiseaseIndex_PC1)`
  - sd(`DiseaseIndex_PC1`) = `r sd(subdata$DiseaseIndex_PC1)` 
  - mean(`DiseaseIndex_PC2`) = `r mean(subdata$DiseaseIndex_PC2)`
  - sd(`DiseaseIndex_PC2`) = `r sd(subdata$DiseaseIndex_PC2)` 
  
However, to ensure comparability among traits, we scale both variables so that their means are exactly 0 and their standard deviations are exactly 1.

For clarity in interpretation, we use the negative of the PC1 component (i.e., -PC1) in subsequent analyses, ensuring that higher values of both indexes (PC1 and PC2) correspond to greater disease susceptibility. 

Below, we show the first ten rows of the final dataset used for subsequent analyses.

```{r FinalDataset, message=F}
subdata <- subdata %>%
  mutate(
    DiseaseIndex_PC1_scaled = -as.numeric(scale(DiseaseIndex_PC1)),
    DiseaseIndex_PC2_scaled = as.numeric(scale(DiseaseIndex_PC2))
    ) %>%
    arrange(FieldCode)

head(subdata, 10) %>% kable_mydf()
```



# Variance partitioning with GRM {#sec-ReacNormVarPart}

!! Warning !! This section has not yet been run using the sign-reversed PC1 component (–PC1).


We partitioned the variance of the trait of interest at the species level, i.e., with all populations combined. For that, we follow the methodology in [de Villemereuil & Chevin (2025)](https://doi.org/10.24072/pcjournal.515) to partition the phenotypic variance of the traits of interest. Below are sections extracted from the paper:

$$V_P =  V_{Plas} + V_{Gen} + V_{Res}$$

$V_P$ is the total phenotypic variance of the trait.
$V_{Plas}$ is the phenotypic variance arising from changes in the mean reaction norm across environments.
$V_{Gen}$ is the total genetic variance in the trait across environments.
$V_{Res}$ is the residual variance, not explained by the reaction norm.

$$V_{Gen} = V_{Add} + V_{nonAdd}$$

$V_{Add}$ is the total additive genetic variance in the trait across environments.
$V_{nonAdd}$ is the non additive genetic variance in the trait.

$$V_{Add} =  V_A + V_{AxE}$$

$V_A$ is the environment-blind additive genetic variance of the trait, i.e., based on the mean breeding values across environments.
$V_{AxE}$ is the additive genetic variance arising from plasticity, i.e., variance of the mean-centered breeding values.

From this, it is possible to derive unitless quantities of interest.

$P_{RN}^2 = V_{Plas}/V_P$ is the proportion of the phenotypic variance arising from average plastic responses to environments.

Variance-standardised additive genetic variances are heritabilities:

$h_{RN}^2=\frac{V_{Add}}{V_P}=\frac{V_A}{V_P}+\frac{V_{AxE}}{V_P}=h^2+h^2_I$

In other words, the heritability of the trait when fully accounting for its reaction norm ($h_{RN}^2$) is equal to the environment-blind heritability of the trait ($h^2$, based on the breeding values averaged across environments) plus the heritability from plasticity ($h_I^2$, based on the breeding values by environment interaction).

$T_{RN}^2=\frac{V_{Plas}+V_{Gen}}{V_P}=P_{RN}^2+h_{RN}^2$

$T_{RN}^2$ is the proportion of the phenotypic variance explained by the (possibly plastic and genetically variable) reaction norm, and thus our ability to predict the individual phenotype from the genotype and the environment.

To apply this methodology, we followed the tutorials associated with the `Reacnorm` R package: <https://github.com/devillemereuil/Reacnorm>.

## Running the model

We load the genomic relationship matrix. Here are the first 10 rows and columns.

```{r LoadGRM}
GRM <- readRDS(file=here("data/ScotsPine/GenomicData/data_update_Fev2025/formatted_data/Gmat_VanRaden.rds"))
GRM[1:10,1:10]
```

Matrix dimensions:

```{r}
dim(GRM)
```

```{r FilterDataAndFormatGRM}
subdata_grm <- subdata %>% 
  #dplyr::select(FieldSite,Block,FieldCode,PopulationCode,Family,Nursery,all_of(traits)) %>% 
  #drop_na() %>% # no NAs
  filter((FieldCode%in%rownames(GRM))) %>%  # 1340 to 1297
  #dplyr::rename(y=any_of(traits)) %>% 
  arrange(FieldCode)

# filter the grm with the remainng trees
GRM <-  GRM[as.character(subdata_grm$FieldCode), as.character(subdata_grm$FieldCode), drop = FALSE]
# dim(GRM)
# GRM[1:10,1:10]
# 
# all.equal(as.character(subdata$FieldCode),colnames(GRM))
# all.equal(as.character(subdata$FieldCode),rownames(GRM))
```

We run the model for PC1 and PC2 of the PCA: `DiseaseIndex_PC1` and `DiseaseIndex_PC2`.

```{r RunTheSpeciesLevelModelWithGRM, eval=F}
n_chains <- 4 # number of chains (MCMC)
n_iter <- 7000 # number of iterations
n_warm <- 3000 # number of iterations in the warm-up phase
n_thin <- 1 # thinning interval
n_seed <- 404

form_1 <- brmsformula(DiseaseIndex_PC1 ~ 0 + FieldSite + (0 + FieldSite | gr(FieldCode, cov = GRM)))
form_2 <- brmsformula(DiseaseIndex_PC2 ~ 0 + FieldSite + (0 + FieldSite | gr(FieldCode, cov = GRM)))

mod <- brm(formula = form_1,
           data = subdata_grm,
           data2 = list(GRM=GRM),
           save_pars = save_pars(group = FALSE),
           chains = n_chains,
           cores = n_chains,
           seed = n_seed,
           iter = n_iter,
           backend = "cmdstanr",
           warmup = n_warm, 
           control = list(adapt_delta = 0.99))

saveRDS(mod, file = here(paste0("outputs/ScotsPine/StanModels/DiseaseIndex_PC1_TwoCG_SpeciesLevelModel_PartVarVillemereuil_AdaptDelta99_NoNAS.rds")))

mod <- brm(formula = form_2,
           data = subdata_grm,
           data2 = list(GRM=GRM),
           save_pars = save_pars(group = FALSE),
           chains = n_chains,
           cores = n_chains,
           seed = n_seed,
           iter = n_iter,
           backend = "cmdstanr",
           warmup = n_warm, 
           control = list(adapt_delta = 0.99))

saveRDS(mod, file = here(paste0("outputs/ScotsPine/StanModels/DiseaseIndex_PC2_scaled_TwoCG_SpeciesLevelModel_PartVarVillemereuil_AdaptDelta99.rds")))
```

## Diagnostic checks

```{r DiagnosticChecks}
# Defining functions
# ==================

compute_bfmi_chain_vec <- function(energy_vec){
  ediff2 <- diff(energy_vec)^2
  mean(ediff2, na.rm = TRUE) / var(energy_vec, na.rm = TRUE)
}

get_sampling_diagnostics <- function(fit_object){
  sp <- rstan::get_sampler_params(fit_object$fit, inc_warmup = FALSE)
  
  # BFMI per chain
  bfmi_per_chain <- sapply(sp, function(x) compute_bfmi_chain_vec(x[, "energy__"]))
  
  # Divergences and treedepth exceedances per chain
  divergences <- sapply(sp, function(x) sum(x[, "divergent__"]))
  treedepth_exceeded <- sapply(sp, function(x) sum(x[, "treedepth__"] >= 10))
  
  data.frame(
    Mean_BFMI = mean(bfmi_per_chain),
    SD_BFMI   = sd(bfmi_per_chain),
    Min_BFMI  = min(bfmi_per_chain),
    Max_BFMI  = max(bfmi_per_chain),
    Divergences = sum(divergences),
    Treedepth_Exceeded = sum(treedepth_exceeded)
  )
}

# Load models
# ===========

mod_PC1 <- readRDS(here("outputs/ScotsPine/StanModels/DiseaseIndex_PC1_TwoCG_SpeciesLevelModel_PartVarVillemereuil_AdaptDelta99.rds"))
mod_PC2 <- readRDS(here("outputs/ScotsPine/StanModels/DiseaseIndex_PC2_TwoCG_SpeciesLevelModel_PartVarVillemereuil_AdaptDelta99.rds"))


# Compute diagnostics
# ===================

diag_PC1 <- get_sampling_diagnostics(mod_PC1)
diag_PC2 <- get_sampling_diagnostics(mod_PC2)


# Combine results into one table
# ==============================

diagnostics_summary <- bind_rows(
  cbind(Model = "DiseaseIndex_PC1", diag_PC1),
  cbind(Model = "DiseaseIndex_PC2", diag_PC2)
)


# Display results
#================
diagnostics_summary %>% kable_mydf()
```

**Interpretation**

> Energy–Bayesian Fraction of Missing Information (E-BFMI)

The E-BFMI quantifies how efficiently the Hamiltonian Monte Carlo (HMC) sampler explores the posterior energy landscape.

Values below 0.3 indicate inefficient energy mixing and potential sampling pathologies (e.g., poor exploration of momentum or funnel-shaped posteriors).

Values above 0.3 are generally considered acceptable.

The `DiseaseIndex_PC1` model has a mean E-BFMI of 0.22, with all chains below the 0.3 threshold. This indicates poor energy mixing, suggesting that the posterior geometry is challenging for the sampler. While no divergences occurred, the low BFMI implies that posterior means and variances may be reliable, but tail behavior and covariance structure could be underestimated or biased. This model would benefit from reparameterization or improved scaling of the random effect covariance structure (e.g., rescaling the GRM or adjusting priors).

The `DiseaseIndex_PC2` model, in contrast, shows a mean E-BFMI of 0.44, comfortably above the 0.3 threshold. The sampler efficiently explored the posterior energy landscape, and there were no divergences or treedepth saturations, indicating well-behaved sampling and reliable posterior estimation.

> Divergent transitions

Neither model exhibited divergent transitions (0 across all chains). This indicates that the numerical integrator successfully followed the Hamiltonian trajectories without hitting regions of extreme curvature — a strong indicator of stable sampling.

> Tree depth exceedances

No chains in either model reached the maximum tree depth limit (0 across all chains). This suggests that the No-U-Turn Sampler (NUTS) was able to adapt its trajectory lengths appropriately without hitting computational limits, further supporting convergence.

## Model summary

### Response variable: PC1
  
```{r InspectTheSpeciesLevelModelWithGRMPC1, fig.height=8, fig.width=10, eval=T}
summary(mod_PC1)
plot(mod_PC1, variable=c("b_FieldSiteFE", "b_FieldSiteFS","sigma"))
plot(mod_PC1, variable=c("cor_FieldCode__FieldSiteFE__FieldSiteFS", "sd_FieldCode__FieldSiteFE","sd_FieldCode__FieldSiteFS"))
```

### Response variable: PC2

```{r InspectTheSpeciesLevelModelWithGRMPC2, fig.height=8, fig.width=10, eval=T}
summary(mod_PC2)
plot(mod_PC2, variable=c("b_FieldSiteFE", "b_FieldSiteFS","sigma"))
plot(mod_PC2, variable=c("cor_FieldCode__FieldSiteFE__FieldSiteFS", "sd_FieldCode__FieldSiteFE","sd_FieldCode__FieldSiteFS"))
```


## Variance partitioning


```{r PartVarSpeciesLevelModelWithGRM, warning=F}
# ─────────────────────────────────────────────
# Function to perform variance decomposition
# ─────────────────────────────────────────────

calculate_var_part <- function(mod, model_name) {
  
  # Mean uncertainty on fixed-effect parameters
  var_uncert <- vcov(mod) |>  # extracts the variance-covariance matrix of fixed effects.
    diag() |> # diagonal values represent the variances of each fixed effect parameter.
    mean()
  
  # Posterior draws for fixed effects
  post_theta <- fixef(mod, summary = FALSE) |> # extracts posterior samples for fixed effects
    as_draws_df() # formats them as a posterior draws object.
  
  # Compute variance of plasticity (V_Plas) = variance of the average reaction norm
  var_plas <- post_theta |> 
    select(starts_with("FieldSite")) |> # Select fixed effects corresponding to the sites 
    as.matrix() |> 
    matrixStats::rowVars() # Compute the row-wise variances of the selected fixed effects
  
  # Correcting V_plas for uncertainty
  post <- data.frame(V_Plas = var_plas - var_uncert) |>   # subtract the mean parameter uncertainty (var_uncert) from the raw V_plas values.
    cbind(select(post_theta, starts_with("."))) |>  # combine with MCMC metadata (e.g., chain, iteration, draw) to create a valid posterior object.
    as_draws_df()
  
  # Getting the G-matrix (variance-covariance of random effects)
  post[["G"]] <- VarCorr(mod, summary = FALSE)[["FieldCode"]][["cov"]] |> # extracts variance-covariance matrices of random effects
    apply(1, \(mat_) { mat_ }, simplify = FALSE) # apply() to structure the posterior samples into a list of matrices for further processing

  
  # Getting the residual variance
  # Extract posterior samples for residual variance
  post[["V_R"]] <- VarCorr(mod, summary = FALSE)[["residual__"]][["sd"]][, 1]^2
  
  # Optional step: thinning
  # Thinning the posterior draws for computational efficiency (to reduce computation time while maintaining sufficient information)
  post <- thin_draws(post, thin = length(var_plas) / 1000)
  
  # Save MCMC metadata (iteration, chain, etc.) for further use
  post_info <- select(post, starts_with("."))
  
  # Summarising V_Plas (posterior estimates and diagnostics)
  # Display summary statistics for V_Plas (mean, median, credible intervals).
  V_plas_summary <- summarise_draws(subset_draws(post, variable = "V_Plas"))
  
  # Compute genetic variance decomposition 
  # Apply the rn_cs_gen function that computes genetic variances from the G-matrix
  post_gen <- map(post[["G"]], rn_cs_gen, .progress = TRUE) |> 
    bind_rows() |> 
    select(where(\(col_) abs(mean(col_)) > 1e-5)) |> 
    cbind(post_info) |> 
    as_draws_df()
  
  # Summarise genetic variance components
  gen_summary <- summarise_draws(post_gen)
  
  # Compute total phenotypic variance and variance-standardised estimates
  post_var_tpc <- bind_draws(post, post_gen) |> 
    subset_draws(variable = c("V_Plas", "V_Add", "V_A", "V_AxE", "V_R")) |> 
    mutate_variables(V_Tot = V_Plas + V_Add + V_R)
  
  post_std_tpc <- post_var_tpc |> 
    transmute(
      P2_RN = V_Plas / V_Tot,
      H2_RN = V_Add / V_Tot,
      H2 = V_A / V_Tot,
      H2_I = V_AxE / V_Tot,
      T2_RN = (V_Plas + V_Add) / V_Tot
    ) |> 
    cbind(post_info) |> 
    as_draws_df()
  
  # Summarising standardised variance estimates
  std_summary <- summarise_draws(post_std_tpc)
  
  # Return results for later use
  list(
    model_name = model_name,
    post_std_tpc_cs = post_std_tpc,
    V_Plas_summary = V_plas_summary,
    GeneticVar_summary = gen_summary,
    StdVar_summary = std_summary
  )
}

# ─────────────────────────────────────────────
# Apply to both models (mod_PC1 and mod_PC2)
# ─────────────────────────────────────────────

results_PC1 <- calculate_var_part(mod_PC1, "DiseaseIndex_PC1")
results_PC2 <- calculate_var_part(mod_PC2, "DiseaseIndex_PC2")

# ─────────────────────────────────────────────
# Combine summary tables in one output
# ─────────────────────────────────────────────

summary_table <- bind_rows(
  mutate(results_PC1$StdVar_summary, Model = "DiseaseIndex_PC1"),
  mutate(results_PC2$StdVar_summary, Model = "DiseaseIndex_PC2")
) %>%
  select(Model, everything())

summary_table %>% kable_mydf()
```


```{r PartVarSpeciesLevelModelWithGRMPC1, fig.height=5, fig.width=10, warning=F, eval=F, warning=F}
# Getting the mean uncertainty on the fixed-effect parameters
var_uncert <- vcov(mod_PC1) |> # extracts the variance-covariance matrix of fixed effects.
  diag() |> # diagonal values represent the variances of each fixed effect parameter.
  mean()  # Calculate mean uncertainty across parameters

# Computing V_plas (variance of the average reaction norm)
post_theta_cs <- fixef(mod_PC1, summary = FALSE) |> # extracts posterior samples for fixed effects
  as_draws_df() # formats them as a posterior draws object.

# Select fixed effects corresponding to the sites and compute their row-wise variances.
var_plas_cs <- post_theta_cs |> 
  select(starts_with("FieldSite")) |>
  as.matrix() |> 
  rowVars()

# Correcting V_plas for uncertainty
post_cs <- data.frame(V_Plas = var_plas_cs - var_uncert) |>  # subtract the mean parameter uncertainty (var_uncert) from the raw V_plas values.
  cbind(select(post_theta_cs, starts_with("."))) |>  # combine with MCMC metadata (e.g., chain, iteration, draw) to create a valid posterior object.
  as_draws_df()

# Getting the G-matrix (variance-covariance of random effects)
post_cs[["G"]] <- VarCorr(mod, summary = FALSE)[["FieldCode"]][["cov"]] |> # extracts variance-covariance matrices of random effects
  apply(1, \(mat_) { mat_ }, simplify = FALSE) # apply() to structure the posterior samples into a list of matrices for further processing

# Getting the residual variance
# Extract posterior samples for residual variance
post_cs[["V_R"]] <- VarCorr(mod, summary = FALSE)[["residual__"]][["sd"]][, 1]^2

# Thinning the posterior draws for computational efficiency
# Thin the posterior draws to reduce computation time while maintaining sufficient information
post_cs <- thin_draws(post_cs, thin = length(var_plas_cs) / 1000)

# Save MCMC metadata (iteration, chain, etc.) for further use
post_cs_info <- select(post_cs, starts_with("."))

# Summarising V_Plas (posterior estimates and diagnostics)
# Display summary statistics for V_Plas (mean, median, credible intervals).
summarise_draws(subset_draws(post_cs, variable = "V_Plas"))  %>% kable_mydf()

# Computing genetic variance decomposition
# Apply function to compute genetic variances from the G-matrix
post_gen_cs <- 
  map(post_cs[["G"]], rn_cs_gen, .progress = TRUE) |> 
  bind_rows() |> 
  select(where(\(col_) { abs(mean(col_)) > 10^-5 })) |>  # Remove negligible components
  cbind(post_cs_info) |> 
  as_draws_df()

# Summarise genetic variance components
summarise_draws(post_gen_cs) %>% kable_mydf()

# Compute total phenotypic variance and variance-standardised estimates
post_var_tpc_cs <- bind_draws(post_cs, post_gen_cs) |> 
  subset_draws(variable = c("V_Plas", "V_Add", "V_A", "V_AxE", "V_R")) |> 
  mutate_variables(V_Tot = V_Plas + V_Add + V_R)

# Compute standardised variance estimates (P2, H2_RN, H2, H2_I, T2)
post_std_tpc_cs <- post_var_tpc_cs |> 
  transmute(
    P2_RN = V_Plas / V_Tot,       # Proportion of phenotypic variance due to plasticity
    H2_RN = V_Add / V_Tot,        # Heritability of reaction norm
    H2 = V_A / V_Tot,             # Heritability of environment-blind genetic variance
    H2_I = V_AxE / V_Tot,         # Heritability of genetic variance due to plasticity
    T2_RN = (V_Plas + V_Add) / V_Tot # Total variance explained by genetic and plasticity components
  ) |> 
  cbind(post_cs_info) |> 
  as_draws_df()

# Summarising standardised variance estimates
summarise_draws(post_std_tpc_cs) %>% kable_mydf()
```

## Plots

MCMC traces of the parameters of interest:

```{r MCMCtracesSpeciesLevelModelWithGRM, fig.height=6, fig.width=10, eval=T}
mcmc_trace(results_PC1$post_std_tpc_cs) +
  ggtitle("MCMC traces - DiseaseIndex_PC1") +
  theme(plot.title = element_text(size = 14))  

mcmc_trace(results_PC2$post_std_tpc_cs) +
  ggtitle("MCMC traces - DiseaseIndex_PC2") +
  theme(plot.title = element_text(size = 14))  
```

Posterior distributions of the parameters of interest:

```{r MCMCAreasSpeciesLevelModelWithGRMPC1, fig.height=5, fig.width=5, eval=T}
mcmc_areas(results_PC1$post_std_tpc_cs, 
           prob = 0.95, 
           area_method = "scaled height") +
  ggtitle("MCMC areas - DiseaseIndex_PC1") +
  theme(plot.title = element_text(size = 14)) 

mcmc_areas(results_PC2$post_std_tpc_cs, 
           prob = 0.95, 
           area_method = "scaled height") +
  ggtitle("MCMC areas - DiseaseIndex_PC2") +
  theme(plot.title = element_text(size = 14)) 
```

<span style="color: orange;">Main take-home message:</span> 
For `DiseaseIndex_PC1`: the heritability of the reaction norm ($h_{RN}^2 = 0.39$) mainly comes from the environment-blind heritability of the trait ($h^2=0.34$, based on the breeding values averaged across environments) and not from the heritability from plasticity ($h_I^2=0.05$, based on the breeding values by environment interaction).
For `DiseaseIndex_PC2`: the heritability of the reaction norm ($h_{RN}^2=0.28$) comes both from the environment-blind heritability of the trait ($h^2 = 17$) and the heritability from plasticity ($h_I^2=0.11$).

# Estimating family reaction norms{#sec-FamReactionNormEstim}

We remove families that are found in only one field site.

```{r RemovingFamiliesInOneSiteOnly}
# Count how many FieldSites each family occurs in
family_sites <- subdata %>%
  group_by(PopulationCode, Family) %>%
  summarise(n_sites = n_distinct(FieldSite), .groups = "drop")

# Families that occur in only one FieldSite
single_site_families <- family_sites %>%
  filter(n_sites == 1)

single_site_families %>% kable_mydf

# Remove families present in only one FieldSite
subdata <- subdata %>%
  group_by(PopulationCode, Family) %>%
  filter(n_distinct(FieldSite) > 1) %>%
  ungroup()
```


## Running the model

```{r RunTheSpeciesLevelModelWithoutGRMNested, eval=F}
n_chains <- 4 # number of chains (MCMC)
n_iter <- 4000 # number of iterations
n_warm <- 1500 # number of iterations in the warm-up phase
n_thin <- 1 # thinning interval
n_seed <- 404

form_1 <- brmsformula(DiseaseIndex_PC1_scaled ~ 0 + FieldSite + (0 + FieldSite | PopulationCode/Family))
form_2 <- brmsformula(DiseaseIndex_PC2_scaled ~ 0 + FieldSite + (0 + FieldSite | PopulationCode/Family))

mod <- brm(formula = form_1,
           data = subdata,
           #save_pars = save_pars(group = FALSE),
           chains = n_chains,
           cores = n_chains,
           seed = n_seed,
           iter = n_iter,
           backend = "cmdstanr",
           warmup = n_warm)

saveRDS(mod, file = here(paste0("outputs/ScotsPine/StanModels/DiseaseIndex_PC1_scaled_TwoCGs_MixedModelWithoutGRM.rds")))

mod <- brm(formula = form_2,
           data = subdata,
           #save_pars = save_pars(group = FALSE),
           chains = n_chains,
           cores = n_chains,
           seed = n_seed,
           iter = n_iter,
           backend = "cmdstanr",
           warmup = n_warm)

saveRDS(mod, file = here(paste0("outputs/ScotsPine/StanModels/DiseaseIndex_PC2_scaled_TwoCGs_MixedModelWithoutGRM.rds")))
```


## Model summary

### Response variable: sign-inverted and scaled PC1
  
```{r InspectTheSpeciesLevelModelWithoutGRMPC1Nested, fig.height=8, fig.width=10, eval=T}
mod_PC1 <- readRDS(file = here(paste0("outputs/ScotsPine/StanModels/DiseaseIndex_PC1_scaled_TwoCGs_MixedModelWithoutGRM.rds")))
summary(mod_PC1)
#plot(mod_PC1, variable = c("^b_", "sigma"), regex = TRUE)
```

### Response variable: scaled PC2

```{r InspectTheSpeciesLevelModelWithoutGRMPC2Nested, fig.height=8, fig.width=10, eval=T}
mod_PC2 <- readRDS(file = here(paste0("outputs/ScotsPine/StanModels/DiseaseIndex_PC2_scaled_TwoCGs_MixedModelWithoutGRM.rds")))
summary(mod_PC2)
#plot(mod_PC2, variable = c("^b_", "sigma"), regex = TRUE)
```

## Family-level reaction norms

### Interval plots of the family-level slopes


The graphs below show the family-level reaction norms between the two field sites (slope between FE and FS). Each point represents the median posterior slope for a family, calculated as the difference in predicted trait values between the sites, and the horizontal error bars indicate the 95% credible interval. Families are ordered along the y-axis by increasing median slope. The dashed vertical line at zero represents no difference between sites; families to the right exhibit higher predicted values in FE than FS, while families to the left exhibit higher values in FS than FE.

```{r IntervalPlotFamilyLevelReactionNormNested, fig.height=15, fig.width=8}
interval_plot_family_reaction_norms_nested <- function(mod, y_label){

  family_diff <- mod %>%
    spread_draws(`r_PopulationCode:Family`[PopFam, FieldSite]) %>% 
    # PopFam is of the form "POP:Family"
    mutate(
      PopulationCode = sub("_.*", "", PopFam),
      Family         = sub(".*_", "", PopFam)
    ) %>%
    dplyr::rename(slope = `r_PopulationCode:Family`) %>%
    pivot_wider(names_from = FieldSite, values_from = slope) %>%
    mutate(
      slope_FE_FS = FieldSiteFE - FieldSiteFS
    ) %>%
    group_by(PopulationCode, Family) %>%
    summarise(
      median_slope = median(slope_FE_FS),
      lower_slope  = quantile(slope_FE_FS, 0.025),
      upper_slope  = quantile(slope_FE_FS, 0.975),
      .groups = "drop"
    ) %>%
    arrange(median_slope) %>%
    mutate(Family = factor(Family, levels = Family))

  ggplot(family_diff, aes(x = median_slope, y = Family)) +
    geom_point() +
    geom_errorbar(aes(xmin = lower_slope, xmax = upper_slope), width = 0.2) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
    theme_bw() +
    labs(
      y = "",
      x = "Family-level reaction norms (FE − FS) with 95% credible intervals",
      title = y_label
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}


interval_plot_family_reaction_norms_nested(mod = mod_PC1, y_label = "Disease index PC1 (sign-inverted and scaled)")
interval_plot_family_reaction_norms_nested(mod = mod_PC2, y_label = "Disease index PC2 (scaled)")
```

### Reaction-norm plots

#### Observed relationships

We compute and visualize the **observed** (non-modeled) family-level reaction norms of disease indices across the two field sites (FE and FS). More precisely, we directly use the **raw data** to calculate the median disease index per family and site, and we then generate two plots.

The first plot displays the family-level median reaction norms of all families together. The background points represent individual tree observations. The colored lines connect the median trait values per family across the two field sites. Colors correspond to populations. This plot provides an overall view of the range of family-level differentiation in reaction norms.

The second plot shows the same information but faceted by population, allowing clearer visualization of within-population variation in median slopes. Each panel corresponds to one population. Background points represent individual trees. Blue lines connect family median values across sites.


```{r PredictionsVizSpeciesLevelModelWithoutGRMNested, fig.height=10, fig.width=10}
observed_family_reaction_norms <- function(mod, subdata, y_var, y_label) {
  
  
  # Compute family medians per FieldSite
  family_means <- subdata %>%
    group_by(Family, FieldSite) %>%
    summarise(median_DI = median(.data[[y_var]], na.rm = TRUE), .groups = "drop") %>%
    mutate(PopulationCode = substr(as.character(Family), 1, 2)) 
  
  # Plots
  p_all_reaction_norms <- 
    ggplot() +
    # Points = individual trees, colored by family
    geom_point(data = subdata,
               aes(x = FieldSite, y = .data[[y_var]], colour = PopulationCode),
               position = position_jitter(width = 0.1, height = 0), alpha = 0.5) +
    # Lines = family median across sites
    geom_line(data = family_means,
              aes(x = FieldSite, y = median_DI, group = Family, colour = PopulationCode),
              linewidth = 1) +
    # Aesthetic tuning
    theme_minimal(base_size = 14) +
    #theme(legend.position = "none") +
    labs(x = "Field site", y = y_label) +
    scale_colour_viridis_d(option = "turbo", name = "Populations")  # nice categorical palette
  
  
  
  p_reaction_norms_per_pop <- ggplot() +
  # raw data points
  geom_jitter(data = subdata,
              aes(x = FieldSite, y = .data[[y_var]]),
              width = 0.1, alpha = 0.4, show.legend = FALSE) +
  
  # median/mean predictions per family
  geom_line(data = family_means,
            aes(x = FieldSite, y = median_DI, group = Family),
            linewidth = 0.8, color = "blue", alpha = 0.3) +

  # facet by population
  facet_wrap(~PopulationCode, scales = "fixed") +
  
  theme_bw() +
  labs(
    x = "Field Site",
    y = y_label,
    title = "Family-level observed reaction norms (median values)"
  ) +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12))
  
  
  return(list(p_all_reaction_norms = p_all_reaction_norms,
              p_reaction_norms_per_pop = p_reaction_norms_per_pop))
}


observed_family_reaction_norms(mod_PC1, subdata, "DiseaseIndex_PC1_scaled", y_label = "Disease index PC1 (sign-inverted and scaled)")
observed_family_reaction_norms(mod_PC2, subdata, "DiseaseIndex_PC2_scaled", y_label = "Disease index PC2 (scaled)")
```

#### Fitted relationships

##### Family-level

In the plots below, family-level reaction norms were obtained by generating posterior expected values from the fitted *brms* models using the function `add_epred_draws()` from *tidybayes*. For each combination of *FieldSite × Population × Family*, we constructed a prediction grid and extracted posterior draws of the expected response (`.epred`), specifying `re_formula = ~(0 + FieldSite | PopulationCode/Family)` so that predictions include the fixed *FieldSite* effect, the population-level random slopes, and the family-level random slopes. The function returns the posterior distribution of the linear predictor after marginalising over residual variance.

The first plot show the family-level reaction norms of all families together. Background points correspond to the observed scaled disease index values, colored by population.

The second plot is faceted by population, allowing visualization of within-population variation in family responses. Background points correspond to the observed scaled disease index values. The red line represents the global median response across all families, providing a reference for the overall trend.


```{r ReactionNormPlotFamilyLevelNested, fig.width=10, fig.height=10}
family_reaction_norms <- function(mod, subdata, y_var, y_label){
  
  valid_combos <- subdata %>% distinct(PopulationCode, Family)
  newdata <- crossing(FieldSite = unique(subdata$FieldSite), valid_combos)
  
  # Posterior fitted values including BOTH random-effect levels
  preds <- add_epred_draws(
    mod,
    newdata = newdata,
    re_formula = ~(0 + FieldSite | PopulationCode/Family))
  
  # Summaries per family × fieldsite
  family_summary <- preds %>%
    group_by(PopulationCode, Family, FieldSite) %>%
    summarise(
      median_pred       = median(.epred),
      mean_pred         = mean(.epred),
      lower_quant_pred  = quantile(.epred, 0.025),
      upper_quant_pred  = quantile(.epred, 0.975),
      sd_pred           = sd(.epred),
      lower_sd_pred     = mean_pred - sd_pred,
      upper_sd_pred     = mean_pred + sd_pred,
      .groups = "drop"
    )
  
  # Global summary across all families
  global_summary <- preds %>%
    group_by(FieldSite) %>%
    summarise(
      median_all       = median(.epred),
      mean_all         = mean(.epred),
      lower_quant_all  = quantile(.epred, 0.025),
      upper_quant_all  = quantile(.epred, 0.975),
      sd_all           = sd(.epred),
      lower_sd_all     = mean_all - sd_all,
      upper_sd_all     = mean_all + sd_all,
      .groups = "drop"
    )
  
  ## ------------------------------
  ## Plot 1 — All families together
  ## ------------------------------
  
  plot_all_fams <- ggplot() +
    geom_jitter(
      data = subdata,
      aes(x = FieldSite, y = .data[[y_var]], color = PopulationCode),
      width = 0.1, alpha = 0.4, show.legend = FALSE
    ) +
    geom_ribbon(
      data = family_summary,
      aes(x = FieldSite, ymin = lower_quant_pred, ymax = upper_quant_pred, group = Family, fill = PopulationCode),
      alpha = 0.05, show.legend = FALSE
    ) +
    geom_line(
      data = family_summary,
      aes(x = FieldSite, y = median_pred, group = Family, color = PopulationCode),
      linewidth = 0.8
    ) +
    theme_bw() +
    labs(x = "Field Site", y = y_label,
         title = "Family-level reaction norms (median ± 95% CI)") +
    theme(axis.text = element_text(size = 12))
  
  
  ## ----------------------------------
  ## Plot 2 — Faceted by population
  ## ----------------------------------
  
  plot_facet_median_with_CI <- ggplot() +
    geom_jitter(
      data = subdata,
      aes(x = FieldSite, y = .data[[y_var]]),
      width = 0.1, alpha = 0.4
    ) +
    geom_ribbon(
      data = family_summary,
      aes(x = FieldSite, ymin = lower_quant_pred, ymax = upper_quant_pred, group = Family),
      alpha = 0.05, fill = "blue"
    ) +
    geom_line(
      data = family_summary,
      aes(x = FieldSite, y = median_pred, group = Family),
      linewidth = 0.8, color = "blue", alpha = 0.3
    ) +
    geom_ribbon(
      data = global_summary,
      aes(x = FieldSite, ymin = lower_quant_all, ymax = upper_quant_all),
      alpha = 0.2, fill = "red"
    ) +
    geom_line(
      data = global_summary,
      aes(x = FieldSite, y = median_all, group = 1),
      color = "red", linewidth = 1.2
    ) +
    facet_wrap(~PopulationCode, scales = "fixed") +
    theme_bw() +
    labs(
      x = NULL, y = y_label,
      title = "Family-level reaction norms by population (median ± 95% CI)"
    ) +
    theme(axis.text = element_text(size = 12))
  
  
  ## -----------------------------------------
  ## Plot 3 — Mean ± SD version (optional)
  ## -----------------------------------------
  
  plot_facet_mean_with_SD <- ggplot() +
    geom_jitter(
      data = subdata,
      aes(x = FieldSite, y = .data[[y_var]]),
      width = 0.1, alpha = 0.4
    ) +
    geom_ribbon(
      data = family_summary,
      aes(x = FieldSite, ymin = lower_sd_pred, ymax = upper_sd_pred, group = Family),
      alpha = 0.05, fill = "blue"
    ) +
    geom_line(
      data = family_summary,
      aes(x = FieldSite, y = mean_pred, group = Family),
      linewidth = 0.8, color = "blue", alpha = 0.3
    ) +
    geom_line(
      data = global_summary,
      aes(x = FieldSite, y = mean_all, group = 1),
      color = "red", linewidth = 1.2
    ) +
    facet_wrap(~PopulationCode, scales = "fixed") +
    theme_bw() +
    labs(
      x = NULL, y = y_label,
      title = "Family-level fitted reaction norms (mean ± SD)"
    ) +
    theme(axis.text = element_text(size = 12))
  
  
  list(
    family_summary = family_summary,
    plot_all_fams = plot_all_fams,
    plot_facet_median_with_CI = plot_facet_median_with_CI,
    plot_facet_mean_with_SD = plot_facet_mean_with_SD
  )
}


family_reaction_norms_PC1 <- family_reaction_norms(mod=mod_PC1, subdata, y_var="DiseaseIndex_PC1", y_label = "Disease index PC1 (sign-inverted and scaled)")
family_reaction_norms_PC2 <- family_reaction_norms(mod=mod_PC2, subdata, y_var="DiseaseIndex_PC2", y_label = "Disease index PC2 (scaled)")

family_reaction_norms_PC1$plot_all_fams
family_reaction_norms_PC2$plot_all_fams
family_reaction_norms_PC1$plot_facet_median_with_CI
family_reaction_norms_PC2$plot_facet_median_with_CI
```

##### Population-level


```{r ReactionNormPlotPopulationLevelNested, fig.width=8, fig.height=8}
population_reaction_norms <- function(mod, subdata, y_var, y_label) {
  
  # Unique population × field site combinations
  valid_pops <- subdata %>% distinct(PopulationCode)
  newdata <- crossing(FieldSite = unique(subdata$FieldSite), valid_pops)
  
  # Posterior fitted values including only population-level random effects
  preds <- add_epred_draws(
    mod,
    newdata = newdata,
    re_formula = ~(0 + FieldSite | PopulationCode)
  )
  
  # Summarise posterior per population × field site
  pop_summary <- preds %>%
    group_by(PopulationCode, FieldSite) %>%
    summarise(
      median_pred      = median(.epred),
      mean_pred        = mean(.epred),
      lower_quant_pred = quantile(.epred, 0.025),
      upper_quant_pred = quantile(.epred, 0.975),
      sd_pred          = sd(.epred),
      lower_sd_pred    = mean_pred - sd_pred,
      upper_sd_pred    = mean_pred + sd_pred,
      .groups = "drop"
    )
  
  ## ------------------------------
  ## Plot 1 — All populations together
  ## ------------------------------
  plot_all_pops <- ggplot() +
    geom_jitter(
      data = subdata,
      aes(x = FieldSite, y = .data[[y_var]], color = PopulationCode),
      width = 0.1, alpha = 0.4
    ) +
    geom_ribbon(
      data = pop_summary,
      aes(x = FieldSite, ymin = lower_quant_pred, ymax = upper_quant_pred, group = PopulationCode, fill = PopulationCode),
      alpha = 0.1, show.legend = FALSE
    ) +
    geom_line(
      data = pop_summary,
      aes(x = FieldSite, y = median_pred, group = PopulationCode, color = PopulationCode),
      linewidth = 1
    ) +
    theme_bw() +
    labs(
      x = "Field Site",
      y = y_label,
      title = "Population-level reaction norms (median ± 95% CI)"
    ) +
    theme(axis.text = element_text(size = 12))
  
  ## ------------------------------
  ## Plot 2 — Faceted by population
  ## ------------------------------
  plot_facet_median_with_CI <- ggplot() +
    geom_jitter(
      data = subdata,
      aes(x = FieldSite, y = .data[[y_var]]),
      width = 0.1, alpha = 0.4
    ) +
    geom_ribbon(
      data = pop_summary,
      aes(x = FieldSite, ymin = lower_quant_pred, ymax = upper_quant_pred),
      alpha = 0.2, fill = "blue"
    ) +
    geom_line(
      data = pop_summary,
      aes(x = FieldSite, y = median_pred, group = 1),
      color = "blue", linewidth = 1
    ) +
    facet_wrap(~PopulationCode, scales = "fixed") +
    theme_bw() +
    labs(
      x = "Field Site",
      y = y_label,
      title = "Population-level reaction norms by population (median ± 95% CI)"
    ) +
    theme(axis.text = element_text(size = 12))
  
  list(
    pop_summary = pop_summary,
    plot_all_pops = plot_all_pops
  )
}

# ------------------------------
# Apply function to PC1 and PC2
# ------------------------------
pop_reaction_norms_PC1 <- population_reaction_norms(mod = mod_PC1, subdata, y_var = "DiseaseIndex_PC1_scaled", y_label = "Disease index PC1 (sign-inverted and scaled)")
pop_reaction_norms_PC2 <- population_reaction_norms(mod = mod_PC2, subdata, y_var = "DiseaseIndex_PC2_scaled", y_label = "Disease index PC2 (scaled)")

# Display plots
pop_reaction_norms_PC1$plot_all_pops
pop_reaction_norms_PC2$plot_all_pops
#pop_reaction_norms_PC1$pop_summary
```


### Extracting parameters of interest

For each response variable (sign-inverted PC1 and PC2), we used posterior expected values (`epred`) to compute, for every *Population × Family*, the difference in disease index between the two field sites (FE – FS). This difference represents the reaction-norm slope.

From these slopes, two population-level parameters were extracted:

  - `median_pop_slope`: The median reaction-norm slope across families within a population. Quantifies the average change in disease susceptibility between sites for that population.
  
  - `median_var_family_slope`: The median variance of family-specific slopes within a population. Measures the within-population variation in responses to the site contrast.

These two parameters summarize, respectively, the population-level environmental response and the heterogeneity among families in that response.

```{r ExtractParamsOfInterest}
extract_slopes_epred <- function(model, subdata) {
  
  # Get all population/family combinations
  newdata <- subdata %>% dplyr::select(FieldSite, PopulationCode, Family) %>% distinct

  # Posterior predictions including both population and family effects
  preds <- add_epred_draws(model, newdata = newdata, re_formula = ~(0 + FieldSite | PopulationCode/Family)) %>% 
    ungroup %>% 
    select(FieldSite, PopulationCode, Family, .draw, .epred)
  
  # Compute slope for each draw: FE - FS per family
  slopes <- preds %>%
    pivot_wider(
      names_from = FieldSite,
      values_from = .epred
    ) %>%
    mutate(
      fam_slope = FE -FS   # adjust to match your site names
    )
  
  # Summarize per draw & population
  slopes_by_draw <- slopes %>%
    group_by(.draw, PopulationCode) %>%
    arrange(PopulationCode) %>% 
    summarise(
      pop_slope = median(fam_slope),
      var_family_slope = if(n() > 1) var(fam_slope) else 0, # zero if only one family
      .groups = "drop"
    )
  
  # Summarize across draws
  summary_pop <- slopes_by_draw %>%
    group_by(PopulationCode) %>%
    summarise(
      median_pop_slope = median(pop_slope),
      median_var_family_slope = median(var_family_slope),
      .groups = "drop"
    )
  
  summary_pop
}

# Apply to both models
summary_PC1 <- extract_slopes_epred(mod_PC1, subdata)
summary_PC2 <- extract_slopes_epred(mod_PC2, subdata)

# Join into one table
sum_pop <- summary_PC1 %>%
  rename(median_pop_slope_PC1 = median_pop_slope,
         median_var_family_slope_PC1 = median_var_family_slope) %>%
  left_join(
    summary_PC2 %>% 
      rename(median_pop_slope_PC2 = median_pop_slope,
             median_var_family_slope_PC2 = median_var_family_slope),
    by = "PopulationCode"
  )

sum_pop %>%  kable_mydf()
```


# Mean trait values{#sec-MeanTraitValues}

We extract the mean trait values of each population in each common garden.

```{r ExtractMeanTraitValue}
# Function to compute mean phenotype per population and field site
compute_means_per_site <- function(subdata, trait) {
  subdata %>%
    group_by(PopulationCode, FieldSite) %>%
    summarise(mean_y = mean(.data[[trait]], na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(
      names_from = FieldSite,
      values_from = mean_y,
      names_prefix = paste0("mean_", trait, "_")
    )
}

# Compute means for both traits
mean_PC1 <- compute_means_per_site(subdata, "DiseaseIndex_PC1_scaled")
mean_PC2 <- compute_means_per_site(subdata, "DiseaseIndex_PC2_scaled")

# Merge with your existing summary table
sum_pop <- sum_pop %>%
  left_join(mean_PC1, by = "PopulationCode") %>%
  left_join(mean_PC2, by = "PopulationCode")
```

# RDPI{#sec-RDPI}


In this section, we calculate the relative distance plasticity index (RDPI) from @valladaresQuantitativeEstimationPhenotypic2006. In this paper, the authors propose to quantify phenotypic plasticity based on phenotypic distances among individuals of a given species (in our study, population) exposed to different environments (in our study, planted in different common gardens), which is summarized in the RDPI index that allows for statistical comparisons of phenotypic plasticity between species (in our study, populations). 

For a given trait, the relative distance plasticity index $RDPI_p$ for a given population $p$ can be calculated as follows:

$$RDPI_p = \frac{\sum_{n=1}^N{\frac{x_{ij}-x_{i'j'}}{x_{ij}+x_{i'j'}}}}{N}$$

where $N$ is the number of pairwise comparisons among individuals of the same population but from different common gardens. $x_{ij}$ and $x_{i'j'}$ are the phenotypic values of individual $j$ and $j'$ in common gardens $i$ and $i’$. As such, the phenotypic plasticity of a population for a given trait corresponds to the sum of the relative phenotypic distances for all pairs of individuals of the population grown in different common gardens. 

We also calculate the RDPI at the population level. In this case, we first compute the mean trait value of each population in each common garden, and then calculate RDPI based on these population means. The population-mean RDPI therefore summarizes overall population-level plasticity, integrating the average phenotypic shift of the population across environments while ignoring within-population variation among individuals.

Comment: Before calculating the RDPI, I shifted the trait values to ensure they are all positive because the index involves dividing by the sum of trait values across environments. When traits include negative or zero values, the denominator can become very small or change sign, producing extreme or negative RDPI values. Shifting the values preserves the relative differences among individuals while avoiding these numerical artifacts.

```{r CalcRDPI}
## ---- RDPI computation for DiseaseIndex_PC1 and PC2 ----

compute_RDPI <- function(subdata, trait) {
  
  # Rename target trait to generic 'y'
  subdata <- subdata %>% mutate(y = .data[[trait]])
  
  # Shift the trait to positive values if necessary
  trait_min <- min(subdata$y, na.rm = TRUE)
  if (trait_min <= 0) {
    shift_value <- abs(trait_min) + 1
    subdata <- subdata %>% mutate(y_shifted = y + shift_value)
  } else {
    subdata <- subdata %>% mutate(y_shifted = y)
  }

  # Identify populations and field sites
  pop_codes <- unique(subdata$PopulationCode) %>% sort()
  site_codes <- unique(subdata$FieldSite) %>% sort()
  site_comb <- combn(site_codes, 2)
  calc_rel_dist <- function(x) abs(x[1] - x[2]) / (x[1] + x[2])

  # Loop over populations
  res_list <- lapply(pop_codes, function(pop) {
    subpop <- subdata %>% filter(PopulationCode == pop)

    # --- Individual-based RDPI ---
    rel_dist_ind <- lapply(1:ncol(site_comb), function(nb_comb) {
      subsite1 <- subpop %>% filter(FieldSite == site_comb[1, nb_comb])
      subsite2 <- subpop %>% filter(FieldSite == site_comb[2, nb_comb])
      
      if (nrow(subsite1) == 0 | nrow(subsite2) == 0) return(NULL)
      
      lapply(1:nrow(subsite1), function(x) {
        rbind(subsite1[x, "y_shifted"], subsite2[, "y_shifted"]) %>%
          pull() %>%
          combn(2, calc_rel_dist) %>%
          .[1:nrow(subsite2)] %>%
          as.numeric()
      }) %>% unlist()
    }) %>% unlist()

    N_ind <- length(rel_dist_ind)
    ind_RDPI <- if (N_ind == 0) NA_real_ else sum(rel_dist_ind) / N_ind

    # --- Population-mean RDPI ---
    mean_tab <- subpop %>%
      drop_na(y) %>%
      group_by(FieldSite) %>%
      summarise(pop_mean = mean(y_shifted, na.rm = TRUE), .groups = "drop")

    rel_dist_pop_means <- lapply(1:ncol(site_comb), function(nb_comb) {
      s1 <- site_comb[1, nb_comb]
      s2 <- site_comb[2, nb_comb]
      mean1 <- mean_tab %>% filter(FieldSite == s1) %>% pull(pop_mean)
      mean2 <- mean_tab %>% filter(FieldSite == s2) %>% pull(pop_mean)
      if (length(mean1) == 0 || length(mean2) == 0) return(NA_real_)
      denom <- mean1 + mean2
      if (denom == 0) return(NA_real_) else return(abs(mean1 - mean2) / denom)
    }) %>% unlist()

    pop_mean_RDPI <- if (all(is.na(rel_dist_pop_means))) NA_real_ else mean(rel_dist_pop_means, na.rm = TRUE)

    tibble(PopulationCode = pop,
           ind_RDPI = ind_RDPI,
           pop_mean_RDPI = pop_mean_RDPI)
  })

  df_rdpi <- bind_rows(res_list) %>%
    mutate(across(c(ind_RDPI, pop_mean_RDPI), as.numeric))

  # Rename columns with trait suffix
  colnames(df_rdpi)[2:3] <- paste0(colnames(df_rdpi)[2:3], "_", trait)
  
  return(df_rdpi)
}

# ---- Apply function to both traits ----
df_rdpi_PC1 <- compute_RDPI(subdata, "DiseaseIndex_PC1_scaled")
df_rdpi_PC2 <- compute_RDPI(subdata, "DiseaseIndex_PC2_scaled")

# ---- Merge all results ----
sum_pop <- sum_pop %>%
  left_join(df_rdpi_PC1, by = "PopulationCode") %>%
  left_join(df_rdpi_PC2, by = "PopulationCode")

# Final summary
#saveRDS(sum_pop, here("outputs/ScotsPine/PhenotypicPlasticity/DiseaseSusceptibilityIndexesSummary.rds"))
sum_pop %>% kable_mydf()
```

# Relationships among all the estimated indexes{#sec-IndexRelationships}

Description for each variable:

  - `PopulationCode`: Identifier of the population.

  - `median_pop_slope_PC1`: Median reaction-norm slope (FE – FS) for PC1, summarizing the population-level difference in disease susceptibility between field sites.

  - `median_var_family_slope_PC1`: Median variance among family-specific slopes for PC1, reflecting within-population heterogeneity in responses.

  - `median_pop_slope_PC2`: Median reaction-norm slope for PC2, summarizing the population-level difference in disease index between field sites.

  - `median_var_family_slope_PC2`: Median variance among family-specific slopes for PC2, indicating within-population variation in responses.

  - `mean_FE_PC1`: Mean PC1 disease index for the population in the FE site.

  - `mean_FS_PC1`: Mean PC1 disease index for the population in the FS site.

  - `mean_FE_PC2`: Mean PC2 disease index for the population in the FE site.

  - `mean_FS_PC2`: Mean PC2 disease index for the population in the FS site.

  - `ind_RDPI_PC1`: Individual-based RDPI for PC1, quantifying relative phenotypic change across sites among individuals within the population.

  - `pop_mean_RDPI_PC1`: RDPI computed from population means for PC1, summarizing average phenotypic divergence across sites.

  - `ind_RDPI_PC2`: Individual-based RDPI for PC2, quantifying relative phenotypic change across sites among individuals within the population.

  - `pop_mean_RDPI_PC2`: RDPI computed from population means for PC2, summarizing average phenotypic divergence across sites.


Note that PC1 and PC2 corresponds to the sign-inverted and scaled PC1 component and scaled PC2 component of @sec-SynthDiseaseIndex.
```{r}
# Rename columns so FE/FS comes first
sum_pop <- sum_pop %>%
  rename(
    mean_FE_PC1 = mean_DiseaseIndex_PC1_scaled_FE,
    mean_FS_PC1 = mean_DiseaseIndex_PC1_scaled_FS,
    mean_FE_PC2 = mean_DiseaseIndex_PC2_scaled_FE,
    mean_FS_PC2 = mean_DiseaseIndex_PC2_scaled_FS,
    ind_RDPI_PC1 = ind_RDPI_DiseaseIndex_PC1_scaled,
    ind_RDPI_PC2 = ind_RDPI_DiseaseIndex_PC2_scaled,
    pop_mean_RDPI_PC1 = pop_mean_RDPI_DiseaseIndex_PC1_scaled,
    pop_mean_RDPI_PC2 = pop_mean_RDPI_DiseaseIndex_PC2_scaled,
    
  )

sum_pop %>% kable_mydf
```

## PCA

```{r PCAonCalaculateIndexes, fig.height=8, fig.width=8}
# keep only populations with complete data for PCA (or use na.omit)
pca_df <- sum_pop %>% drop_na()

# run PCA (center & scale)
pca_res <- prcomp(pca_df %>% select(-PopulationCode), center = TRUE, scale. = TRUE)

# scores (PC coordinates for populations)
scores <- as.data.frame(pca_res$x)
scores$PopulationCode <- pca_df$PopulationCode

# loadings (variable contributions)
loadings <- as.data.frame(pca_res$rotation)
loadings$var <- rownames(loadings)

# get % variance explained for axis labels
eig <- (pca_res$sdev)^2
pct_var <- round(100 * eig / sum(eig), 1)

# scale loadings to make arrows visible on the same plotting scale as scores
# choose a multiplier so arrows reach ~30-50% of the score cloud
mult <- min(
  (max(scores$PC1) - min(scores$PC1)) / (max(loadings$PC1) - min(loadings$PC1)),
  (max(scores$PC2) - min(scores$PC2)) / (max(loadings$PC2) - min(loadings$PC2))
) * 0.6

loadings_plot <- loadings %>%
  mutate(arrow_x = PC1 * mult,
         arrow_y = PC2 * mult)

# create a color palette with one color per population
n_pops <- nrow(scores)
pal <- scales::hue_pal()(n_pops) # distinct hues; change palette if you prefer

# Plot
ggplot() +
  # points for populations
  geom_point(data = scores, aes(x = PC1, y = PC2, color = PopulationCode), size = 3, alpha = 0.9) +
  # population labels
  geom_text_repel(data = scores, aes(x = PC1, y = PC2, label = PopulationCode, color = PopulationCode),
                  size = 3.5, show.legend = FALSE, max.overlaps = 100) +
  # arrows for variable loadings (from origin to scaled loading)
  geom_segment(data = loadings_plot,
               aes(x = 0, y = 0, xend = arrow_x, yend = arrow_y),
               arrow = arrow(length = unit(0.25, "cm")), colour = "black", linewidth = 0.6) +
  # variable labels at arrow tips
  geom_text_repel(data = loadings_plot,
                  aes(x = arrow_x, y = arrow_y, label = var),
                  color = "black", size = 4, segment.size = 0.3) +
  scale_color_manual(values = pal) +
  labs(x = paste0("PC1 (", pct_var[1], "%)"),
       y = paste0("PC2 (", pct_var[2], "%)"),
       title = "",
       color = "Population") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none") # legend suppressed because each pop has its own color label

```

## Correlations 

```{r PCAonCalculatedPhenoIndexes, fig.height=8, fig.width=8}
 sum_pop %>% 
  dplyr::select(-PopulationCode) %>% 
  cor() %>% 
  corrplot(method = 'number',
           type = 'lower', diag = FALSE,
           title=paste0(""),
           mar=c(0,0,2,0),
           number.cex=0.9,tl.cex=0.9)
```


## Box plots

```{r BoxPlots, fig.height=10, fig.width=10}
# Pivot to long format
sum_pop_long <- sum_pop %>%
  pivot_longer(
    cols = -PopulationCode,
    names_to = c("variable", "PC"),
    names_pattern = "(.*_FE|.*_FS|.*)_PC(\\d+)$",
    values_to = "value"
  )

# --- ensure correct variable grouping ---
sum_pop_long$variable <- factor(sum_pop_long$variable,
                                levels = c("median_pop_slope",
                                           "median_var_family_slope",
                                           "mean_FS",
                                           "mean_FE",
                                           "ind_RDPI",
                                           "pop_mean_RDPI"))

# --- boxplot with points behind ---
p_all <- ggplot(sum_pop_long, aes(x = variable, y = value, fill = PC)) +
  geom_jitter(aes(color = PC), 
              width = 0.15, alpha = 0.5, size = 2, show.legend = FALSE) +  # points behind boxes
  geom_boxplot(alpha = 0.7, width = 0.6, outlier.shape = NA, color = "black") +
  scale_fill_manual(values = c("1" = "#1f77b4", "2" = "#d62728"),
                    name = "Principal Component",
                    labels = c("PC1", "PC2")) +
  scale_color_manual(values = c("1" = "#1f77b4", "2" = "#d62728")) +
  theme_bw(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  ) +
  labs(
    x = "Variable type",
    y = "Value",
    title = "Distribution of population-level estimates across variables and PCs"
  )

# Split long table into two sets of variables
vars_means <- c("mean_FS", "mean_FE", "median_pop_slope")
vars_other <- c("median_var_family_slope", "ind_RDPI", "pop_mean_RDPI")

df_means <- sum_pop_long %>% filter(variable %in% vars_means)
df_other <- sum_pop_long %>% filter(variable %in% vars_other)

# --- Plot 1: mean_FE / mean_FS ---
p_means <- ggplot(df_means, aes(x = variable, y = value, fill = PC)) +
  geom_jitter(aes(color = PC), width = 0.15, alpha = 0.5, size = 2, show.legend = FALSE) +
  geom_boxplot(alpha = 0.7, width = 0.6, outlier.shape = NA, color = "black") +
  scale_fill_manual(values = c("1" = "#1f77b4", "2" = "#d62728"),name = "Disease Index", labels=c("PC1","PC2")) +
  scale_color_manual(values = c("1" = "#1f77b4", "2" = "#d62728")) +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.minor = element_blank()) +
  labs(x = "", y = "Index value", title = "Population median slope (FE - FS) and population mean trait values per field site")

# --- Plot 2: Other variables ---
p_other <- ggplot(df_other, aes(x = variable, y = value, fill = PC)) +
  geom_jitter(aes(color = PC), width = 0.15, alpha = 0.5, size = 2, show.legend = FALSE) +
  geom_boxplot(alpha = 0.7, width = 0.6, outlier.shape = NA, color = "black") +
  scale_fill_manual(values = c("1" = "#1f77b4", "2" = "#d62728"),name = "Disease Index", labels=c("PC1","PC2")) +
  scale_color_manual(values = c("1" = "#1f77b4", "2" = "#d62728")) +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.minor = element_blank()) +
  labs(x = "", y = "Index value", title = "Population-level proxies of disease susceptibility plasticity")

# --- Combine plots vertically ---
combined_plot <- plot_grid(p_means, p_other, ncol = 1, align = "v", rel_heights = c(1, 1.2))
# Show the combined plot
print(combined_plot)
```


# Correlations with genomic offset predictions{#sec-CorrWithGenomicOffset}

```{r DownloadGOpred}
df_go_clim_dist <- readRDS(here("outputs/ScotsPine/df_gen_off_pred_clim_dist.rds"))

df_go_wide <- df_go_clim_dist %>%
  select(PopulationCode, rcp_scenario, pred_timeslice, method, ensemble_member, var_code,val) %>%
  mutate(var_name_comb = paste(rcp_scenario, pred_timeslice, method, ensemble_member, var_code,sep = "_") %>%  gsub("[ .-]", "", .)) %>%
  dplyr::select(PopulationCode, var_name_comb,val) %>% 
  pivot_wider(names_from = var_name_comb, values_from = val)


df_go_wide <- df_go_clim_dist %>%
  dplyr::filter(var_code == "all_snps" & rcp_scenario == "RCP2.6" & pred_timeslice == "2030-2050" & ensemble_member == "Average across EMs") %>% 
  select(PopulationCode, method, val) %>%
  mutate(var_name_comb = paste(method, sep = "_") %>%  gsub("[ .-]", "", .)) %>%
  dplyr::select(PopulationCode, var_name_comb, val) %>% 
  pivot_wider(names_from = var_name_comb, values_from = val)
```


```{r PCAwithGO, fig.height=8, fig.width=8}
# keep only populations with complete data for PCA (or use na.omit)
pca_df <- sum_pop %>% left_join(df_go_wide, by="PopulationCode")

# run PCA (center & scale)
pca_res <- prcomp(pca_df %>% select(-PopulationCode), center = TRUE, scale. = TRUE)

# scores (PC coordinates for populations)
scores <- as.data.frame(pca_res$x)
scores$PopulationCode <- pca_df$PopulationCode

# loadings (variable contributions)
loadings <- as.data.frame(pca_res$rotation)
loadings$var <- rownames(loadings)

# get % variance explained for axis labels
eig <- (pca_res$sdev)^2
pct_var <- round(100 * eig / sum(eig), 1)

# scale loadings to make arrows visible on the same plotting scale as scores
# choose a multiplier so arrows reach ~30-50% of the score cloud
mult <- min(
  (max(scores$PC1) - min(scores$PC1)) / (max(loadings$PC1) - min(loadings$PC1)),
  (max(scores$PC2) - min(scores$PC2)) / (max(loadings$PC2) - min(loadings$PC2))
) * 0.6

loadings_plot <- loadings %>%
  mutate(arrow_x = PC1 * mult,
         arrow_y = PC2 * mult)

# create a color palette with one color per population
n_pops <- nrow(scores)
pal <- scales::hue_pal()(n_pops) # distinct hues; change palette if you prefer

# Plot
ggplot() +
  # points for populations
  geom_point(data = scores, aes(x = PC1, y = PC2, color = PopulationCode), size = 3, alpha = 0.9) +
  # population labels
  geom_text_repel(data = scores, aes(x = PC1, y = PC2, label = PopulationCode, color = PopulationCode),
                  size = 3.5, show.legend = FALSE, max.overlaps = Inf) +
  # arrows for variable loadings (from origin to scaled loading)
  geom_segment(data = loadings_plot,
               aes(x = 0, y = 0, xend = arrow_x, yend = arrow_y),
               arrow = arrow(length = unit(0.25, "cm")), colour = "black", linewidth = 0.6) +
  # variable labels at arrow tips
  geom_text_repel(data = loadings_plot,
                  aes(x = arrow_x, y = arrow_y, label = var),
                  color = "black", size = 4, segment.size = 0.3, max.overlaps = Inf) +
  scale_color_manual(values = pal) +
  labs(x = paste0("PC1 (", pct_var[1], "%)"),
       y = paste0("PC2 (", pct_var[2], "%)"),
       title = "",
       color = "Population") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none") # legend suppressed because each pop has its own color label

```
 
```{r CorrelationsWithGO, fig.height=10, fig.width=10}
sum_pop %>% left_join(df_go_wide, by="PopulationCode") %>% 
  dplyr::select(-PopulationCode) %>% 
  cor() %>% 
  corrplot(method = 'number',
           type = 'lower', diag = FALSE,
           title=paste0(""),
           mar=c(0,0,2,0),
           number.cex=0.9,tl.cex=0.9)
```

There are no or very weak correlations between genomic offset predictions and plasticity indices. A weak correlation was observed between some genomic offset predictions (mostly pRDA and RDA) and population variance in family-level slopes for PC2 (`median_var_family_slope_PC2`) and the individual-level RDPI for PC2 (`ind_RDPI_PC2`).

Population variance in family-level slopes for PC1 (`median_var_family_slope_PC1`) is very low, and this variable is not correlated with other plasticity indices: `ind_RDPI_PC1`, `pop_mean_RDPI_PC1`, or `median_pop_slope_PC1.`

This is not the case for PC2, where all plasticity indices are correlated: `ind_RDPI_PC2`, `pop_mean_RDPI_PC2`, `median_pop_slope_PC2`, and `median_var_family_slope_PC2.`

