---
title: "Plastic responses in phenology at the population level"
author: "Juliette Archambeau"
date: last-modified
format: 
  html:
    toc: true
    toc-depth: 4
    code-fold: true
    page-layout: full
    number-sections: true
embed-resources: true
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

<style type="text/css">
body {
   font-size: 15px;
}
code.r{
  font-size: 11px;
}
pre {
  font-size: 11px
}

table {
  font-size: 11px
}
</style>

```{r setup, include=F}
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))
knitr::opts_chunk$set(cache=F)
options(width = 300)
library(knitr)
library(kableExtra)
library(tidyverse)
library(here)
library(magrittr)
library(corrplot)
library(readxl)
library(rstan)
library(tidybayes)
library(bayesplot)
library(ggplot2)
library(parallel)
library(tictoc)
library(latex2exp)
library(brms)
library(Reacnorm)
library(matrixStats)
library(posterior)
library(missMDA)
library(ggrepel)

rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

theme_set(theme_bw(base_size = 20))

# my own function for building tables in reports
source(here("scripts/functions/kable_mydf.R"))
```

# Introduction

The goal of this report is to determine whether the populations have different plastic responses for phenology. 

In @sec-SynthPhenoIndex, we summarized the phenology-related variables into a single index. We performed a PCA on the phenology-related variables and extracted the first principal component (PC1), which we used as a synthetic trait capturing overall phenological timing. We considered two options: (i) including all phenology variables (timing of budburst and budburst duration) or (ii) including only variables related to budburst timing. In the current version of the report, I decided to include only the variables related to the timing of budburst in the PCA for two reasons: (i) I considered that information about the duration of budburst is already partly captured by the variables related to budburst timing (since the variables related to budburst duration are derived from the variables related to budburst timing if I well remember, <span style="color: orange;"> is it right?</span> ), and (ii) the first PC from the PCA including only budburst timing variables explained more variation than the first PC from the PCA including all phenology variables. <span style="color: orange;">Do you agree with this step or would you prefer to use all the phenology-related variables in the PCA? (to discuss!).</span> 

To account for nursery effects (i.e., the fact that trees planted in some common gardens originated from different nurseries), we estimated plasticity to environmental variation arising from the combined effects of site and nursery. Consequently, we restricted the analysis to the two common gardens where all trees originated from a single nursery: Yair (FS), where all trees came from NG, and Glensaugh (FE), where all trees came from NE. In Glensaugh, four trees originated from NG; these were excluded from the dataset used to estimate plasticity.

In section @sec-SPmodelGRM, we applied the approach developed by Villemereuil & Chevin (2025) to assess whether the heritability of the reaction norm ($h_{RN}^2$) primarily stems from the environment-blind heritability of the trait ($h^2$, based on breeding values averaged across environments) or from the heritability of plasticity ($h_I^2$, based on breeding values by environment interaction). We implemented this approach at the species level, rather than at the population level, because it requires a large amount of data. Even at the species level, the model required many iterations for the Markov chains to converge, and some warnings still occurred. <span style="color: orange;">I plan to test the use of informative priors to improve convergence, but this remains to be done.</span> 

In @sec-ModelVaryingSlopes, plastic responses were estimated as population-level varying slopes (@sec-SPmodelPopVaryingSlopes) and family-level varying slopes (@sec-SPmodelFamVaryingSlopes). In @sec-SPmodelPopVaryingSlopes, our objective was to determine whether populations differ in their plastic responses across sites, as captured by the variation in slopes. In @sec-SPmodelFamVaryingSlopes, we estimated the variation in the plastic responses among families within populations (i.e., variance in family-level slopes within each population).

In @sec-RDPI, we quantified phenotypic plasticity using the relative distance plasticity index (RDPI; Valladares et al., 2006), which measures the mean relative phenotypic distance among individuals of the same population grown in different environments. We calculated RDPI in two ways: (i) at the individual level (i.e., the common practice), where all pairwise phenotypic distances between individuals of a given population and from different common gardens are considered, and (ii) using population means per site, which summarizes differences in average population responses across sites.

In @sec-IndexRelationships, we estimated the relationships among all these indices—including the mean trait values averaged within each common garden (@sec-MeanTraitValues)—using a principal component analysis (PCA) and a correlation matrix.

In @sec-CorrWithGenomicOffset, we estimated the relationships between our phenology indexes with the predicted genomic offsets. We used genomic offset predictions from the RCP scenario 2.6, the prediction time period 2030-2050 and the average across the four climatic models (ensemble members).


# A synthetic trait capturing phenology{#sec-SynthPhenoIndex}


We load the phenotypic data from the common gardens and filter for either the common garden FS combined with the NG nursery, or the common garden FE and the nursery NE.

```{r LoadPhenoCGData}
data <- readRDS(file=here("data/ScotsPine/formatted_CG_data.rds")) %>% # 1848 obs
  filter(FieldSite %in% c("FE","FS")) %>% # 1344 obs
  filter(!(FieldSite=="FE" & Nursery =="NG")) # 1340 obs
```

The dataset contains `r length(unique(data$FieldCode))` individuals, `r length(unique(data$PopulationCode))` populations, `r length(unique(data$Family))` families. Here are the first 10 rows:

```{r VizData, fig.height=12, fig.width=12}
data[1:10,] %>% kable_mydf()

# Correlation matrix plot for phenotypic variables
data %>%
  dplyr::select(contains("BT"), contains("BD")) %>%
  cor(use = "pairwise.complete.obs") %>%
  corrplot::corrplot(
    method = "color",
    type = "lower",
    order = "hclust",
    addCoef.col = "black",
    tl.col = "black",
    tl.srt = 45,
    col = colorRampPalette(c("#6D9EC1", "white", "#E46726"))(200),
    title = "Correlation among phenotypic variables",
    mar = c(0,0,2,0)
  )
```

We build a single synthetic trait that captures overall phenological timing.

We consider two options: either (i) including all phenology variables (timing of budburst and budburst duration) or (ii) including only the variables related to budburst timing. We run a PCA for both sets of variables and then choose which to retain for the subsequent analyses, using the first principal component as a phenology index for each tree.

<span style="color: orange;">We imputed missing values (NAs) in the phenology dataset to ensure that all individuals could be included in the principal component analysis (PCA), as PCA requires complete data. The imputation was performed using the regularized iterative PCA algorithm implemented in the `imputePCA()` function from the missMDA package. This method estimates the missing entries based on the underlying correlation structure among variables, iteratively refining the estimates until convergence. The optimal number of components for imputation was determined with `estim_ncpPCA()`, which minimizes the prediction error of the imputed values. This approach preserves the multivariate relationships among phenological traits and avoids biases that would result from listwise deletion of incomplete observations.</span> 

```{r runningPhenoPCA, fig.height=6, fig.width=6}
# Simple function wrapping your code
compute_pheno_pca <- function(data, var_pattern, pc1_name = NULL) {
  # create TreeID if not present
  if (!"TreeID" %in% colnames(data)) {
    data <- data %>% mutate(TreeID = paste0(Family, "_", Block, "_", FieldCode))
  }
  
  # select phenology variables
  pheno <- data %>%
    column_to_rownames("TreeID") %>%
    select(matches(var_pattern))
  
  # NA check (kept here for info — printed)
  # print(colSums(is.na(pheno)))
  
  # Estimate optimal number of components for imputation
  ncp <- estim_ncpPCA(pheno, method = "Regularized")$ncp
  
  # Impute missing values
  imp <- imputePCA(pheno, ncp = ncp)
  pheno_imp <- imp$completeObs
  
  # PCA on imputed data
  pca_imp <- prcomp(pheno_imp, scale. = TRUE)
  #print(summary(pca_imp))
  
  # decide PC1 column name
  if (is.null(pc1_name)) {
    safe_pattern <- gsub("\\W+", "_", var_pattern)
    pc1_name <- paste0("PhenoIndex_PC1_", safe_pattern)
  }
  
  # Add PC1 as phenology index
  data[[pc1_name]] <- pca_imp$x[, 1]
  
  # Extract PCA scores (individuals) and loadings (variables)
  scores_df <- as.data.frame(pca_imp$x) %>%
    mutate(TreeID = rownames(pca_imp$x)) %>%
    left_join(data, ., by = "TreeID")
  
  loadings_df <- as.data.frame(pca_imp$rotation) %>%
    rownames_to_column(var = "Trait")
  
  # Scale loadings for plotting as arrows
  scale_factor <- min(
    (max(scores_df$PC1) - min(scores_df$PC1)) / (max(loadings_df$PC1) - min(loadings_df$PC1)),
    (max(scores_df$PC2) - min(scores_df$PC2)) / (max(loadings_df$PC2) - min(loadings_df$PC2))
  ) * 0.6
  
  loadings_df <- loadings_df %>%
    mutate(arrow_x = PC1 * scale_factor,
           arrow_y = PC2 * scale_factor)
  
  # Plot: individuals as points, coloured by FieldSite, arrows for variables
  p <- ggplot() +
    geom_point(data = scores_df, aes(x = PC1, y = PC2, color = FieldSite), alpha = 0.7, size = 2) +
    geom_segment(data = loadings_df,
                 aes(x = 0, y = 0, xend = arrow_x, yend = arrow_y),
                 arrow = arrow(length = unit(0.3, "cm")), colour = "black", linewidth = 0.6) +
    geom_text_repel(data = loadings_df,
                    aes(x = arrow_x, y = arrow_y, label = Trait),
                    size = 3.5,
                    max.overlaps = 20) +
    labs(x = paste0("PC1 (", round(100 * summary(pca_imp)$importance[2,1], 1), "%)"),
         y = paste0("PC2 (", round(100 * summary(pca_imp)$importance[2,2], 1), "%)"),
         color = "Site") +
    theme_classic(base_size = 14) +
    theme(legend.position = "right")
  
  #print(p)
  
  centroids <- scores_df %>%
    group_by(PopulationCode, FieldSite) %>%
    summarise(PC1 = mean(PC1, na.rm = TRUE),
              PC2 = mean(PC2, na.rm = TRUE),
              n = n(), .groups = "drop")
  
  p_centroids <- ggplot() +
    geom_point(data = scores_df, aes(x = PC1, y = PC2), color = "grey80", alpha = 0.4, size = 1) +
    geom_point(data = centroids, aes(x = PC1, y = PC2, color = FieldSite), size = 3) +
    geom_text_repel(data = centroids, aes(x = PC1, y = PC2, label = PopulationCode), size = 3, max.overlaps = Inf) +
    geom_segment(data = loadings_df,
                 aes(x = 0, y = 0, xend = arrow_x, yend = arrow_y),
                 arrow = arrow(length = unit(0.3, "cm")), colour = "black", linewidth = 0.6) +
    geom_text_repel(data = loadings_df, aes(x = arrow_x, y = arrow_y, label = Trait), size = 3.5) +
    labs(x = paste0("PC1 (", round(100 * summary(pca_imp)$importance[2,1], 1), "%)"),
         y = paste0("PC2 (", round(100 * summary(pca_imp)$importance[2,2], 1), "%)")) +
    theme_minimal(base_size = 13)
  
  #print(p_centroids)
  
  # Return a simple list with relevant objects
  list(
    data = data,
    pca_imp = pca_imp,
    scores_df = scores_df,
    loadings_df = loadings_df,
    plot = p,
    plot_centroids = p_centroids,
    pc1_name = pc1_name
  )
}

# Apply to both sets:
# 1) all phenology variables (BD|BT)
res_all <- compute_pheno_pca(data = data, var_pattern = "BD|BT", pc1_name = "PhenoIndex_PC1_all")

# 2) only budburst timing variables (BT)
res_bt  <- compute_pheno_pca(data = data, var_pattern = "BT",    pc1_name = "PhenoIndex_PC1_BT")
```

## PCA with all phenology variables

```{r PCASynthIndexWithAllPhenoVariables, fig.height=6, fig.width=6}
print(res_all$plot)
```

## PCA with budburst timing variables only

```{r PCASynthIndexWithBudburstTimingVariables, fig.height=6, fig.width=6}
print(res_bt$plot)
```

We decide to use the phenology index `PhenoIndex_PC1_BT` calculated on variables related to budburst timing only. Populations with higher `PhenoIndex_PC1_BT` show earlier phenology.

```{r SelectedPhenoIndex, message=F}
data <- res_all$data %>% 
  left_join(res_bt$data)

trait <- "PhenoIndex_PC1_BT"
```


# Species level model with GRM{#sec-SPmodelGRM}


We partitioned the variance of the trait of interest at the species level, i.e., with all populations combined. For that, we follow the methodology in [de Villemereuil & Chevin (2025)](https://doi.org/10.24072/pcjournal.515) to partition the phenotypic variance of the traits of interest. Below are sections extracted from the paper:

$$V_P =  V_{Plas} + V_{Gen} + V_{Res}$$

$V_P$ is the total phenotypic variance of the trait.
$V_{Plas}$ is the phenotypic variance arising from changes in the mean reaction norm across environments.
$V_{Gen}$ is the total genetic variance in the trait across environments.
$V_{Res}$ is the residual variance, not explained by the reaction norm.

$$V_{Gen} = V_{Add} + V_{nonAdd}$$

$V_{Add}$ is the total additive genetic variance in the trait across environments.
$V_{nonAdd}$ is the non additive genetic variance in the trait.

$$V_{Add} =  V_A + V_{AxE}$$

$V_A$ is the environment-blind additive genetic variance of the trait, i.e., based on the mean breeding values across environments.
$V_{AxE}$ is the additive genetic variance arising from plasticity, i.e., variance of the mean-centered breeding values.

From this, it is possible to derive unitless quantities of interest.

$P_{RN}^2 = V_{Plas}/V_P$ is the proportion of the phenotypic variance arising from average plastic responses to environments.

Variance-standardised additive genetic variances are heritabilities:

$h_{RN}^2=\frac{V_{Add}}{V_P}=\frac{V_A}{V_P}+\frac{V_{AxE}}{V_P}=h^2+h^2_I$

In other words, the heritability of the trait when fully accounting for its reaction norm ($h_{RN}^2$) is equal to the environment-blind heritability of the trait ($h^2$, based on the breeding values averaged across environments) plus the heritability from plasticity ($h_I^2$, based on the breeding values by environment interaction).

$T_{RN}^2=\frac{V_{Plas}+V_{Gen}}{V_P}=P_{RN}^2+h_{RN}^2$

$T_{RN}^2$ is the proportion of the phenotypic variance explained by the (possibly plastic and genetically variable) reaction norm, and thus our ability to predict the individual phenotype from the genotype and the environment.

To apply this methodology, we followed the tutorials associated with the `Reacnorm` R package: <https://github.com/devillemereuil/Reacnorm>.

## Loading the GRM

We load the genomic relationship matrix. Here are the first 10 rows and columns.

```{r LoadGRM}
GRM <- readRDS(file=here("data/ScotsPine/GenomicData/data_update_Fev2025/formatted_data/Gmat_VanRaden.rds"))
GRM[1:10,1:10]
```

Matrix dimensions:

```{r}
dim(GRM)
```


```{r FilterDataAndFormatGRM}
subdata <- data %>% 
  dplyr::select(FieldSite,Block,FieldCode,PopulationCode,Family,Nursery,all_of(trait)) %>% 
  drop_na() %>% # no NAs
  filter((FieldCode%in%rownames(GRM))) %>%  # 1340 to 1297
  dplyr::rename(y=any_of(trait)) %>% 
  arrange(FieldCode)

# filter the grm with the remainng trees
GRM <-  GRM[as.character(subdata$FieldCode), as.character(subdata$FieldCode), drop = FALSE]
# dim(GRM)
# GRM[1:10,1:10]
# 
# all.equal(as.character(subdata$FieldCode),colnames(GRM))
# all.equal(as.character(subdata$FieldCode),rownames(GRM))
```


```{r RunTheSpeciesLevelModelWithGRM, eval=F}
n_chains <- 4 # number of chains (MCMC)
n_iter <- 6000 # number of iterations
n_warm <- 3000 # number of iterations in the warm-up phase
n_thin <- 1 # thinning interval
n_seed <- 404

form <- brmsformula(y ~ 0 + FieldSite + (0 + FieldSite | gr(FieldCode, cov = GRM)))

mod <- brm(formula = form,
           data = subdata,
           data2 = list(GRM=GRM),
           save_pars = save_pars(group = FALSE),
           chains = n_chains,
           cores = n_chains,
           seed = n_seed,
           iter = n_iter,
           backend = "cmdstanr",
           warmup = n_warm)

saveRDS(mod, file = here(paste0("outputs/ScotsPine/StanModels/",trait,"_TwoCG_SpeciesLevelModel_PartVarVillemereuil.rds")))
```

Model summary:
  
```{r InspectTheSpeciesLevelModelWithGRM, fig.height=8, fig.width=10, eval=T}
mod <- readRDS(here(paste0("outputs/ScotsPine/StanModels/",trait,"_TwoCG_SpeciesLevelModel_PartVarVillemereuil.rds")))
summary(mod)
plot(mod, variable=c("b_FieldSiteFE", "b_FieldSiteFS","sigma"))
plot(mod, variable=c("cor_FieldCode__FieldSiteFE__FieldSiteFS", "sd_FieldCode__FieldSiteFE","sd_FieldCode__FieldSiteFS"))
```

```{r PartVarSpeciesLevelModelWithGRM, fig.height=5, fig.width=10, warning=F, eval=T}
# Getting the mean uncertainty on the fixed-effect parameters
var_uncert <- vcov(mod) |> # extracts the variance-covariance matrix of fixed effects.
  diag() |> # diagonal values represent the variances of each fixed effect parameter.
  mean()  # Calculate mean uncertainty across parameters

# Computing V_plas (variance of the average reaction norm)
post_theta_cs <- fixef(mod, summary = FALSE) |> # extracts posterior samples for fixed effects
  as_draws_df() # formats them as a posterior draws object.

# Select fixed effects corresponding to the sites and compute their row-wise variances.
var_plas_cs <- post_theta_cs |> 
  select(starts_with("FieldSite")) |>
  as.matrix() |> 
  rowVars()

# Correcting V_plas for uncertainty
post_cs <- data.frame(V_Plas = var_plas_cs - var_uncert) |>  # subtract the mean parameter uncertainty (var_uncert) from the raw V_plas values.
  cbind(select(post_theta_cs, starts_with("."))) |>  # combine with MCMC metadata (e.g., chain, iteration, draw) to create a valid posterior object.
  as_draws_df()

# Getting the G-matrix (variance-covariance of random effects)
post_cs[["G"]] <- VarCorr(mod, summary = FALSE)[["FieldCode"]][["cov"]] |> # extracts variance-covariance matrices of random effects
  apply(1, \(mat_) { mat_ }, simplify = FALSE) # apply() to structure the posterior samples into a list of matrices for further processing

# Getting the residual variance
# Extract posterior samples for residual variance
post_cs[["V_R"]] <- VarCorr(mod, summary = FALSE)[["residual__"]][["sd"]][, 1]^2

# Thinning the posterior draws for computational efficiency
# Thin the posterior draws to reduce computation time while maintaining sufficient information
post_cs <- thin_draws(post_cs, thin = length(var_plas_cs) / 1000)

# Save MCMC metadata (iteration, chain, etc.) for further use
post_cs_info <- select(post_cs, starts_with("."))

# Summarising V_Plas (posterior estimates and diagnostics)
# Display summary statistics for V_Plas (mean, median, credible intervals).
summarise_draws(subset_draws(post_cs, variable = "V_Plas"))  %>% kable_mydf()

# Computing genetic variance decomposition
# Apply function to compute genetic variances from the G-matrix
post_gen_cs <- 
  map(post_cs[["G"]], rn_cs_gen, .progress = TRUE) |> 
  bind_rows() |> 
  select(where(\(col_) { abs(mean(col_)) > 10^-5 })) |>  # Remove negligible components
  cbind(post_cs_info) |> 
  as_draws_df()

# Summarise genetic variance components
summarise_draws(post_gen_cs) %>% kable_mydf()

# Compute total phenotypic variance and variance-standardised estimates
post_var_tpc_cs <- bind_draws(post_cs, post_gen_cs) |> 
  subset_draws(variable = c("V_Plas", "V_Add", "V_A", "V_AxE", "V_R")) |> 
  mutate_variables(V_Tot = V_Plas + V_Add + V_R)

# Compute standardised variance estimates (P2, H2_RN, H2, H2_I, T2)
post_std_tpc_cs <- post_var_tpc_cs |> 
  transmute(
    P2_RN = V_Plas / V_Tot,       # Proportion of phenotypic variance due to plasticity
    H2_RN = V_Add / V_Tot,        # Heritability of reaction norm
    H2 = V_A / V_Tot,             # Heritability of environment-blind genetic variance
    H2_I = V_AxE / V_Tot,         # Heritability of genetic variance due to plasticity
    T2_RN = (V_Plas + V_Add) / V_Tot # Total variance explained by genetic and plasticity components
  ) |> 
  cbind(post_cs_info) |> 
  as_draws_df()

# Summarising standardised variance estimates
summarise_draws(post_std_tpc_cs) %>% kable_mydf()
```

MCMC traces of the parameters of interest:

```{r MCMCtracesSpeciesLevelModelWithGRM, fig.height=6, fig.width=10, eval=T}
mcmc_trace(post_std_tpc_cs)
```

Posterior distributions of the parameters of interest:

```{r MCMCAreasSpeciesLevelModelWithGRM, fig.height=5, fig.width=5, eval=T}
mcmc_areas(post_std_tpc_cs, prob = 0.95, area_method = "scaled height")
```

<span style="color: orange;">Main take-home message:</span> At the species level, the heritability of the reaction norm ($h_{RN}^2$) mainly comes from the environment-blind heritability of the trait ($h^2$, based on the breeding values averaged across environments) and not from the heritability from plasticity ($h_I^2$, based on the breeding values by environment interaction).



# Species level model with varying slopes{#sec-ModelVaryingSlopes}

In this section, we estimate population or family varying slopes (i.e., random slopes) to quantify the mean change in the trait of interest across two different sites (Yair and Glensaugh) for each population (or family). While this estimate does not indicate the genetic variation in reaction norms within populations, it allows for comparisons of the mean plastic responses across sites among populations/families. 

## Population varying slopes{#sec-SPmodelPopVaryingSlopes}


```{r SubsetDataSpeciesLevelModelRandomSlopes}
subdata <- data %>% 
  dplyr::select(FieldSite,Block,FieldCode,PopulationCode,Family,Nursery,all_of(trait)) %>% 
  drop_na() %>% # no NAs
  dplyr::rename(y=any_of(trait)) %>% 
  arrange(FieldCode) %>% 
  mutate(FieldSite_num = as.numeric(as_factor(FieldSite)))
```

```{r RunSpeciesLevelModelRandomSlopes, eval=F}
# Model options
n_chains <- 4 # number of chains (MCMC)
n_iter <- 8000 # number of iterations
n_warm <- 4000 # number of iterations in the warm-up phase
n_thin <- 1 # thinning interval
n_seed <- 404

form <- brmsformula(y ~ 1 + FieldSite_num + (1 + FieldSite_num | PopulationCode))

mod <- brm(formula = form,
           data = subdata,
           chains = n_chains,
           cores = n_chains,
           seed = n_seed,
           iter = n_iter,
           warmup = n_warm,
           backend = "cmdstanr",
           control = list(adapt_delta = 0.95))

# All 4 chains finished successfully.
# Mean chain execution time: 56.7 seconds.
# Total execution time: 66.2 seconds.
# 
# Warning: 3 of 16000 (0.0%) transitions ended with a divergence.
# See https://mc-stan.org/misc/warnings for details.

saveRDS(mod, file = here(paste0("outputs/ScotsPine/StanModels/",trait,"_TwoCG_SpeciesLevelModel_PopulationVaryingSlopes.rds")))
```

Summary of the model:

```{r LoadSpeciesLevelModelRandomSlopes}
mod <- readRDS(file = here(paste0("outputs/ScotsPine/StanModels/",trait,"_TwoCG_SpeciesLevelModel_PopulationVaryingSlopes.rds"))) 

summary(mod)
```

Interval plots of the population varying slopes (i.e., random slopes).

```{r IntervalPlotsPopulationSpecificSlopesSpeciesLevelModelRandomSlopes, fig.height=10, fig.width=10}
# Extract posterior samples for population-level slopes
population_slopes <- as_draws_df(mod) %>%
  spread_draws(r_PopulationCode[PopulationCode, FieldSite_num]) %>% 
  filter(FieldSite_num=="FieldSite_num")

population_slopes %>%
  ggplot(aes(x = r_PopulationCode, y = PopulationCode)) +
  stat_halfeye(
    .width = c(0.80, 0.95),   # Specify intervals
    fill = "skyblue",
    point_interval = "median_qi",
    color = "blue") +
  theme_bw() +
  labs(title = "MCMC areas of population varying slopes (median, 80% and 95% credible intervals)",
       x = "Slope",
       y = "Population")
```

Global slope of the model and its 95% credible interval.

```{r PlotGlobalSlopePopulationLevel, fig.height=5, fig.width=5}
mod_coeff <- mod %>% 
  broom.mixed::tidyMCMC(
    robust = T, # median and mean absolute deviation are used to compute point estimates and uncertainty
    ess = F, 
    rhat = F, 
    conf.int = T,
    conf.level = 0.95)

global_intercept <- mod_coeff$estimate[["b_Intercept"]]
global_slope <- mod_coeff$estimate[["b_FieldSite_num"]]
global_slope_low95 <- mod_coeff[mod_coeff$term=="b_FieldSite_num","conf.low"] %>% pull()
global_slope_high95 <- mod_coeff[mod_coeff$term=="b_FieldSite_num","conf.high"] %>% pull()

ggplot(data = subdata) +
  geom_point(aes(x = FieldSite_num, y = y), color = "gray") +
  scale_x_continuous(breaks = c(1, 2), 
                     limits = c(0.8, 2.2),
                     labels = c("Glensaugh (FE)", "Yair (FS)")) +  # Set x-axis limits and breaks
    geom_ribbon(aes(x = FieldSite_num, 
                    ymin = global_intercept + global_slope_low95 * FieldSite_num, 
                    ymax = global_intercept + global_slope_high95 * FieldSite_num),  
                inherit.aes = FALSE, 
                fill = "blue", 
                alpha = 0.2) + 
  geom_abline(slope = global_slope, 
              intercept = global_intercept, 
              color = "blue", 
              linetype = "dashed", 
              linewidth = 1) +
  theme_bw() +
  labs(title = "Global slope (median and 95% intervals)",
       x = "",
       y = "Tree height (mm)")
```

Now we look the estimated intercept and slope of each population. The credible intervals reflect uncertainty in the estimation of the slope only.

```{r PlotPopRandomSlopesPopLEvel,fig.height=5, fig.width=5}
pop_names <- subdata$PopulationCode %>% unique %>% sort

lapply(pop_names, function(pop_i){
  
  term_intercept <- paste0("r_PopulationCode[",pop_i,",Intercept]")
  term_slope <- paste0("r_PopulationCode[",pop_i,",FieldSite_num]")
  pop_intercept <- mod_coeff[mod_coeff$term==term_intercept,"estimate"] %>% pull
  pop_slope <- mod_coeff[mod_coeff$term==term_slope,"estimate"] %>% pull
  pop_slope_low95 <- mod_coeff[mod_coeff$term==term_slope,"conf.low"] %>% pull
  pop_slope_high95 <- mod_coeff[mod_coeff$term==term_slope,"conf.high"] %>% pull
  
ggplot(data = subdata) +
  geom_point(aes(x = FieldSite_num, y = y), color="gray") +
  scale_x_continuous(breaks = c(1, 2), 
                     limits = c(0.5, 2.5),
                     labels = c("Glensaugh (FE)", "Yair (FS)")) +  # Set x-axis limits and breaks
  geom_ribbon(aes(x = FieldSite_num, 
                  ymin = global_intercept + global_slope_low95 * FieldSite_num, 
                  ymax = global_intercept + global_slope_high95 * FieldSite_num),  
              inherit.aes = FALSE, 
              fill = "blue", 
              alpha = 0.2) + 
  geom_ribbon(aes(x = FieldSite_num, 
                  ymin = (global_intercept + pop_intercept) + (global_slope + pop_slope_low95) * FieldSite_num, 
                  ymax = (global_intercept + pop_intercept) + (global_slope + pop_slope_high95) * FieldSite_num),  
              inherit.aes = FALSE, 
              fill = "red", 
              alpha = 0.2) + 
  geom_point(data=subdata %>% filter(subdata$PopulationCode==pop_i), aes(x=FieldSite_num, y=y), color="red", alpha=0.8) +
  geom_abline(slope = global_slope, 
              intercept = global_intercept, 
              color = "blue", 
              linetype = "dashed", 
              linewidth = 1) +
  geom_abline(slope=(global_slope + pop_slope) , intercept=(global_intercept + pop_intercept), color="red", alpha=0.8, linewidth=0.5) +
  scale_color_manual(values=c("red",rep("gray",20))) +
  theme_bw() +
  labs(title = paste0("Population ",pop_i),
       x = "",
       y = "Tree height (mm)")
})
```

We extract the slopes for comparison.

```{r extractMedianSlopes}
  median_slope_df <- mod_coeff %>% 
  filter(stringr::str_detect(term, ",FieldSite_num]$")) %>% 
  mutate(PopulationCode = str_extract(term, "\\[([A-Z]{2}),")) %>%
  mutate(PopulationCode = str_remove_all(PopulationCode, "\\[|,")) %>% 
  dplyr::select(PopulationCode, estimate) %>% 
  dplyr::rename(median_slope = estimate)
```


<span style="color: orange;">Main take-home message:</span> Very low variation in the slope across sites among populations, and the differences in slope are not statistically meaningful.

We extract the median of the posterior distribution of the slope for each population and name it `median_slope`.

## Family varying slopes{#sec-SPmodelFamVaryingSlopes}

```{r RunSpeciesLevelModelFamilyRandomSlopes, eval=F}
# Model options
n_chains <- 4 # number of chains (MCMC)
n_iter <- 8000 # number of iterations
n_warm <- 4000 # number of iterations in the warm-up phase
n_thin <- 1 # thinning interval
n_seed <- 404

form <- brmsformula(y ~ 1 + FieldSite_num + (1 + FieldSite_num | Family))

mod <- brm(formula = form,
           data = subdata,
           chains = n_chains,
           cores = n_chains,
           seed = n_seed,
           iter = n_iter,
           warmup = n_warm,
           backend = "cmdstanr",
           control = list(adapt_delta = 0.95))

# All 4 chains finished successfully.
# Mean chain execution time: 56.5 seconds.
# Total execution time: 64.7 seconds.

saveRDS(mod, file = here(paste0("outputs/ScotsPine/StanModels/",trait,"_TwoCG_SpeciesLevelModel_FamilyVaryingSlopes.rds")))
```

Summary of the model:

```{r LoadSpeciesLevelModelFamilyRandomSlopes}
mod <- readRDS(file = here(paste0("outputs/ScotsPine/StanModels/",trait,"_TwoCG_SpeciesLevelModel_FamilyVaryingSlopes.rds"))) 

summary(mod)
```  

Global slope of the model and its 95% credible interval.

```{r PlotGlobalSlopefamilyLevel, fig.height=5, fig.width=5}
mod_coeff <- mod %>% 
  broom.mixed::tidyMCMC(
    robust = T, # median and mean absolute deviation are used to compute point estimates and uncertainty
    ess = F, 
    rhat = F, 
    conf.int = T,
    conf.level = 0.95)

global_intercept <- mod_coeff$estimate[["b_Intercept"]]
global_slope <- mod_coeff$estimate[["b_FieldSite_num"]]
global_slope_low95 <- mod_coeff[mod_coeff$term=="b_FieldSite_num","conf.low"] %>% pull()
global_slope_high95 <- mod_coeff[mod_coeff$term=="b_FieldSite_num","conf.high"] %>% pull()

ggplot(data = subdata) +
  geom_point(aes(x = FieldSite_num, y = y), color = "gray") +
  scale_x_continuous(breaks = c(1, 2), 
                     limits = c(0.8, 2.2),
                     labels = c("Glensaugh (FE)", "Yair (FS)")) +  # Set x-axis limits and breaks
    geom_ribbon(aes(x = FieldSite_num, 
                    ymin = global_intercept + global_slope_low95 * FieldSite_num, 
                    ymax = global_intercept + global_slope_high95 * FieldSite_num),  
                inherit.aes = FALSE, 
                fill = "blue", 
                alpha = 0.2) + 
  geom_abline(slope = global_slope, 
              intercept = global_intercept, 
              color = "blue", 
              linetype = "dashed", 
              linewidth = 1) +
  theme_bw() +
  labs(title = "Global slope (median and 95% intervals)",
       x = "",
       y = "Tree height (mm)")
```

We plot the plastic response of each family (i.e. family-level slopes across the two sites) within each population (one graph per population).

```{r PlotFamilyRandomSlopes,fig.height=5, fig.width=5}
pop_names <- subdata$PopulationCode %>% unique %>% sort


plot_pop_families_only <- function(pop_names,
                                   subdata,
                                   mod_coeff,
                                   global_intercept,
                                   global_slope,
                                   x_var = "FieldSite_num",
                                   y_var = "y",
                                   family_col = "Family",
                                   pop_col = "PopulationCode",
                                   x_breaks = c(1, 2),
                                   x_limits = c(0.5, 2.5),
                                   x_labels = c("Glensaugh (FE)", "Yair (FS)"),
                                   x_seq_len = 60) {
  # helper to generate candidate term names for a family
  gen_terms <- function(fam, x_var) {
    list(
      intercept_terms = c(
        paste0("r_Family[", fam, ",Intercept]"),
        paste0("r_Family[", fam, ",(Intercept)]"),
        paste0("Family[", fam, ",Intercept]"),
        paste0("Family[", fam, ",(Intercept)]")
      ),
      slope_terms = c(
        paste0("r_Family[", fam, ",", x_var, "]"),
        paste0("r_Family[", fam, ",(FieldSite_num)]"),
        paste0("Family[", fam, ",", x_var, "]"),
        paste0("Family[", fam, ",FieldSite_num]")
      )
    )
  }
  extract_first <- function(terms, what = "estimate") {
    for (t in terms) {
      idx <- which(mod_coeff$term == t)
      if (length(idx) >= 1) return(mod_coeff[[what]][idx[1]])
    }
    return(NA_real_)
  }

  plots <- purrr::map(pop_names, function(pop_i) {
    families <- subdata %>% filter(.data[[pop_col]] == pop_i) %>% pull(.data[[family_col]]) %>% unique()
    if (length(families) == 0) {
      warning("No families for population: ", pop_i)
      return(NULL)
    }

    fam_df <- tibble::tibble(Family = families) %>%
      dplyr::rowwise() %>%
      dplyr::mutate(
        fam_intercept = extract_first(gen_terms(Family, x_var)$intercept_terms, "estimate"),
        fam_slope     = extract_first(gen_terms(Family, x_var)$slope_terms, "estimate"),
        fam_slope_low = extract_first(gen_terms(Family, x_var)$slope_terms, "conf.low"),
        fam_slope_high= extract_first(gen_terms(Family, x_var)$slope_terms, "conf.high")
      ) %>% 
      dplyr::ungroup()

    x_seq <- seq(x_limits[1], x_limits[2], length.out = x_seq_len)

    fam_pred <- fam_df %>%
      dplyr::filter(!is.na(fam_intercept) & !is.na(fam_slope)) %>%
      dplyr::group_by(Family) %>%
      dplyr::do({
        fi <- .$fam_intercept
        fs <- .$fam_slope
        fsl <- .$fam_slope_low
        fsh <- .$fam_slope_high
        data.frame(
          x = x_seq,
          y = (global_intercept + fi) + (global_slope + fs) * x_seq,
          ymin = if(!is.na(fsl)) (global_intercept + fi) + (global_slope + fsl) * x_seq else NA_real_,
          ymax = if(!is.na(fsh)) (global_intercept + fi) + (global_slope + fsh) * x_seq else NA_real_
        )
      }) %>%
      dplyr::ungroup() %>%
      dplyr::mutate(Family = rep(fam_df$Family[!is.na(fam_df$fam_intercept) & !is.na(fam_df$fam_slope)], each = x_seq_len))

    ggplot(data = subdata) +
      geom_point(aes(x = .data[[x_var]], y = .data[[y_var]]), color = "gray80", alpha = 0.5) +
      geom_ribbon(data = fam_pred %>% filter(!is.na(ymin) & !is.na(ymax)),
                  aes(x = x, ymin = ymin, ymax = ymax, group = Family),
                  inherit.aes = FALSE, alpha = 0.05, fill = "red") +
      geom_line(data = fam_pred, aes(x = x, y = y, group = Family),
                color = "red", linewidth = 0.5, alpha = 0.8) +
      geom_point(data = subdata %>% filter(.data[[pop_col]] == pop_i),
                 aes(x = .data[[x_var]], y = .data[[y_var]]),
                 color = "red", alpha = 0.9) +
      geom_abline(slope = global_slope, intercept = global_intercept,
                  color = "blue", linetype = "dashed", linewidth = 0.9) +
      scale_x_continuous(breaks = x_breaks, limits = x_limits, labels = x_labels) +
      theme_bw() +
      labs(title = paste0("Population ", pop_i),
           x = "", y = "Tree height (mm)")
  }) %>% purrr::set_names(pop_names)

  return(plots)
}

plots_list <- plot_pop_families_only(
  pop_names = pop_names,
  subdata = subdata,
  mod_coeff = mod_coeff,
  global_intercept = global_intercept,
  global_slope = global_slope
)

# show a single population plot:
print(plots_list)
```


We compute the posterior variance of family slopes per population and we compare those variances among populations in the table below.

```{r PosteriorDrawsFamilySlopes, warning=F}
# Extract all posterior draws
draws <- as_draws_df(mod)

# Extract population-level slope (fixed effect)
beta_slope <- draws$b_FieldSite_num

# Extract family-level deviations (random effects for slope)
family_slopes <- ranef(mod, pars = "FieldSite_num")$Family[, , "FieldSite_num"]

# ranef() gives summary stats, not full posteriors. To get the posterior samples of each family slope, use:
family_slopes_post <- as_draws_df(mod) %>% dplyr::select(starts_with("r_Family[") & ends_with(",FieldSite_num]"))
# Each column is the deviation from the global slope for one family across posterior samples

# We add global slope and compute family-specific posterior slopes.
family_slopes_post <- family_slopes_post %>%  mutate(across(everything(), ~ . + beta_slope))  # add fixed effect slope

# Compute posterior variance of family slopes per population
fam_to_pop <- subdata %>%  distinct(Family, PopulationCode)

# reshape long
slopes_long <- family_slopes_post %>%
  mutate(draw = row_number()) %>%
  pivot_longer(-draw, names_to = "term", values_to = "slope") %>%
  mutate(Family = sub("r_Family\\[(.*),FieldSite_num\\]", "\\1", term)) %>%
  left_join(fam_to_pop, by = "Family")

# Variance of slopes per population, per posterior draw
var_post <- slopes_long %>%
  group_by(draw, PopulationCode) %>%
  summarise(var_slope = var(slope), .groups = "drop")

# Summarise with credible intervals
summary_var <- var_post %>%
  group_by(PopulationCode) %>%
  summarise(
    median = median(var_slope),
    low95 = quantile(var_slope, 0.025),
    high95 = quantile(var_slope, 0.975),
    .groups = "drop"
  )

sum_pop <- summary_var %>% 
  dplyr::select(PopulationCode,median) %>% 
  dplyr::rename(var_fam_slope=median) %>% 
  left_join(median_slope_df, ., by = "PopulationCode")

# all pairwise comparisons of populations
pop_pairs <- combn(unique(var_post$PopulationCode), 2, simplify = FALSE)

# for each pair, compute posterior differences
pairwise_diffs <- purrr::map_dfr(pop_pairs, function(p) {
  df1 <- var_post %>% filter(PopulationCode == p[1])
  df2 <- var_post %>% filter(PopulationCode == p[2])
  
  stopifnot(nrow(df1) == nrow(df2))  # should match draws
  tibble(
    draw = df1$draw,
    pop1 = p[1],
    pop2 = p[2],
    diff = df1$var_slope - df2$var_slope
  )
})

# Summarise differences
pairwise_summary <- pairwise_diffs %>%
  group_by(pop1, pop2) %>%
  summarise(
    median = median(diff),
    low95 = quantile(diff, 0.025),
    high95 = quantile(diff, 0.975),
    p_greater0 = mean(diff > 0),   # posterior probability var(pop1) > var(pop2)
    p_less0    = mean(diff < 0),   # posterior probability var(pop1) < var(pop2)
    .groups = "drop"
  )

pairwise_summary %>% kable_mydf()
```

This tables gives us:

  - `median`: posterior median difference (difference in the variance of the family-level slopes among two populations)

  - `low95`, `high95`: 95% credible interval of the difference

  - `p_greater0`: probability variance(pop1) > variance(pop2)

  - `p_less0`: probability variance(pop1) < variance(pop2)

Two populations can be considered to have a statistically meaningful difference in the variance of their family slopes if the 95% credible interval does not include zero.

Alternatively, posterior probabilities can be used for interpretation:

  - `p_greater0 > 0.95` → strong evidence that variance(pop1) > variance(pop2)

  - `p_greater0 ≈ 0.5` → no clear difference between populations.
  
In our analysis, we did not observe any statistically meaningful differences between populations.


We extract the median of the posterior distribution of the variance of family slopes for each population and name it `var_fam_slope`.


# Mean trait values{#sec-MeanTraitValues}

We extract the mean trait values of each population in each common garden.

```{r ExtractMeanTraitValue}
sum_pop <- subdata %>%
  group_by(PopulationCode,FieldSite) %>%
  summarise(
    mean_y = mean(y, na.rm = TRUE),
    .groups = "drop") %>% 
   pivot_wider(
    names_from = FieldSite,
    values_from = mean_y,
    names_prefix = "mean_y_"
  ) %>% 
  left_join(sum_pop, by="PopulationCode")
```

# RDPI{#sec-RDPI}


In this section, we calculate the relative distance plasticity index (RDPI) from @valladaresQuantitativeEstimationPhenotypic2006. In this paper, the authors propose to quantify phenotypic plasticity based on phenotypic distances among individuals of a given species (in our study, population) exposed to different environments (in our study, planted in different common gardens), which is summarized in the RDPI index that allows for statistical comparisons of phenotypic plasticity between species (in our study, populations). 

For a given trait, the relative distance plasticity index $RDPI_p$ for a given population $p$ can be calculated as follows:

$$RDPI_p = \frac{\sum_{n=1}^N{\frac{x_{ij}-x_{i'j'}}{x_{ij}+x_{i'j'}}}}{N}$$

where $N$ is the number of pairwise comparisons among individuals of the same population but from different common gardens. $x_{ij}$ and $x_{i'j'}$ are the phenotypic values of individual $j$ and $j'$ in common gardens $i$ and $i’$. As such, the phenotypic plasticity of a population for a given trait corresponds to the sum of the relative phenotypic distances for all pairs of individuals of the population grown in different common gardens. 

We also calculate the RDPI at the population level. In this case, we first compute the mean trait value of each population in each common garden, and then calculate RDPI based on these population means. The population-mean RDPI therefore summarizes overall population-level plasticity, integrating the average phenotypic shift of the population across environments while ignoring within-population variation among individuals.

Comment: Before calculating the RDPI, I shifted the trait values to ensure they are all positive because the index involves dividing by the sum of trait values across environments. When traits include negative or zero values, the denominator can become very small or change sign, producing extreme or negative RDPI values. Shifting the values preserves the relative differences among individuals while avoiding these numerical artifacts.

```{r CalcRDPI}
# Find the minimum value of the trait
trait_min <- min(subdata$y, na.rm = TRUE)

# If the minimum is <= 0, shift the trait so all values are positive
if(trait_min <= 0) {
  shift_value <- abs(trait_min) + 1  # add 1 to ensure strictly positive
  subdata <- subdata %>% mutate(y_shifted = y + shift_value)
} else {
  subdata <- subdata %>% mutate(y_shifted = y)
}


# my objects
pop_codes <- unique(subdata$PopulationCode) %>% sort() # population codes
site_codes <- unique(subdata$FieldSite) %>% sort() # field site codes
site_comb <- combn(site_codes,2)
calc_rel_dist <- function(x){ abs(x[1] - x[2]) / (x[1] + x[2]) }

# --- main loop: compute both individual-based RDPI and population-mean RDPI ---
res_list <- lapply(pop_codes, function(pop){

  subpop <- subdata %>% dplyr::filter(PopulationCode %in% pop)

  # ---------- individual-based RDPI (your original calculation) ----------
  rel_dist_ind <- lapply(1:dim(site_comb)[[2]], function(nb_comb){

    subsite1 <- subpop %>% dplyr::filter(FieldSite %in% site_comb[1, nb_comb])
    subsite2 <- subpop %>% dplyr::filter(FieldSite %in% site_comb[2, nb_comb])

    lapply(1:nrow(subsite1), function(x) {
      rbind(subsite1[x, "y_shifted"], subsite2[, "y_shifted"]) %>%
        pull() %>%
        combn(2, calc_rel_dist) %>%
        .[1:nrow(subsite2)] %>%
        as.numeric()
    }) %>% unlist()

  }) %>% unlist()

  N_ind <- length(rel_dist_ind)
  ind_RDPI <- if (N_ind == 0) NA_real_ else sum(rel_dist_ind) / N_ind

  # ---------- population-mean RDPI (new, minimal addition) ----------
  # compute per-site population means for this population
  mean_tab <- subpop %>%
    drop_na(y) %>%
    group_by(FieldSite) %>%
    summarise(pop_mean = mean(y_shifted, na.rm = TRUE), n = n(), .groups = "drop")

  # for each site combination compute RDPI on the two site means (if both present)
  rel_dist_pop_means <- lapply(1:dim(site_comb)[[2]], function(nb_comb){
    s1 <- site_comb[1, nb_comb]
    s2 <- site_comb[2, nb_comb]

    mean1 <- mean_tab %>% filter(FieldSite == s1) %>% pull(pop_mean)
    mean2 <- mean_tab %>% filter(FieldSite == s2) %>% pull(pop_mean)

    # if either mean is missing, return NA for this combo
    if(length(mean1) == 0 || length(mean2) == 0) return(NA_real_)

    denom <- mean1 + mean2
    if(denom == 0) return(NA_real_) else return(abs(mean1 - mean2) / denom)
  }) %>% unlist()

  # average across combos, ignoring NA combos (if all NA -> result NA)
  if(all(is.na(rel_dist_pop_means))) {
    pop_mean_RDPI <- NA_real_
  } else {
    pop_mean_RDPI <- mean(rel_dist_pop_means, na.rm = TRUE)
  }

  # return a named vector/list for this population
  c(PopulationCode = pop,
    ind_RDPI = ind_RDPI,
    pop_mean_RDPI = pop_mean_RDPI)
})

# bind into a tibble and ensure numeric columns are numeric
df_rdpi <- do.call(rbind, res_list) %>%
  as_tibble() %>%
  mutate(
    ind_RDPI = as.numeric(ind_RDPI),
    pop_mean_RDPI = as.numeric(pop_mean_RDPI)
  )

sum_pop <- df_rdpi %>% left_join(sum_pop, by="PopulationCode")

# final object: df_rdpi with columns PopulationCode, ind_RDPI, pop_mean_RDPI
df_rdpi %>% kable_mydf()
```


# Relationships among all the estimated indexes{#sec-IndexRelationships}

## PCA

```{r PCAonCalaculateIndexes, fig.height=8, fig.width=8}
# keep only populations with complete data for PCA (or use na.omit)
pca_df <- sum_pop %>% drop_na()

# run PCA (center & scale)
pca_res <- prcomp(pca_df %>% select(-PopulationCode), center = TRUE, scale. = TRUE)

# scores (PC coordinates for populations)
scores <- as.data.frame(pca_res$x)
scores$PopulationCode <- pca_df$PopulationCode

# loadings (variable contributions)
loadings <- as.data.frame(pca_res$rotation)
loadings$var <- rownames(loadings)

# get % variance explained for axis labels
eig <- (pca_res$sdev)^2
pct_var <- round(100 * eig / sum(eig), 1)

# scale loadings to make arrows visible on the same plotting scale as scores
# choose a multiplier so arrows reach ~30-50% of the score cloud
mult <- min(
  (max(scores$PC1) - min(scores$PC1)) / (max(loadings$PC1) - min(loadings$PC1)),
  (max(scores$PC2) - min(scores$PC2)) / (max(loadings$PC2) - min(loadings$PC2))
) * 0.6

loadings_plot <- loadings %>%
  mutate(arrow_x = PC1 * mult,
         arrow_y = PC2 * mult)

# create a color palette with one color per population
n_pops <- nrow(scores)
pal <- scales::hue_pal()(n_pops) # distinct hues; change palette if you prefer

# Plot
ggplot() +
  # points for populations
  geom_point(data = scores, aes(x = PC1, y = PC2, color = PopulationCode), size = 3, alpha = 0.9) +
  # population labels
  geom_text_repel(data = scores, aes(x = PC1, y = PC2, label = PopulationCode, color = PopulationCode),
                  size = 3.5, show.legend = FALSE, max.overlaps = 100) +
  # arrows for variable loadings (from origin to scaled loading)
  geom_segment(data = loadings_plot,
               aes(x = 0, y = 0, xend = arrow_x, yend = arrow_y),
               arrow = arrow(length = unit(0.25, "cm")), colour = "black", linewidth = 0.6) +
  # variable labels at arrow tips
  geom_text_repel(data = loadings_plot,
                  aes(x = arrow_x, y = arrow_y, label = var),
                  color = "black", size = 4, segment.size = 0.3) +
  scale_color_manual(values = pal) +
  labs(x = paste0("PC1 (", pct_var[1], "%)"),
       y = paste0("PC2 (", pct_var[2], "%)"),
       title = "",
       color = "Population") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none") # legend suppressed because each pop has its own color label

```

## Correlations 

```{r PCAonCalculatedPhenoIndexes, fig.height=8, fig.width=8}
 sum_pop %>% 
  dplyr::select(-PopulationCode) %>% 
  cor() %>% 
  corrplot(method = 'number',
           type = 'lower', diag = FALSE,
           title=paste0(""),
           mar=c(0,0,2,0),
           number.cex=0.9,tl.cex=0.9)
```


The two RDPI indexes (@sec-RDPI) and the population-level slopes (@sec-SPmodelPopVaryingSlopes) are highly correlated, which is reassuring. No correlation with the variance of the family-level slopes (@sec-SPmodelFamVaryingSlopes), which is not surprising as this index does not capture the plastic reponses across sites, but whther their is variation in the plastic responses of the families in a given population.
Interestingly, the populations that are the most plastic (even though we have to keep in mind that the diffferneces in the plastic responses are very small) are those that have an early phenology in FE. No correlation between the plastic responses of the populations and the phenology in FS.

# Correlations with genomic offset predictions{#sec-CorrWithGenomicOffset}

```{r DownloadGOpred}
df_go_clim_dist <- readRDS(here("outputs/ScotsPine/df_gen_off_pred_clim_dist.rds"))

df_go_wide <- df_go_clim_dist %>%
  select(PopulationCode, rcp_scenario, pred_timeslice, method, ensemble_member, var_code,val) %>%
  mutate(var_name_comb = paste(rcp_scenario, pred_timeslice, method, ensemble_member, var_code,sep = "_") %>%  gsub("[ .-]", "", .)) %>%
  dplyr::select(PopulationCode, var_name_comb,val) %>% 
  pivot_wider(names_from = var_name_comb, values_from = val)


df_go_wide <- df_go_clim_dist %>%
  dplyr::filter(var_code == "all_snps" & rcp_scenario == "RCP2.6" & pred_timeslice == "2030-2050" & ensemble_member == "Average across EMs") %>% 
  select(PopulationCode, method, val) %>%
  mutate(var_name_comb = paste(method, sep = "_") %>%  gsub("[ .-]", "", .)) %>%
  dplyr::select(PopulationCode, var_name_comb, val) %>% 
  pivot_wider(names_from = var_name_comb, values_from = val)
```


```{r PCAwithGO, fig.height=8, fig.width=8}
# keep only populations with complete data for PCA (or use na.omit)
pca_df <- sum_pop %>% left_join(df_go_wide, by="PopulationCode")

# run PCA (center & scale)
pca_res <- prcomp(pca_df %>% select(-PopulationCode), center = TRUE, scale. = TRUE)

# scores (PC coordinates for populations)
scores <- as.data.frame(pca_res$x)
scores$PopulationCode <- pca_df$PopulationCode

# loadings (variable contributions)
loadings <- as.data.frame(pca_res$rotation)
loadings$var <- rownames(loadings)

# get % variance explained for axis labels
eig <- (pca_res$sdev)^2
pct_var <- round(100 * eig / sum(eig), 1)

# scale loadings to make arrows visible on the same plotting scale as scores
# choose a multiplier so arrows reach ~30-50% of the score cloud
mult <- min(
  (max(scores$PC1) - min(scores$PC1)) / (max(loadings$PC1) - min(loadings$PC1)),
  (max(scores$PC2) - min(scores$PC2)) / (max(loadings$PC2) - min(loadings$PC2))
) * 0.6

loadings_plot <- loadings %>%
  mutate(arrow_x = PC1 * mult,
         arrow_y = PC2 * mult)

# create a color palette with one color per population
n_pops <- nrow(scores)
pal <- scales::hue_pal()(n_pops) # distinct hues; change palette if you prefer

# Plot
ggplot() +
  # points for populations
  geom_point(data = scores, aes(x = PC1, y = PC2, color = PopulationCode), size = 3, alpha = 0.9) +
  # population labels
  geom_text_repel(data = scores, aes(x = PC1, y = PC2, label = PopulationCode, color = PopulationCode),
                  size = 3.5, show.legend = FALSE, max.overlaps = Inf) +
  # arrows for variable loadings (from origin to scaled loading)
  geom_segment(data = loadings_plot,
               aes(x = 0, y = 0, xend = arrow_x, yend = arrow_y),
               arrow = arrow(length = unit(0.25, "cm")), colour = "black", linewidth = 0.6) +
  # variable labels at arrow tips
  geom_text_repel(data = loadings_plot,
                  aes(x = arrow_x, y = arrow_y, label = var),
                  color = "black", size = 4, segment.size = 0.3, max.overlaps = Inf) +
  scale_color_manual(values = pal) +
  labs(x = paste0("PC1 (", pct_var[1], "%)"),
       y = paste0("PC2 (", pct_var[2], "%)"),
       title = "",
       color = "Population") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none") # legend suppressed because each pop has its own color label

```
 
```{r CorrelationsWithGO, fig.height=8, fig.width=8}
sum_pop %>% left_join(df_go_wide, by="PopulationCode") %>% 
  dplyr::select(-PopulationCode) %>% 
  cor() %>% 
  corrplot(method = 'number',
           type = 'lower', diag = FALSE,
           title=paste0(""),
           mar=c(0,0,2,0),
           number.cex=0.9,tl.cex=0.9)
```

**Conclusion**

There appears to be no correlation between genomic offset predictions under future climates and phenology plastic responses. In other words, populations that are more plastic or exhibit greater family-level variation in plasticity are not more or less at risk of maladaptation under climate change.

We were unable to identify statistically meaningful differences among populations in their plastic responses. Although such differences may exist, they may not be detectable due to limited sample sizes rather than a true absence of variation. This is why we still examined the relationships with genomic offset predictions—because even small differences in plasticity can sometimes have a substantial impact on population adaptive responses. However, no patterns were observed between plasticity and genomic offset. In this case study, there is no evidence that populations predicted to be more at risk of climate maladaptation compensate through higher phenotypic plasticity.

