---
title: "Plastic responses in disease susceptibility at the population level"
author: "Juliette Archambeau"
date: last-modified
format: 
  html:
    toc: true
    toc-depth: 4
    code-fold: true
    page-layout: full
    number-sections: true
embed-resources: true
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

<style type="text/css">
body {
   font-size: 15px;
}
code.r{
  font-size: 11px;
}
pre {
  font-size: 11px
}

table {
  font-size: 11px
}
</style>

```{r setup, include=F}
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))
knitr::opts_chunk$set(cache=F)
options(width = 300)
library(knitr)
library(kableExtra)
library(tidyverse)
library(here)
library(magrittr)
library(corrplot)
library(readxl)
library(rstan)
library(tidybayes)
library(bayesplot)
library(ggplot2)
library(parallel)
library(tictoc)
library(latex2exp)
library(brms)
library(Reacnorm)
library(matrixStats)
library(posterior)
library(missMDA)
library(ggrepel)
library(cmdstanr)
library(cowplot)
library(ggExtra)

rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

theme_set(theme_bw(base_size = 20))

# my own function for building tables in reports
source(here("scripts/functions/kable_mydf.R"))
```

# Introduction

The goal of this report is to determine whether the populations have different plastic responses for disease susceptibility. We have four disease-related variables: `DNB16`and `DNB22`, which are the percentage of needles with lesions measured in 2016 and 2022, and`NR16`and `NR22`, which are the number of years needle retention, also measured in 2016 and 2022. Because these variables capture distinct symptoms and are expressed on different scales, in @sec-SynthDiseaseIndex, we constructed year-specific synthetic susceptibility indices. For each year, we inverted NR values so that higher values consistently reflect higher susceptibility, standardized DNB and NR to remove scale differences, averaged them to give equal weight to both symptoms, and finally applied a min–max rescaling to obtain strictly positive indices on a common [0,1] scale. The resulting variables, `Susc16` and `Susc22`, provide comparable measures of disease susceptibility across years, where high values indicate high susceptibility. To avoid confounding with nursery origin, we restricted analyses to the two common gardens where trees originated from a single nursery (FS from NG and FE from NE), removing the four FE trees that originated from NG.

In @sec-ReacNormVarPart, we applied the approach developed by Villemereuil & Chevin (2025) to assess whether the heritability of the reaction norm ($h_{RN}^2$) primarily stems from the environment-blind heritability of the trait ($h^2$, based on breeding values averaged across environments) or from the heritability of plasticity ($h_I^2$, based on breeding values by environment interaction). We implemented this approach at the species level, rather than at the population level, because it requires a large amount of data. We implemented the "animal model" version of the Villemereuil and Chevin approach, i.e. the version that uses a genomic relationship matrix (GRM) to estimate additive genetic effects. These models take a very long time to run because they require many iterations for the Markov chains to converge and careful parameter tuning. Some warnings still occurred for `DiseaseIndex_Susc16`, but I do not believe these are concerning. I will try to re-run this model with further parameter tuning (it may take about two days to complete), in the hope that the warnings will disappear. 
<!-- Because these models were very difficult to fit using the GRM, I also implemented the Villemereuil and Chevin (2025) approach without the genomic relationship matrix in section @sec-SPmodelWithoutGRM. In this case, additive genetic effects cannot still be estimated, only broad-sense heritabilities are accessible. The results differ slightly between the two approaches, as expected, but the main patterns remain consistent, so I believe they can be interpreted with confidence. -->

In @sec-FamReactionNormEstim, we run a standard reaction-norm model by estimating family-level reaction-norms and we then extract two variables of interest: the median reaction-norm slope across families within a population (quantifies the average change in disease susceptibility between sites for that population), and the median variance of family-specific slopes within a population (wich measures the within-population variation in responses to the site contrast). These two parameters summarize, respectively, the population-level environmental response and the heterogeneity among families in that response.

In @sec-MeanTraitValues, we simply quickly extract the mean trait values (sign-inverted and scaled PC1 and scaled PC2) of each population in each field site to compare their values to those of the parameters of interest.

In @sec-RDPI, we quantified phenotypic plasticity using the relative distance plasticity index (RDPI; Valladares et al., 2006), which measures the mean relative phenotypic distance among individuals of the same population grown in different environments. We calculated RDPI in two ways: (i) at the individual level (i.e., the common practice), where all pairwise phenotypic distances between individuals of a given population and from different common gardens are considered, and (ii) using population means per site, which summarizes differences in average population responses across sites.

In @sec-IndexRelationships, we estimated the relationships among all these indice using a principal component analysis (PCA) and a correlation matrix.

In @sec-CorrWithGenomicOffset, we estimated the relationships between our phenology indexes with the predicted genomic offsets. We used genomic offset predictions from the RCP scenario 2.6, the prediction time period 2030-2050 and the average across the four climatic models (ensemble members).


# Indexes capturing disease susceptibility{#sec-SynthDiseaseIndex}

## The data

We load the disease data and show the first 10 rows of the dataset.

```{r LoadDiseaseData}
disease_data <- read.delim(here("data/ScotsPine/Disease_raw_Juliette.txt")) %>% 
  dplyr::rename(FieldCode=id,
                FieldSite=Site)
disease_data[1:10,] %>% kable_mydf()
```


We merge this data with the other phenotypic data from the common gardens filter for either the common garden FS combined with the NG nursery, or the common garden FE and the nursery NE.

```{r LoadPhenoCGData}
data <- readRDS(file=here("data/ScotsPine/formatted_CG_data.rds")) %>% # 1848 obs
  full_join(disease_data,by=c("FieldSite","FieldCode")) %>% 
  filter(FieldSite %in% c("FE","FS")) %>% # 1344 obs
  filter(!(FieldSite=="FE" & Nursery =="NG")) # 1340 obs
```

The dataset contains `r length(unique(data$FieldCode))` individuals, `r length(unique(data$PopulationCode))` populations, `r length(unique(data$Family))` families. Here are the first 10 rows:

```{r VizData, fig.width=8, fig.height=8}
data[1:10,] %>% kable_mydf()

# Correlation matrix plot for phenotypic variables
data %>%
  dplyr::select(HA20, contains("16"), contains("22"), contains("DNB"), contains("NR"), -DI16, -DA16, -HI16) %>%
  cor(use = "pairwise.complete.obs") %>%
  corrplot::corrplot(
    method = "color",
    type = "lower",
    order = "hclust",
    diag=F,
    addCoef.col = "black",
    tl.col = "black",
    tl.srt = 45,
    col = colorRampPalette(c("#E46726", "white","#6D9EC1"))(200),
    title = "Correlation among phenotypic variables",
    mar = c(0,0,2,0)
  )

data %>% 
  dplyr::select(contains("NR"),contains("DNB")) %>% 
  pivot_longer(everything(),names_to="variable",values_drop_na = TRUE) %>% 
  ggplot(aes(x=value)) +  
  geom_histogram(aes(y=after_stat(density)), colour="blue",fill="white",bins = 100) +
  #geom_density(alpha=.2,fill="pink") +
  xlab("") +
  facet_wrap(~variable,scales="free") + 
  theme_bw() 
```

A weak correlation ($\rho = 0.4$) was observed between `NR16` and `NR22`, indicating that needle retention was somewhat consistent across the two years, 2016 and 2022. This was not the case for `DNB16` and `DNB22` ($\rho = 0.05$). Interestingly, `NR22`, and even more so `NR16`, were slightly positively correlated with height. $\rho = -0.45$ between `NR16` and `DNB16`, so trees that had higher percentage of needles with lesions had also less needle retention, which is what we might expect. This was not the case in 2022: very low correlation between `NR22` and `DNB22` ($\rho = 0.15$; unexpected direction). This highlights the importance of accounting for both variables, as they capture different symptoms of the disease. A slightly negative correlation was found between `NR16` and both budburst duration (`BD16`) and stage 6 of budburst (`BT16_6`), suggesting that trees that completed budburst earlier were slightly more likely to retain their needles.

## Removing families in only one field site

We remove families that are found in only one field site.

```{r RemovingFamiliesInOneSiteOnly}
# Count how many FieldSites each family occurs in
family_sites <- data %>%
  group_by(PopulationCode, Family) %>%
  summarise(n_sites = n_distinct(FieldSite), .groups = "drop")

# Families that occur in only one FieldSite
single_site_families <- family_sites %>%
  filter(n_sites == 1)

single_site_families %>% kable_mydf

# Remove families present in only one FieldSite
data <- data %>%
  group_by(PopulationCode, Family) %>%
  filter(n_distinct(FieldSite) > 1) %>%
  ungroup()
```


## Removing missing values

We first identify the number of missing values for each trait:

```{r CountNAs}
subdata <- data %>%
  dplyr::select(FieldSite,Block,FieldCode,PopulationCode,Family,Nursery,matches("NR|DNB"))
  
print(sapply(subdata, function(x) sum(is.na(x))))
```

We then check whether the missing values correspond to the same individuals:

```{r PrintRowsWithNAs}
subdata_with_na <- subdata %>%
  filter(if_any(everything(), is.na))

subdata_with_na %>% kable_mydf()
```

Most missing values occur in the same individuals. We therefore remove these individuals before further analyses.


```{r removingNAs}
subdata <- subdata %>% drop_na()
```


## Year-specific synthetic indexes


To get year-specific indexes of disease susceptibility, we:

  - invert the NR variables so that higher values consistently reflect higher susceptibility
  - standardize each variable to remove scale differences between DNB and NR
  - average the standardized scores within each year to give equal weight to DNB and NR
  - rescale the year-specific scores with a min–max transformation to obtain strictly positive values on a common [0,1] scale
  
The final variables `Susc16` and `Susc22` are standardized indices, where high values indicate high susceptibility, fully comparable across years and suitable for downstream plasticity metrics.


```{r CalcSuscIndexes}
subdata <- subdata %>%
  # invert NR so that higher = higher susceptibility
  mutate(
    NR16_inv = -NR16,
    NR22_inv = -NR22
  ) %>%
  # standardize each variable
  mutate(
    DNB16_z = scale(DNB16)[,1],
    NR16_z  = scale(NR16_inv)[,1],
    DNB22_z = scale(DNB22)[,1],
    NR22_z  = scale(NR22_inv)[,1]
  ) %>%
  # build raw susceptibility indices
  mutate(
    Susc16_raw = (DNB16_z + NR16_z) / 2,
    Susc22_raw = (DNB22_z + NR22_z) / 2
  ) %>%
  mutate(
    Susc16 = (Susc16_raw - min(Susc16_raw)) /
             (max(Susc16_raw) - min(Susc16_raw)),
    Susc22 = (Susc22_raw - min(Susc22_raw)) /
             (max(Susc22_raw) - min(Susc22_raw))
  ) %>%
  select(-ends_with("inv"), -ends_with("raw"))

# to check
# mean(subdata$Susc16);sd(subdata$Susc16)
# mean(subdata$Susc22);sd(subdata$Susc22)
```

Correlation between `Susc16` and `Susc22`: `r cor(subdata$Susc16, subdata$Susc22)`.


```{r PlotSuscIndexes, fig.height=8, fig.width=8}
p <- ggplot(subdata, aes(x = Susc16, y = Susc22, colour = PopulationCode)) +
  geom_point(alpha = 0.7) +
  labs(
    x = "Susceptibility index 2016 (Susc16)",
    y = "Susceptibility index 2022 (Susc22)",
    colour = "Population"
  ) +
  theme_bw() +
  theme(
    legend.position = "right"
  )

ggMarginal(
  p,
  type = "histogram",
  margins = "both",
  bins = 30,
  groupColour = TRUE,
  groupFill = TRUE
)
```

Higher values of `Susc16` and `Susc22` correspond to greater disease susceptibility. 


# Variance partitioning with GRM {#sec-ReacNormVarPart}

To redo.

# Family-level reaction norms{#sec-FamReactionNormEstim}

## Running the model

```{r RunTheSpeciesLevelModelWithoutGRMNested, eval=F}
n_chains <- 4  # number of chains (MCMC)
n_iter <- 4000 # number of iterations
n_warm <- 1500 # number of iterations in the warm-up phase
n_thin <- 1    # thinning interval
n_seed <- 404

form_1 <- brmsformula(Susc16 ~ 0 + FieldSite + (0 + FieldSite | PopulationCode/Family))
form_2 <- brmsformula(Susc22 ~ 0 + FieldSite + (0 + FieldSite | PopulationCode/Family))

mod <- brm(formula = form_1,
           data = subdata,
           #save_pars = save_pars(group = FALSE),
           chains = n_chains,
           cores = n_chains,
           seed = n_seed,
           iter = n_iter,
           backend = "cmdstanr",
           warmup = n_warm)

saveRDS(mod, file = here(paste0("outputs/ScotsPine/StanModels/Susc16_TwoCGs_MixedModelWithoutGRM.rds")))

mod <- brm(formula = form_2,
           data = subdata,
           #save_pars = save_pars(group = FALSE),
           chains = n_chains,
           cores = n_chains,
           seed = n_seed,
           iter = n_iter,
           backend = "cmdstanr",
           warmup = n_warm)

saveRDS(mod, file = here(paste0("outputs/ScotsPine/StanModels/Susc22_TwoCGs_MixedModelWithoutGRM.rds")))
```


## Model summary

### Response variable: Susc16
  
```{r MixedModelSummarySusc16, fig.height=8, fig.width=10, eval=T}
mod_Susc16 <- readRDS(file = here(paste0("outputs/ScotsPine/StanModels/Susc16_TwoCGs_MixedModelWithoutGRM.rds")))
summary(mod_Susc16)
#plot(mod_Susc16, variable = c("^b_", "sigma"), regex = TRUE)
```

### Response variable: Susc22

```{r MixedModelSummarySusc22, fig.height=8, fig.width=10, eval=T}
mod_Susc22 <- readRDS(file = here(paste0("outputs/ScotsPine/StanModels/Susc22_TwoCGs_MixedModelWithoutGRM.rds")))
summary(mod_Susc22)
#plot(mod_Susc22, variable = c("^b_", "sigma"), regex = TRUE)
```

## Interval plots of the family-level slopes

The graphs below show the family-level reaction norms between the two field sites (slope between FE and FS). Each point represents the median posterior slope for a family, calculated as the difference in predicted trait values between the sites, and the horizontal error bars indicate the 95% credible interval. Families are ordered along the y-axis by increasing median slope. The dashed vertical line at zero represents no difference between sites; families to the right exhibit higher predicted values in FE than FS, while families to the left exhibit higher values in FS than FE.

```{r IntervalPlotFamilyLevelReactionNormNested, fig.height=15, fig.width=8}
interval_plot_family_reaction_norms_nested <- function(mod, y_label){

  family_diff <- mod %>%
    spread_draws(`r_PopulationCode:Family`[PopFam, FieldSite]) %>% 
    # PopFam is of the form "POP:Family"
    mutate(
      PopulationCode = sub("_.*", "", PopFam),
      Family         = sub(".*_", "", PopFam)
    ) %>%
    dplyr::rename(slope = `r_PopulationCode:Family`) %>%
    pivot_wider(names_from = FieldSite, values_from = slope) %>%
    mutate(
      slope_FE_FS = FieldSiteFE - FieldSiteFS
    ) %>%
    group_by(PopulationCode, Family) %>%
    summarise(
      median_slope = median(slope_FE_FS),
      lower_slope  = quantile(slope_FE_FS, 0.025),
      upper_slope  = quantile(slope_FE_FS, 0.975),
      .groups = "drop"
    ) %>%
    arrange(median_slope) %>%
    mutate(Family = factor(Family, levels = Family))

  ggplot(family_diff, aes(x = median_slope, y = Family)) +
    geom_point() +
    geom_errorbar(aes(xmin = lower_slope, xmax = upper_slope), width = 0.2) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
    theme_bw() +
    labs(
      y = "",
      x = "Family-level reaction norms (FE − FS) with 95% credible intervals",
      title = y_label
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}


interval_plot_family_reaction_norms_nested(mod = mod_Susc16, y_label = "Susceptibility index 2016 (Susc16)")
interval_plot_family_reaction_norms_nested(mod = mod_Susc22, y_label = "Susceptibility index 2022 (Susc22)")
```

## Reaction-norm plots

### Observed relationships

We compute and visualize the **observed** (non-modeled) family-level reaction norms of disease indices across the two field sites (FE and FS). More precisely, we directly use the **raw data** to calculate the median disease index per family and site, and we then generate two plots.

The first plot displays the family-level median reaction norms of all families together. The background points represent individual tree observations. The colored lines connect the median trait values per family across the two field sites. Colors correspond to populations. This plot provides an overall view of the range of family-level differentiation in reaction norms.

The second plot shows the same information but faceted by population, allowing clearer visualization of within-population variation in median slopes. Each panel corresponds to one population. Background points represent individual trees. Blue lines connect family median values across sites.


```{r ObservedReactionNormsPlot, fig.height=10, fig.width=10}
observed_family_reaction_norms <- function(mod, subdata, y_var, y_label) {
  
  # Compute family medians per FieldSite
  family_means <- subdata %>%
    group_by(Family, FieldSite) %>%
    summarise(median_y = median(.data[[y_var]], na.rm = TRUE), .groups = "drop") %>%
    mutate(PopulationCode = substr(as.character(Family), 1, 2)) 
  
  # reshape wide to compute slopes
  family_slopes <- family_means %>%
    pivot_wider(names_from = FieldSite, values_from = median_y) %>%
    mutate(slope = FE - FS) %>% # family-level slope = difference between the two sites
    select(Family, PopulationCode, slope)
  
  sum_pop <- family_slopes %>%
    group_by(PopulationCode) %>%
    summarise(
      obs_median_pop_slope = median(slope, na.rm = TRUE), # median of the family slopes
      obs_var_family_slope = var(slope, na.rm = TRUE),
      .groups = "drop")
  
  # Plots
  p_all_reaction_norms <- 
    ggplot() +
    # Points = individual trees, colored by family
    geom_point(data = subdata,
               aes(x = FieldSite, y = .data[[y_var]], colour = PopulationCode),
               position = position_jitter(width = 0.1, height = 0), alpha = 0.5) +
    # Lines = family median across sites
    geom_line(data = family_means,
              aes(x = FieldSite, y = median_y, group = Family, colour = PopulationCode),
              linewidth = 1) +
    # Aesthetic tuning
    theme_minimal(base_size = 14) +
    #theme(legend.position = "none") +
    labs(x = "Field site", y = y_label) +
    scale_colour_viridis_d(option = "turbo", name = "Populations")  # nice categorical palette
  
  
  
  p_reaction_norms_per_pop <- ggplot() +
  # raw data points
  geom_jitter(data = subdata,
              aes(x = FieldSite, y = .data[[y_var]]),
              width = 0.1, alpha = 0.4, show.legend = FALSE) +
  
  # median/mean predictions per family
  geom_line(data = family_means,
            aes(x = FieldSite, y = median_y, group = Family),
            linewidth = 0.8, color = "blue", alpha = 0.3) +

  # facet by population
  facet_wrap(~PopulationCode, scales = "fixed") +
  
  theme_bw() +
  labs(
    x = "Field Site",
    y = y_label,
    title = "Family-level observed reaction norms (median values)"
  ) +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12))
  
  
  return(list(sum_pop = sum_pop,
              p_all_reaction_norms = p_all_reaction_norms,
              p_reaction_norms_per_pop = p_reaction_norms_per_pop))
}

# Run the function for each trait
obs_Susc16 <- observed_family_reaction_norms(mod_Susc16, subdata, "Susc16", y_label = "Susceptibility index 2016 (Susc16)")
obs_Susc22 <- observed_family_reaction_norms(mod_Susc22, subdata, "Susc22", y_label = "Susceptibility index 2022 (Susc22)")


# Join into one table
sum_pop <- obs_Susc16$sum_pop %>%
  rename(obs_median_pop_slope_Susc16 = obs_median_pop_slope,
         obs_var_family_slope_Susc16 = obs_var_family_slope) %>%
  left_join(
    obs_Susc22$sum_pop %>% 
      rename(obs_median_pop_slope_Susc22 = obs_median_pop_slope,
             obs_var_family_slope_Susc22 = obs_var_family_slope),
    by = "PopulationCode"
  )

# Show the plots
obs_Susc16$p_all_reaction_norms
obs_Susc16$p_reaction_norms_per_pop

obs_Susc22$p_all_reaction_norms
obs_Susc22$p_reaction_norms_per_pop
```

For each population and each year, two empirical descriptors were extracted from the raw phenotypic data. First, the *observed median population slope* (`obs_median_pop_slope`) corresponds to the median of the family-level reaction norm slopes, where each family slope is calculated as the difference between its median phenotype at the two field sites (FE – FS). This metric reflects the typical magnitude and direction of population-level plasticity. Second, the *observed variance in family-level slopes* (`obs_var_family_slope`) is the variance of these family-specific slopes within each population and quantifies the heterogeneity of plastic responses among families.

### Fitted relationships

#### Family-level reaction-norms

In the plots below, family-level reaction norms were obtained by generating posterior expected values from the fitted *brms* models using the function `add_epred_draws()` from *tidybayes*. For each combination of *FieldSite × Population × Family*, we constructed a prediction grid and extracted posterior draws of the expected response (`.epred`), specifying `re_formula = ~(0 + FieldSite | PopulationCode/Family)` so that predictions include the fixed *FieldSite* effect, the population-level random slopes, and the family-level random slopes. The function returns the posterior distribution of the linear predictor after marginalising over residual variance.

The first plot show the family-level reaction norms of all families together. Background points correspond to the observed disease susceptibility values, colored by population.

The second plot is faceted by population, allowing visualization of within-population variation in family responses. Background points correspond to the observed disease susceptibility values. The red line represents the global median response across all families, providing a reference for the overall trend.


```{r ReactionNormPlotFamilyLevelNested, fig.width=10, fig.height=10}
family_reaction_norms <- function(mod, subdata, y_var, y_label){
  
  valid_combos <- subdata %>% distinct(PopulationCode, Family)
  newdata <- crossing(FieldSite = unique(subdata$FieldSite), valid_combos)
  
  # Posterior fitted values including BOTH random-effect levels
  preds <- add_epred_draws(
    mod,
    newdata = newdata,
    re_formula = ~(0 + FieldSite | PopulationCode/Family))
  
  # Summaries per family × fieldsite
  family_summary <- preds %>%
    group_by(PopulationCode, Family, FieldSite) %>%
    summarise(
      median_pred       = median(.epred),
      mean_pred         = mean(.epred),
      lower_quant_pred  = quantile(.epred, 0.025),
      upper_quant_pred  = quantile(.epred, 0.975),
      sd_pred           = sd(.epred),
      lower_sd_pred     = mean_pred - sd_pred,
      upper_sd_pred     = mean_pred + sd_pred,
      .groups = "drop"
    )
  
  # Global summary across all families
  global_summary <- preds %>%
    group_by(FieldSite) %>%
    summarise(
      median_all       = median(.epred),
      mean_all         = mean(.epred),
      lower_quant_all  = quantile(.epred, 0.025),
      upper_quant_all  = quantile(.epred, 0.975),
      sd_all           = sd(.epred),
      lower_sd_all     = mean_all - sd_all,
      upper_sd_all     = mean_all + sd_all,
      .groups = "drop"
    )
  
  ## ------------------------------
  ## Plot 1 — All families together
  ## ------------------------------
  
  plot_all_fams <- ggplot() +
    geom_jitter(
      data = subdata,
      aes(x = FieldSite, y = .data[[y_var]], color = PopulationCode),
      width = 0.1, alpha = 0.4, show.legend = FALSE
    ) +
    geom_ribbon(
      data = family_summary,
      aes(x = FieldSite, ymin = lower_quant_pred, ymax = upper_quant_pred, group = Family, fill = PopulationCode),
      alpha = 0.05, show.legend = FALSE
    ) +
    geom_line(
      data = family_summary,
      aes(x = FieldSite, y = median_pred, group = Family, color = PopulationCode),
      linewidth = 0.8
    ) +
    theme_bw() +
    labs(x = "Field Site", y = y_label,
         title = "Family-level reaction norms (median ± 95% CI)") +
    theme(axis.text = element_text(size = 12))
  
  
  ## ----------------------------------
  ## Plot 2 — Faceted by population
  ## ----------------------------------
  
  plot_facet_median_with_CI <- ggplot() +
    geom_jitter(
      data = subdata,
      aes(x = FieldSite, y = .data[[y_var]]),
      width = 0.1, alpha = 0.4
    ) +
    geom_ribbon(
      data = family_summary,
      aes(x = FieldSite, ymin = lower_quant_pred, ymax = upper_quant_pred, group = Family),
      alpha = 0.05, fill = "blue"
    ) +
    geom_line(
      data = family_summary,
      aes(x = FieldSite, y = median_pred, group = Family),
      linewidth = 0.8, color = "blue", alpha = 0.3
    ) +
    geom_ribbon(
      data = global_summary,
      aes(x = FieldSite, ymin = lower_quant_all, ymax = upper_quant_all),
      alpha = 0.2, fill = "red"
    ) +
    geom_line(
      data = global_summary,
      aes(x = FieldSite, y = median_all, group = 1),
      color = "red", linewidth = 1.2
    ) +
    facet_wrap(~PopulationCode, scales = "fixed") +
    theme_bw() +
    labs(
      x = NULL, y = y_label,
      title = "Family-level reaction norms by population (median ± 95% CI)"
    ) +
    theme(axis.text = element_text(size = 12))
  
  
  ## -----------------------------------------
  ## Plot 3 — Mean ± SD version (optional)
  ## -----------------------------------------
  
  plot_facet_mean_with_SD <- ggplot() +
    geom_jitter(
      data = subdata,
      aes(x = FieldSite, y = .data[[y_var]]),
      width = 0.1, alpha = 0.4
    ) +
    geom_ribbon(
      data = family_summary,
      aes(x = FieldSite, ymin = lower_sd_pred, ymax = upper_sd_pred, group = Family),
      alpha = 0.05, fill = "blue"
    ) +
    geom_line(
      data = family_summary,
      aes(x = FieldSite, y = mean_pred, group = Family),
      linewidth = 0.8, color = "blue", alpha = 0.3
    ) +
    geom_line(
      data = global_summary,
      aes(x = FieldSite, y = mean_all, group = 1),
      color = "red", linewidth = 1.2
    ) +
    facet_wrap(~PopulationCode, scales = "fixed") +
    theme_bw() +
    labs(
      x = NULL, y = y_label,
      title = "Family-level fitted reaction norms (mean ± SD)"
    ) +
    theme(axis.text = element_text(size = 12))
  
  
  list(
    family_summary = family_summary,
    plot_all_fams = plot_all_fams,
    plot_facet_median_with_CI = plot_facet_median_with_CI,
    plot_facet_mean_with_SD = plot_facet_mean_with_SD
  )
}


family_reaction_norms_Susc16 <- family_reaction_norms(mod=mod_Susc16, subdata, y_var="Susc16", y_label = "Susceptibility index 2016 (Susc16)")
family_reaction_norms_Susc22 <- family_reaction_norms(mod=mod_Susc22, subdata, y_var="Susc22", y_label = "Susceptibility index 2022 (Susc22)")

family_reaction_norms_Susc16$plot_all_fams
family_reaction_norms_Susc22$plot_all_fams

family_reaction_norms_Susc16$plot_facet_median_with_CI
family_reaction_norms_Susc22$plot_facet_median_with_CI
```

#### Population-level reaction-norms


```{r ReactionNormPlotPopulationLevelNested, fig.width=8, fig.height=8}
population_reaction_norms <- function(mod, subdata, y_var, y_label) {
  
  # Unique population × field site combinations
  valid_pops <- subdata %>% distinct(PopulationCode)
  newdata <- crossing(FieldSite = unique(subdata$FieldSite), valid_pops)
  
  # Posterior fitted values including only population-level random effects
  preds <- add_epred_draws(
    mod,
    newdata = newdata,
    re_formula = ~(0 + FieldSite | PopulationCode)
  )
  
  # Summarise posterior per population × field site
  pop_summary <- preds %>%
    group_by(PopulationCode, FieldSite) %>%
    summarise(
      median_pred      = median(.epred),
      mean_pred        = mean(.epred),
      lower_quant_pred = quantile(.epred, 0.025),
      upper_quant_pred = quantile(.epred, 0.975),
      sd_pred          = sd(.epred),
      lower_sd_pred    = mean_pred - sd_pred,
      upper_sd_pred    = mean_pred + sd_pred,
      .groups = "drop"
    )
  
  ## ------------------------------
  ## Plot 1 — All populations together
  ## ------------------------------
  plot_all_pops <- ggplot() +
    geom_jitter(
      data = subdata,
      aes(x = FieldSite, y = .data[[y_var]], color = PopulationCode),
      width = 0.1, alpha = 0.4
    ) +
    geom_ribbon(
      data = pop_summary,
      aes(x = FieldSite, ymin = lower_quant_pred, ymax = upper_quant_pred, group = PopulationCode, fill = PopulationCode),
      alpha = 0.1, show.legend = FALSE
    ) +
    geom_line(
      data = pop_summary,
      aes(x = FieldSite, y = median_pred, group = PopulationCode, color = PopulationCode),
      linewidth = 1
    ) +
    theme_bw() +
    labs(
      x = "Field Site",
      y = y_label,
      title = "Population-level reaction norms (median ± 95% CI)"
    ) +
    theme(axis.text = element_text(size = 12))
  
  ## ------------------------------
  ## Plot 2 — Faceted by population
  ## ------------------------------
  plot_facet_median_with_CI <- ggplot() +
    geom_jitter(
      data = subdata,
      aes(x = FieldSite, y = .data[[y_var]]),
      width = 0.1, alpha = 0.4
    ) +
    geom_ribbon(
      data = pop_summary,
      aes(x = FieldSite, ymin = lower_quant_pred, ymax = upper_quant_pred),
      alpha = 0.2, fill = "blue"
    ) +
    geom_line(
      data = pop_summary,
      aes(x = FieldSite, y = median_pred, group = 1),
      color = "blue", linewidth = 1
    ) +
    facet_wrap(~PopulationCode, scales = "fixed") +
    theme_bw() +
    labs(
      x = "Field Site",
      y = y_label,
      title = "Population-level reaction norms by population (median ± 95% CI)"
    ) +
    theme(axis.text = element_text(size = 12))
  
  list(
    pop_summary = pop_summary,
    plot_all_pops = plot_all_pops
  )
}

# ------------------------------
# Apply function to PC1 and PC2
# ------------------------------
pop_reaction_norms_Susc16 <- population_reaction_norms(mod = mod_Susc16, subdata, y_var = "Susc16", y_label = "Susceptibility index 2016 (Susc16)")
pop_reaction_norms_Susc22 <- population_reaction_norms(mod = mod_Susc22, subdata, y_var = "Susc22", y_label = "Susceptibility index 2022 (Susc22)")

# Display plots
pop_reaction_norms_Susc16$plot_all_pops
pop_reaction_norms_Susc22$plot_all_pops
#pop_reaction_norms_Susc16$pop_summary
```


#### Fitted parameters of interest

For each response variable (`Susc16` and `Susc22`), we used posterior expected values (`epred`) to compute, for every *Population × Family*, the difference in disease index between the two field sites (FE – FS). This difference represents the reaction-norm slope.

From these slopes, two population-level parameters were extracted:

  - `fitted_median_pop_slope`: The median reaction-norm slope across families within a population. Quantifies the average change in disease susceptibility between sites for that population.
  
  - `fitted_var_family_slope`: The median variance of family-specific slopes within a population. Measures the within-population variation in responses to the site contrast.

These two parameters summarize, respectively, the population-level environmental response and the heterogeneity among families in that response.

```{r ExtractParamsOfInterest}
extract_slopes_epred <- function(model, subdata) {
  
  # Get all population/family combinations
  newdata <- subdata %>% dplyr::select(FieldSite, PopulationCode, Family) %>% distinct

  # Posterior predictions including both population and family effects
  preds <- add_epred_draws(model, newdata = newdata, re_formula = ~(0 + FieldSite | PopulationCode/Family)) %>% 
    ungroup %>% 
    select(FieldSite, PopulationCode, Family, .draw, .epred)
  
  # Compute slope for each draw: FE - FS per family
  slopes <- preds %>%
    pivot_wider(
      names_from = FieldSite,
      values_from = .epred
    ) %>%
    mutate(
      fam_slope = FE -FS
    )
  
  # Summarize per draw & population
  slopes_by_draw <- slopes %>%
    group_by(.draw, PopulationCode) %>%
    arrange(PopulationCode) %>% 
    summarise(
      pop_slope = median(fam_slope),
      var_family_slope = if(n() > 1) var(fam_slope) else 0, # zero if only one family
      .groups = "drop"
    )
  
  # Summarize across draws
  summary_pop <- slopes_by_draw %>%
    group_by(PopulationCode) %>%
    summarise(
      fitted_median_pop_slope = median(pop_slope),
      fitted_var_family_slope = median(var_family_slope),
      .groups = "drop"
    )
  
  summary_pop
}

# Apply to both models
summary_Susc16 <- extract_slopes_epred(mod_Susc16, subdata)
summary_Susc22 <- extract_slopes_epred(mod_Susc22, subdata)

# Join into one table
sum_pop <- summary_Susc16 %>%
  rename(fitted_median_pop_slope_Susc16 = fitted_median_pop_slope,
         fitted_var_family_slope_Susc16 = fitted_var_family_slope) %>%
  left_join(
    summary_Susc22 %>% 
      rename(fitted_median_pop_slope_Susc22 = fitted_median_pop_slope,
             fitted_var_family_slope_Susc22 = fitted_var_family_slope),
    by = "PopulationCode"
  ) %>% left_join(sum_pop, by="PopulationCode")

sum_pop %>%  kable_mydf()
```


# Mean trait values{#sec-MeanTraitValues}

We extract the mean trait values of each population in each common garden.

```{r ExtractMeanTraitValue}
# Function to compute mean phenotype per population and field site
compute_means_per_site <- function(subdata, trait) {
  subdata %>%
    group_by(PopulationCode, FieldSite) %>%
    summarise(mean_y = mean(.data[[trait]], na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(
      names_from = FieldSite,
      values_from = mean_y,
      names_prefix = paste0("mean_", trait, "_")
    )
}

# Compute means for both traits
mean_Susc16 <- compute_means_per_site(subdata, "Susc16")
mean_Susc22 <- compute_means_per_site(subdata, "Susc22")

# Merge with your existing summary table
sum_pop <- sum_pop %>%
  left_join(mean_Susc16, by = "PopulationCode") %>%
  left_join(mean_Susc22, by = "PopulationCode")
```

# RDPI{#sec-RDPI}


In this section, we calculate the relative distance plasticity index (RDPI) from @valladaresQuantitativeEstimationPhenotypic2006. In this paper, the authors propose to quantify phenotypic plasticity based on phenotypic distances among individuals of a given species (in our study, population) exposed to different environments (in our study, planted in different common gardens), which is summarized in the RDPI index that allows for statistical comparisons of phenotypic plasticity between species (in our study, populations). 

For a given trait, the relative distance plasticity index $RDPI_p$ for a given population $p$ can be calculated as follows:

$$RDPI_p = \frac{\sum_{n=1}^N{\frac{x_{ij}-x_{i'j'}}{x_{ij}+x_{i'j'}}}}{N}$$

where $N$ is the number of pairwise comparisons among individuals of the same population but from different common gardens. $x_{ij}$ and $x_{i'j'}$ are the phenotypic values of individual $j$ and $j'$ in common gardens $i$ and $i’$. As such, the phenotypic plasticity of a population for a given trait corresponds to the sum of the relative phenotypic distances for all pairs of individuals of the population grown in different common gardens. 

We also calculate the RDPI at the population level. In this case, we first compute the mean trait value of each population in each common garden, and then calculate RDPI based on these population means. The population-mean RDPI therefore summarizes overall population-level plasticity, integrating the average phenotypic shift of the population across environments while ignoring within-population variation among individuals.


```{r CalcRDPI}
## ---- RDPI computation for Susc16 and Susc22 ----

compute_RDPI <- function(subdata, trait) {
  
  # Rename target trait to generic 'y'
  subdata <- subdata %>% mutate(y = .data[[trait]])
  
  # Identify populations and field sites
  pop_codes <- unique(subdata$PopulationCode) %>% sort()
  site_codes <- unique(subdata$FieldSite) %>% sort()
  site_comb <- combn(site_codes, 2)
  calc_rel_dist <- function(x) abs(x[1] - x[2]) / (x[1] + x[2])

  # Loop over populations
  res_list <- lapply(pop_codes, function(pop) {
    subpop <- subdata %>% filter(PopulationCode == pop)

    # --- Individual-based RDPI ---
    rel_dist_ind <- lapply(1:ncol(site_comb), function(nb_comb) {
      subsite1 <- subpop %>% filter(FieldSite == site_comb[1, nb_comb])
      subsite2 <- subpop %>% filter(FieldSite == site_comb[2, nb_comb])
      
      if (nrow(subsite1) == 0 | nrow(subsite2) == 0) return(NULL)
      
      lapply(1:nrow(subsite1), function(x) {
        rbind(subsite1[x, "y"], subsite2[, "y"]) %>%
          pull() %>%
          combn(2, calc_rel_dist) %>%
          .[1:nrow(subsite2)] %>%
          as.numeric()
      }) %>% unlist()
    }) %>% unlist()

    N_ind <- length(rel_dist_ind)
    ind_RDPI <- if (N_ind == 0) NA_real_ else sum(rel_dist_ind) / N_ind

    # --- Population-mean RDPI ---
    mean_tab <- subpop %>%
      drop_na(y) %>%
      group_by(FieldSite) %>%
      summarise(pop_mean = mean(y, na.rm = TRUE), .groups = "drop")

    rel_dist_pop_means <- lapply(1:ncol(site_comb), function(nb_comb) {
      s1 <- site_comb[1, nb_comb]
      s2 <- site_comb[2, nb_comb]
      mean1 <- mean_tab %>% filter(FieldSite == s1) %>% pull(pop_mean)
      mean2 <- mean_tab %>% filter(FieldSite == s2) %>% pull(pop_mean)
      if (length(mean1) == 0 || length(mean2) == 0) return(NA_real_)
      denom <- mean1 + mean2
      if (denom == 0) return(NA_real_) else return(abs(mean1 - mean2) / denom)
    }) %>% unlist()

    pop_mean_RDPI <- if (all(is.na(rel_dist_pop_means))) NA_real_ else mean(rel_dist_pop_means, na.rm = TRUE)

    tibble(PopulationCode = pop,
           ind_RDPI = ind_RDPI,
           pop_mean_RDPI = pop_mean_RDPI)
  })

  df_rdpi <- bind_rows(res_list) %>%
    mutate(across(c(ind_RDPI, pop_mean_RDPI), as.numeric))

  # Rename columns with trait suffix
  colnames(df_rdpi)[2:3] <- paste0(colnames(df_rdpi)[2:3], "_", trait)
  
  return(df_rdpi)
}

# ---- Apply function to both traits ----
df_rdpi_Susc16 <- compute_RDPI(subdata, "Susc16")
df_rdpi_Susc22 <- compute_RDPI(subdata, "Susc22")

# ---- Merge all results ----
sum_pop <- sum_pop %>%
  left_join(df_rdpi_Susc16, by = "PopulationCode") %>%
  left_join(df_rdpi_Susc22, by = "PopulationCode")

# Final summary
#saveRDS(sum_pop, here("outputs/ScotsPine/PhenotypicPlasticity/DiseaseSusceptibilityIndexesSummary.rds"))
sum_pop %>% kable_mydf()
```

# Relationships among all the estimated indexes{#sec-IndexRelationships}

Description for each variable:

  - `PopulationCode`: Identifier of the population.

  - `fitted_median_pop_slope_Susc16` and `fitted_median_pop_slope_Susc22`: Median reaction-norm slope of a population predicted by the model, computed from posterior draws and averaged across its families. Summarizing the predicted population-level difference in disease index between field sites.

  - `fitted_var_family_slope_Susc16` and `fitted_var_family_slope_Susc22`: Model-based variance in family-level slopes within a population, quantifying predicted among-family heterogeneity in reaction norms (i.e., within-population variation in responses).
  
  - `obs_median_pop_slope_Susc16` and `obs_median_pop_slope_Susc22`: Observed median reaction-norm slope (FE – FS) of a population, calculated directly from the empirical family slopes. Summarizing the observed population-level difference in disease susceptibility between the two field sites.
  
  - `obs_var_family_slope_Susc16` and `obs_var_family_slope_Susc22`: observed variance in family-level slopes within a population; an empirical measure of how differently families respond across sites.

  - `mean_FE_Susc16` and `mean_FE_Susc22`: Mean susceptibility value of the population in the FE site.

  - `mean_FS_Susc16` and `mean_FS_Susc22`: Mean susceptibility value of the population in the FS site.

  - `pop_mean_RDPI_Susc16` and `pop_mean_RDPI_Susc22`: RDPI computed from population means, summarizing average phenotypic divergence across sites.
  
  - `ind_RDPI_Susc16` and `ind_RDPI_Susc22`: Individual-based RDPI, quantifying relative phenotypic change across sites among individuals within the population.


```{r RenameMeanTraitValues}
# Rename columns so FE/FS comes first
sum_pop <- sum_pop %>%
  rename(
    mean_FE_Susc16 = mean_Susc16_FE,
    mean_FS_Susc16 = mean_Susc16_FS,
    mean_FE_Susc22 = mean_Susc22_FE,
    mean_FS_Susc22 = mean_Susc22_FS)

sum_pop %>% kable_mydf
```

## PCA

```{r PCACalculatedIndexes, fig.height=8, fig.width=8}
# keep only populations with complete data for PCA (or use na.omit)
pca_df <- sum_pop %>% drop_na()

# run PCA (center & scale)
pca_res <- prcomp(pca_df %>% select(-PopulationCode), center = TRUE, scale. = TRUE)

# scores (PC coordinates for populations)
scores <- as.data.frame(pca_res$x)
scores$PopulationCode <- pca_df$PopulationCode

# loadings (variable contributions)
loadings <- as.data.frame(pca_res$rotation)
loadings$var <- rownames(loadings)

# get % variance explained for axis labels
eig <- (pca_res$sdev)^2
pct_var <- round(100 * eig / sum(eig), 1)

# scale loadings to make arrows visible on the same plotting scale as scores
# choose a multiplier so arrows reach ~30-50% of the score cloud
mult <- min(
  (max(scores$PC1) - min(scores$PC1)) / (max(loadings$PC1) - min(loadings$PC1)),
  (max(scores$PC2) - min(scores$PC2)) / (max(loadings$PC2) - min(loadings$PC2))
) * 0.6

loadings_plot <- loadings %>%
  mutate(arrow_x = PC1 * mult,
         arrow_y = PC2 * mult)

# create a color palette with one color per population
n_pops <- nrow(scores)
pal <- scales::hue_pal()(n_pops) # distinct hues; change palette if you prefer

# Plot
ggplot() +
  # points for populations
  geom_point(data = scores, aes(x = PC1, y = PC2, color = PopulationCode), size = 3, alpha = 0.9) +
  # population labels
  geom_text_repel(data = scores, aes(x = PC1, y = PC2, label = PopulationCode, color = PopulationCode),
                  size = 3.5, show.legend = FALSE, max.overlaps = 100) +
  # arrows for variable loadings (from origin to scaled loading)
  geom_segment(data = loadings_plot,
               aes(x = 0, y = 0, xend = arrow_x, yend = arrow_y),
               arrow = arrow(length = unit(0.25, "cm")), colour = "black", linewidth = 0.6) +
  # variable labels at arrow tips
  geom_text_repel(data = loadings_plot,
                  aes(x = arrow_x, y = arrow_y, label = var),
                  color = "black", size = 4, segment.size = 0.3) +
  scale_color_manual(values = pal) +
  labs(x = paste0("PC1 (", pct_var[1], "%)"),
       y = paste0("PC2 (", pct_var[2], "%)"),
       title = "",
       color = "Population") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none") # legend suppressed because each pop has its own color label

```

## Correlations 

```{r PCAonCalculatedPhenoIndexes, fig.height=10, fig.width=10}
sum_pop %>% 
  dplyr::select(-PopulationCode) %>% 
  cor() %>% 
  corrplot(method = 'number',
           type = 'lower', diag = FALSE,
           title=paste0(""),
           mar=c(0,0,2,0),
           number.cex=0.9,tl.cex=0.9)
```


## Box plots

```{r BoxPlots, fig.height=5, fig.width=10}
# Pivot to long format
sum_pop_long <- sum_pop %>%
  pivot_longer(
    cols = -PopulationCode,
    names_to = c("variable", "Susc"),
    names_pattern = "(.*_FE|.*_FS|.*)_Susc(\\d+)$",
    values_to = "value"
  ) %>% 
  mutate(Year = paste0("20",Susc))

# --- ensure correct variable grouping ---
sum_pop_long$variable <- factor(sum_pop_long$variable,
                                levels = c("fitted_median_pop_slope",
                                           "fitted_var_family_slope",
                                           "obs_median_pop_slope",
                                           "obs_var_family_slope",
                                           "mean_FS",
                                           "mean_FE",
                                           "ind_RDPI",
                                           "pop_mean_RDPI"))

# --- boxplot with points behind ---
# p_all <- ggplot(sum_pop_long, aes(x = variable, y = value, fill = Year)) +
#   geom_jitter(aes(color = Year), 
#               width = 0.15, alpha = 0.5, size = 2, show.legend = FALSE) +  # points behind boxes
#   geom_boxplot(alpha = 0.7, width = 0.6, outlier.shape = NA, color = "black") +
#   scale_fill_manual(values = c("#1f77b4","#d62728"),name = "Year") +
#   scale_color_manual(values = c("#1f77b4","#d62728"),name = "Year") +
#   theme_bw(base_size = 13) +
#   theme(
#     axis.text.x = element_text(angle = 45, hjust = 1),
#     panel.grid.minor = element_blank()
#   ) +
#   labs(
#     x = "Variable type",
#     y = "Value",
#     title = "All indexes"
#   )

# Split long table into two sets of variables
pop_traits <- c("mean_FS","mean_FE")
pop_slopes <- c("ind_RDPI", "obs_median_pop_slope", "fitted_median_pop_slope", "pop_mean_RDPI")
pop_vars <- c("obs_var_family_slope", "fitted_var_family_slope")

df_traits <- sum_pop_long %>% filter(variable %in% pop_traits)
df_slopes <- sum_pop_long %>% filter(variable %in% pop_slopes)
df_vars <- sum_pop_long %>% filter(variable %in% pop_vars)

# Traits
ggplot(df_traits, aes(x = variable, y = value, fill = Year)) +
  geom_jitter(aes(color = Susc), width = 0.15, alpha = 0.5, size = 2, show.legend = FALSE) +
  geom_boxplot(alpha = 0.7, width = 0.6, outlier.shape = NA, color = "black") +
  scale_fill_manual(values = c("#1f77b4","#d62728"),name = "Year") +
  scale_color_manual(values = c("#1f77b4","#d62728"),name = "Year") +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.minor = element_blank()) +
  labs(x = "", y = "Index value", title = "Population mean trait values in each field site")

ggplot(df_slopes, aes(x = variable, y = value, fill = Year)) +
  geom_jitter(aes(color = Susc), width = 0.15, alpha = 0.5, size = 2, show.legend = FALSE) +
  geom_boxplot(alpha = 0.7, width = 0.6, outlier.shape = NA, color = "black") +
  scale_fill_manual(values = c("#1f77b4","#d62728"),name = "Year") +
  scale_color_manual(values = c("#1f77b4","#d62728"),name = "Year") +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.minor = element_blank()) +
  labs(x = "", y = "Index value", title = "Population median plastic responses")

ggplot(df_vars, aes(x = Year, y = value, fill = Year)) +
  geom_boxplot(alpha = 0.7, width = 0.6, outlier.shape = NA, color = "black") +
  geom_jitter(aes(color = Year), width = 0.15, alpha = 0.5, size = 2, show.legend = FALSE) +
  scale_fill_manual(values = c("2016" = "#1f77b4", "2022" = "#d62728")) +
  scale_color_manual(values = c("2016" = "#1f77b4", "2022" = "#d62728")) +
  facet_wrap(~ variable, scales = "free_y") +
  theme_bw(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "", y = "Value", title = "Fitted and observed variance of the family-level slopes")
```


## Scatter plots among indexes

Some scatter plots.

```{r ScatterPlotsAmongIndexes, fig.height=7, fig.width=9}

# Select metrics for both years
df_diag <- sum_pop %>%
  select(PopulationCode,
         obs_median_pop_slope_Susc16, fitted_median_pop_slope_Susc16, pop_mean_RDPI_Susc16,ind_RDPI_Susc16,
         fitted_var_family_slope_Susc16, fitted_var_family_slope_Susc22, obs_var_family_slope_Susc16, obs_var_family_slope_Susc22,
         obs_median_pop_slope_Susc22, fitted_median_pop_slope_Susc22, pop_mean_RDPI_Susc22,ind_RDPI_Susc22) %>%
  pivot_longer(
    cols = -PopulationCode,
    names_to = c("metric", "year"),
    names_sep = "_Susc"
  ) %>%
  pivot_wider(names_from = metric, values_from = value) %>% 
  mutate(year=paste0("20",year))


pop_n <- subdata %>% dplyr::count(PopulationCode, name = "n_pop")
df_diag <- df_diag %>%  left_join(pop_n, by = "PopulationCode")

# ggplot(df_diag,
#        aes(x = fitted_median_pop_slope,
#            y = obs_median_pop_slope)) +
# 
#   geom_point(
#     aes(color = year, size = n_pop),
#     alpha = 0.7
#   ) +
# 
#   geom_text(
#     aes(label = PopulationCode),
#     vjust = -0.5,
#     size = 3
#   ) +
# 
#   geom_smooth(
#     aes(color = year, group = year),
#     method = "lm",
#     se = FALSE,
#     linewidth = 1
#   ) +
# 
#   scale_size_continuous(name = "Sample size") +
# 
#   labs(
#     x = "Fitted median population slope",
#     y = "Observed median population slope",
#     title = "Relationship between observed and fitted population slopes"
#   ) +
#   theme_bw(base_size = 14)

# Variables to compare
vars <- c(
  "fitted_median_pop_slope",
  "obs_median_pop_slope",
  "fitted_var_family_slope",
  "obs_var_family_slope",
  "pop_mean_RDPI",
  "ind_RDPI"
)

# All pairwise combinations
pairs <- t(combn(vars, 2)) %>% as.data.frame()
colnames(pairs) <- c("xvar", "yvar")


# Function to build one plot
make_pairplot <- function(xvar, yvar, df) {
  
  ggplot(df, aes(x = .data[[xvar]], y = .data[[yvar]])) +
    
    geom_point(
      aes(color = year, size = n_pop),
      alpha = 0.7
    ) +
    
    geom_text(
      aes(label = PopulationCode),
      vjust = -0.5,
      size = 3
    ) +
    
    geom_smooth(
      aes(color = year, group = year),
      method = "lm",
      se = FALSE,
      formula = y ~ x,
      linewidth = 1
    ) +
    
    scale_size_continuous(name = "Sample size") +
    
    labs(
      x = xvar,
      y = yvar,
      title = paste(yvar, "vs", xvar)
    ) +
    theme_bw(base_size = 14)
}

# Build all plots
pairplots <- pmap(
  pairs,
  ~ make_pairplot(..1, ..2, df_diag)
)

# Name them
names(pairplots) <- paste0(pairs$yvar, "_vs_", pairs$xvar)

# Display some plots
pairplots$fitted_var_family_slope_vs_fitted_median_pop_slope
pairplots$obs_var_family_slope_vs_obs_median_pop_slope
pairplots$obs_var_family_slope_vs_fitted_var_family_slope
pairplots$obs_median_pop_slope_vs_fitted_median_pop_slope
pairplots$ind_RDPI_vs_pop_mean_RDPI
pairplots$ind_RDPI_vs_fitted_var_family_slope
pairplots$ind_RDPI_vs_fitted_median_pop_slope 
```


# Correlations with genomic offset predictions{#sec-CorrWithGenomicOffset}

```{r DownloadGOpred}
df_go_clim_dist <- readRDS(here("outputs/ScotsPine/df_gen_off_pred_clim_dist.rds"))

df_go_wide <- df_go_clim_dist %>%
  select(PopulationCode, rcp_scenario, pred_timeslice, method, ensemble_member, var_code,val) %>%
  mutate(var_name_comb = paste(rcp_scenario, pred_timeslice, method, ensemble_member, var_code,sep = "_") %>%  gsub("[ .-]", "", .)) %>%
  dplyr::select(PopulationCode, var_name_comb,val) %>% 
  pivot_wider(names_from = var_name_comb, values_from = val)


df_go_wide <- df_go_clim_dist %>%
  dplyr::filter(var_code == "all_snps" & rcp_scenario == "RCP2.6" & pred_timeslice == "2030-2050" & ensemble_member == "Average across EMs") %>% 
  select(PopulationCode, method, val) %>%
  mutate(var_name_comb = paste(method, sep = "_") %>%  gsub("[ .-]", "", .)) %>%
  dplyr::select(PopulationCode, var_name_comb, val) %>% 
  pivot_wider(names_from = var_name_comb, values_from = val)
```

## PCA

```{r PCAwithGO, fig.height=8, fig.width=8}
# keep only populations with complete data for PCA (or use na.omit)
pca_df <- sum_pop %>% left_join(df_go_wide, by="PopulationCode")

# run PCA (center & scale)
pca_res <- prcomp(pca_df %>% select(-PopulationCode), center = TRUE, scale. = TRUE)

# scores (PC coordinates for populations)
scores <- as.data.frame(pca_res$x)
scores$PopulationCode <- pca_df$PopulationCode

# loadings (variable contributions)
loadings <- as.data.frame(pca_res$rotation)
loadings$var <- rownames(loadings)

# get % variance explained for axis labels
eig <- (pca_res$sdev)^2
pct_var <- round(100 * eig / sum(eig), 1)

# scale loadings to make arrows visible on the same plotting scale as scores
# choose a multiplier so arrows reach ~30-50% of the score cloud
mult <- min(
  (max(scores$PC1) - min(scores$PC1)) / (max(loadings$PC1) - min(loadings$PC1)),
  (max(scores$PC2) - min(scores$PC2)) / (max(loadings$PC2) - min(loadings$PC2))
) * 0.6

loadings_plot <- loadings %>%
  mutate(arrow_x = PC1 * mult,
         arrow_y = PC2 * mult)

# create a color palette with one color per population
n_pops <- nrow(scores)
pal <- scales::hue_pal()(n_pops) # distinct hues; change palette if you prefer

# Plot
ggplot() +
  # points for populations
  geom_point(data = scores, aes(x = PC1, y = PC2, color = PopulationCode), size = 3, alpha = 0.9) +
  # population labels
  geom_text_repel(data = scores, aes(x = PC1, y = PC2, label = PopulationCode, color = PopulationCode),
                  size = 3.5, show.legend = FALSE, max.overlaps = Inf) +
  # arrows for variable loadings (from origin to scaled loading)
  geom_segment(data = loadings_plot,
               aes(x = 0, y = 0, xend = arrow_x, yend = arrow_y),
               arrow = arrow(length = unit(0.25, "cm")), colour = "black", linewidth = 0.6) +
  # variable labels at arrow tips
  geom_text_repel(data = loadings_plot,
                  aes(x = arrow_x, y = arrow_y, label = var),
                  color = "black", size = 4, segment.size = 0.3, max.overlaps = Inf) +
  scale_color_manual(values = pal) +
  labs(x = paste0("PC1 (", pct_var[1], "%)"),
       y = paste0("PC2 (", pct_var[2], "%)"),
       title = "",
       color = "Population") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none") # legend suppressed because each pop has its own color label

```
 
 
## Correlations

```{r CorrelationsWithGO, fig.height=12, fig.width=12}
sum_pop %>% left_join(df_go_wide, by="PopulationCode") %>% 
  dplyr::select(-PopulationCode) %>% 
  cor() %>% 
  corrplot(method = 'number',
           type = 'lower', diag = FALSE,
           title=paste0(""),
           mar=c(0,0,2,0),
           number.cex=0.9,tl.cex=0.9)
```

## Scatter plots with RDA predictions

```{r PlasticityIndexesVsRDApredictionsScatterPlots, fig.height=7, fig.width=9}
df_diag <- df_diag %>% left_join(df_go_wide[,c("PopulationCode", "RDA")],by = "PopulationCode")

# Build a tibble of pairs: each xvar paired with RDA
pairs <- tibble(
  xvar = vars,
  yvar = "RDA"
)

# Function to build one plot
make_RDA_plot <- function(xvar, yvar, df) {
  
  ggplot(df, aes(x = .data[[xvar]], y = .data[[yvar]])) +
    
  geom_point(
    aes(color = year, size = n_pop),
    alpha = 0.7
  ) +
    
    geom_text(
      aes(label = PopulationCode),
      vjust = -0.5,
      size = 3
    ) +
    
    geom_smooth(
      aes(color = year, group = year),
      method = "lm",
      formula = y ~ x,
      se = FALSE,
      linewidth = 1
    ) +
    scale_size_continuous(name = "Sample size") +
    
    labs(
      x = xvar,
      y = yvar,
      title = paste(yvar, "vs", xvar)
    ) +
    
    theme_bw(base_size = 14)
}

# Build all plots
RDA_plots <- pmap(
  pairs,
  ~ make_RDA_plot(..1, ..2, df_diag)
)

# Name the plots
names(RDA_plots) <- paste0("RDA_vs_", vars)

# Example: display one plot
for(p in RDA_plots) print(p)
```


