---
title: "Scots pine genomic data"
subtitle: "Formatting, filtering, imputation of missing data, genetic relationships"
author: "Juliette Archambeau"
date: "`r format(Sys.time(), '%d %B, %Y')`"
number-sections: true
format: 
  html:
    toc: true
    toc-depth: 4
    code-fold: true
    page-layout: full
embed-resources: true
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

<style type="text/css">
body {
   font-size: 15px;
}
code.r{
  font-size: 11px;
}
pre {
  font-size: 11px
}

table {
  font-size: 10px
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 7,fig.height = 5,cache=F)
options(width = 300)
library(knitr)
library(dplyr)
library(tidyverse)
library(readxl)
library(magrittr)
library(kableExtra)
library(genetics)
library(xtable)
library(reshape2)
library(janitor)
library(diffdf)
library(parallel)
library(here)
library(AGHmatrix) # to estimate the genetic relatedness matrix 

# my own function for building tables in reports
source(here("scripts/functions/kable_mydf.R"))
```

# Raw genomic data

## Loading

Here are the 10 first rows and columns of the original genomic dataset.

```{r LoadRawGenomicData}
# Load the raw genomic data:
raw_data <- read_table(here("data/ScotsPine/GenomicData/data_update_Fev2025/raw_data/PineTrial_genotypes.txt"),show_col_types=F) %>%
  dplyr::select(probeset_id,sort(colnames(.)))

# viz
raw_data[1:10,1:10] %>% kable_mydf(boldfirstcolumn = T, font_size = 11)
```

The original genomic dataset includes **`r ncol(raw_data)-1` individuals** (columns) and **`r nrow(raw_data)` SNPs** (rows).

```{r FormattingRawGenomicDataFromLettersToNumbers, eval=F}
# We apply a function to each column (i.e SNP) of the raw genomic data, 
# after removing the first two columns, i.e. the clone ID and the assay in which the clone was genotyped

data <- raw_data %>% column_to_rownames(var="probeset_id") %>% t() %>% as.data.frame()

list_alleles <- lapply(colnames(data), function(x){ 
  
  vec <- data[,x]
  
  tab <- table(vec) %>%  as.matrix() %>%  t() %>% as.data.frame() %>% dplyr::select(-contains("NN"))
  
  
  if(length(colnames(tab))==1){         # monomorphic case, e.g. only A/A
    major <- names(which.max(tab))
    
    case_when(vec == major ~ 0) 
    
  } else if(length(colnames(tab))==2){  # case where there is no homozygote with the rare allele, e.g. A/A and A/G
    
    major <- names(which.max(tab))
    mid <- names(which.min(tab))
    
    case_when(vec == major ~ 0,
              vec == mid   ~ 1) 
    
  } else if (length(colnames(tab))==3){ # common case, e.g. A/A, A/G and G/G
    
    if(tab[,1]==tab[,3]){               # to account for cases where there is the same number of A/A and G/G 
      major <- colnames(tab)[1]
      minor <- colnames(tab)[3]
      } else {
    major <- names(which.max(tab[,c(1,3)]))
    minor <- names(which.min(tab[,c(1,3)]))
      }
    mid <- setdiff(names(tab),c(major,minor))  
    
    case_when(vec == major ~ 0,
              vec == minor ~ 2,
              vec == mid   ~ 1) 
  }
}) %>% setNames(colnames(data)) # re-attribute the SNP ID to list names

df <- list_alleles %>% 
  as_tibble() %>% 
  t() %>% 
  set_colnames(rownames(data)) %>% # attribute individual ID
  as.data.frame()

# save the dataset
df %>% saveRDS(here("data/ScotsPine/GenomicData/data_update_Fev2025/formatted_data/unfiltered_allele_counts.rds"))
```

Here are the 10 first rows and columns of the formatted dataset. 

```{r LoadgGenomicData}
rm(raw_data)
df <- readRDS(here("data/ScotsPine/GenomicData/data_update_Fev2025/formatted_data/unfiltered_allele_counts.rds"))

# viz
df[1:10,1:10] %>% kable_mydf(boldfirstcolumn = T, font_size = 11)
```



# Filtering genomic data

## MAC and monomorphic SNPs{#MAC}

```{r FunctionCalculateMAC}
# function to calculate the minor allele counts
sum_minor_allele <- function(x){df %>% 
  t() %>% 
  as_tibble() %>% 
  sapply(function(x) sum(x,na.rm=T)) %>% 
  as.data.frame() %>%
  set_colnames(c("sum")) %>% 
  rownames_to_column(var = "snp") %>% 
  as_tibble() 
}
```

```{r IdentifyingMonomorphicSNPs, eval=T}
mono <- df %>% sum_minor_allele %>% filter(sum==0)
low_mac <- df %>% sum_minor_allele %>% filter(sum %in% c(1,2))
```

`r nrow(mono)` SNPs are monomorphic. We remove them.

`r nrow(low_mac)` SNPs have one or two minor allele count (MAC). We also remove these SNPs as the minor alleles may be due to PCR or sequencing errors.


```{r RemovingMonomorphicSNPs, eval=T}
df <- df[!(row.names(df) %in% mono$snp),]
df <- df[!(row.names(df) %in% low_mac$snp),]
```

<span style="color: red;">After filtering monomorphic SNPs and MAC, `r nrow(df)` SNPs left.</span>


## NAs and MAF

### Missing data per clone

```{r MissingDataPerIndividual, warning=F, fig.height=6,fig.width=8}
tab <- df %>% 
  as_tibble() %>% 
  sapply(function(x) sum(is.na(x))*100/nrow(df)) %>% 
  as.data.frame() %>%
  set_colnames(c("na_freq")) %>% 
  rownames_to_column(var = "clone") %>% 
  as_tibble() %>% 
  arrange(na_freq) %>% 
  mutate(id=1:nrow(.))

plot(tab$id,tab$na_freq,
     pch=20,
     xlab="Individual number",
     ylab="Percentage of missing data per individual")
```

All individuals have less than `r max(tab$na_freq) %>% round(2)`% missing data. Therefore, we do not apply a filter of missing data per individual. 

### Minor allele frequencies

We want to remove *minor allele frequencies* (MAF) for gene-environment association analyses, but not for population genetic structure estimation. Therefore, we produce two datasets: one with MAF and the other without.

```{r MAF, warning=F, fig.height=6,fig.width=8}
tab <- df %>% 
  t() %>% 
  as_tibble() %>% 
  sapply(function(x) sum(x,na.rm=T)*100/(2*nrow(.))) %>% 
  as.data.frame() %>%
  set_colnames(c("MAF")) %>% 
  rownames_to_column(var = "snp") %>% 
  as_tibble() %>% 
  arrange(MAF) %>% 
  mutate(id=1:nrow(.))

plot(tab$id,tab$MAF,
     pch=20,
     xlab="SNP number",
     ylab="Minor allele frequencies")
```

`r nrow(tab[tab$MAF<1,])` SNPs have MAF < 1%.
`r nrow(tab[tab$MAF<5,])` SNPs have MAF < 5%.

```{r RmMissingDataMAF}
# We export the two datasets in a list.
maf_threshold <- 5
list_geno <- list(list_withmaf=list(df = df),
                  list_withoutmaf=list(df = df[!(row.names(df) %in% tab[tab$MAF<maf_threshold,]$snp),]))
```

In the dataset without MAF, we remove the SNPs with MAF < `r maf_threshold`%: there are `r nrow(list_geno$list_withoutmaf$df)` SNPs left.


### Missing data per SNP

```{r MissingDataPerSNP, warning=F}
identify_NA_perSNP <- function(x){
  
  x$df %>% 
    t() %>% 
    as_tibble() %>% 
    sapply(function(x) sum(is.na(x))*100/nrow(.)) %>% 
    as.data.frame() %>%
    set_colnames(c("na_freq")) %>% 
    rownames_to_column(var = "snp") %>% 
    as_tibble() %>% 
    arrange(na_freq) %>% 
    mutate(id=1:nrow(.))
}

NAfreq_per_SNP <- lapply(list_geno,identify_NA_perSNP)
```


In the dataset without MAF, the maximum percentage of missing data per SNP is `r max(NAfreq_per_SNP$list_withoutmaf$na_freq) %>% round(2)`%.

In the dataset with MAF, the maximum percentage of missing data per SNP is `r max(NAfreq_per_SNP$list_withmaf$na_freq) %>% round(2)`%.

Therefore, we do not apply a filter for the percentage of missing data per SNP and we keep all the SNPs.

```{r PlotMissingDataPerSNP, fig.height=6,fig.width=8}
plot(NAfreq_per_SNP$list_withoutmaf$id,NAfreq_per_SNP$list_withoutmaf$na_freq,
     pch=20,
     xlab="SNP number",
     ylab="Percentage of missing data per SNP",
     main="Genomic dataset without MAF")

plot(NAfreq_per_SNP$list_withmaf$id,NAfreq_per_SNP$list_withmaf$na_freq,
     pch=20,
     xlab="SNP number",
     ylab="Percentage of missing data per SNP",
     main="Genomic dataset with MAF")
```


After the filtering steps: 

  - <span style="color: red;">`r nrow(list_geno$list_withmaf$df)` SNPs and `r ncol(list_geno$list_withmaf$df)` individuals in the dataset not filtered for MAF.</span>

  - <span style="color: red;">`r nrow(list_geno$list_withoutmaf$df)` SNPs and `r ncol(list_geno$list_withoutmaf$df)` individuals in the dataset filtered for MAF.</span>
  
## Exporting the filtered datasets

When exporting the datasets, we add the population and family information.

In February 2025, the family assignment of some individuals were modified. We load the original family and populations assignments and update it with the new assignments.

```{r ExtractingAndMerginIndividualInformation}
# Original family and population assignement
field_data <- read_excel(here("data/ScotsPine/Field.xlsx"), na = "NA") %>% 
  dplyr::select(PopulationCode,Family, FieldCode) 

# Updated family assignement
fam_rea <- read_table(here("data/ScotsPine/GenomicData/data_update_Fev2025/raw_data/FamilyReassignment.txt"),show_col_types = FALSE)

# With the code below, we see that the "family" column of "Field.xlsx" and the "OriginalFamily" column of "FamilyReassignment.txt" are the same, which is necessary.
# field_data %>% 
#   left_join(fam_rea, by="FieldCode") %>%  
#   filter(OriginalFamily != AssignedFamily | OriginalFamily != Family | AssignedFamily != Family) 

# We are going to use the "AssignedFamily" column (as advised by Annika Perry)
field_data <- field_data %>% 
  left_join(fam_rea, by="FieldCode") %>% 
  dplyr::select(-Family, -OriginalFamily) %>% 
  dplyr::rename(Family = AssignedFamily) %>% 
  mutate(FieldCode = as.character(FieldCode))
```

We format the two datasets with SNPs in columns and individuals in rows. And we add two columns with information about the family and the population of each individual. 

We show below the first 10 rows and columns of the dataset not filtered for MAF.

```{r FormattingFilteredDatasets}
list_geno <- lapply(list_geno, function(x){
  
  # we invert the dataset with tidyverse (quite long)
  df <- x$df %>% 
  rownames_to_column("SNP") %>%
  pivot_longer(-SNP, names_to = "FieldCode", values_to = "Value") %>%
  pivot_wider(names_from = SNP, values_from = Value)

  # we merge the two datasets
  field_data %>% right_join(df, by="FieldCode")
  
})

list_geno$list_withmaf[1:10,1:10] %>% kable_mydf()
```


### Allele counts

We export allele counts.

```{r SavingFilteredDatasets, eval =F}
# we save the dataset not filtered for MAF
saveRDS(list_geno$list_withmaf, here("data/ScotsPine/GenomicData/data_update_Fev2025/formatted_data/filtered_allele_counts_withMAF.rds"))

# we save the dataset without MAF
saveRDS(list_geno$list_withoutmaf, here("data/ScotsPine/GenomicData/data_update_Fev2025/formatted_data/filtered_allele_counts_withoutMAF.rds"))
```

### Allele frequencies

We calculate the population allele frequencies.

```{r CalculatePopulationAlleledFrequencies, eval = F}
# Function to calculate the population allele frequencies
compute_allele_freq <- function(x){
  x %>% 
    dplyr::select(PopulationCode, contains("AX")) %>% 
    group_by(PopulationCode) %>% 
    summarise_all(~sum(., na.rm = TRUE)/((n()-sum(is.na(.)))*2))
}
```

```{r ExportPopAlleleFrequencies, eval =F}
# For the dataset with MAF
list_geno$list_withmaf %>% 
  compute_allele_freq %>% 
  saveRDS(here("data/ScotsPine/GenomicData/data_update_Fev2025/formatted_data/filtered_allele_freq_withMAF.rds"))

# For the dataset without MAF
list_geno$list_withoutmaf %>% 
  compute_allele_freq %>% 
  saveRDS(here("data/ScotsPine/GenomicData/data_update_Fev2025/formatted_data/filtered_allele_freq_withoutMAF.rds"))
```


# Imputation of missing data

We impute missing data based on the most common genotype within the family.

We first impute missing data in the dataset not filtered for MAF.

```{r ImputingNAbasedOnFamilies}
# we load filtered genomic data not filtered for MAF
df <- readRDS(here("data/ScotsPine/GenomicData/data_update_Fev2025/formatted_data/filtered_allele_counts_withMAF.rds"))

# Function to calculate mode (i.e., most common value)
Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

# Replace NA values with the mode within each family for each SNP
impdf <- df %>%
  group_by(Family) %>%
  mutate(across(starts_with("AX"), ~replace(., is.na(.), Mode(.[!is.na(.)])))) %>% 
  ungroup()
```

But, there are `r sum(is.na(impdf))` NAs left. Here are the two families and populations they belong to.

```{r RowsAndColumsWithNAsAfterImputation}
impdf %>%  
  dplyr::select(PopulationCode, FieldCode, Family, where(~ any(is.na(.)))) %>% # Select columns with NA values
  filter(if_any(everything(), is.na)) %>% # selects rows with NA values
  kable_mydf
```

We look at the counts of each allele in the two families with NAs remaining after imputation.

```{r AlleleCountsPopsWithNAsAfterImputation}
impdf %>%  dplyr::select(PopulationCode, FieldCode, Family, where(~ any(is.na(.)))) %>% # Selects columns where there are NA values
  filter(PopulationCode %in% c("CR", "GT")) %>%
  pivot_longer(cols = starts_with("AX"), names_to = "SNP", values_to = "allele_value") %>% 
  group_by(PopulationCode, allele_value) %>% 
  summarize(allele_count = n(), .groups = "drop") %>% 
  filter(allele_value %in% c(0, 1, 2)) %>%  # Include only counts for 0, 1, and 2
  kable_mydf()
```

Based on this, we attribute the value of 0 to the allele `AX-388239058` in family `GT4` and to the allele `AX-388336314` in family `CR8`.

```{r ImputationForRemaingNAs}
impdf <- impdf %>%
  mutate(
    `AX-388239058` = case_when(
      Family == "GT4" ~ 0,  # Assign 0 when Family is GT4
      TRUE ~ `AX-388239058` # Retain original value otherwise
    ),
    `AX-388336314` = case_when(
      Family == "CR8" ~ 0,  # Assign 0 when Family is CR8
      TRUE ~ `AX-388336314` # Retain original value otherwise
    )
  )

# We save the dataset
saveRDS(impdf, here("data/ScotsPine/GenomicData/data_update_Fev2025/formatted_data/imputed_allele_counts_withMAF.rds"))
```

We then subset the dataset by keeping only SNPs without MAF (to get an imputed dataset without MAF).

```{r SavingImputedDatasetWithoutMAF}
impdf <- readRDS(here("data/ScotsPine/GenomicData/data_update_Fev2025/formatted_data/imputed_allele_counts_withMAF.rds"))

# Extract SNPs filtered for MAF
snps_dfnomaf <- readRDS(here("data/ScotsPine/GenomicData/data_update_Fev2025/formatted_data/filtered_allele_counts_withoutMAF.rds")) %>% 
  dplyr::select(-PopulationCode,-Family, -FieldCode) %>% 
  colnames()

# Save dataset filtered for MAF and with imputed missing data
impdf %>% 
  dplyr::select(PopulationCode, Family, FieldCode, any_of(snps_dfnomaf)) %>% 
  saveRDS(here("data/ScotsPine/GenomicData/data_update_Fev2025/formatted_data/imputed_allele_counts_withoutMAF.rds"))
```


# Genetic relatedness coefficients

We use the genomic data not filtered for MAF and with no imputation of missing data to estimate the genetic relatedness coefficients $r_{ij}$. Assuming no inbreeding, $r_{ij}$ can be interpreted as the proportion of the genome that is shared between individuals $i$ and $j$. 

```{r EstimateGRM, eval=F}
# we load genomic data not filtered for MAF and without imputation for missing data
df <- readRDS(here("data/ScotsPine/GenomicData/data_update_Fev2025/formatted_data/filtered_allele_counts_withMAF.rds"))

# change the dataset to a matrix
mat <- df %>% 
  dplyr::select(-Family, -PopulationCode) %>% 
  column_to_rownames("FieldCode") %>% 
  as.matrix()

# We estimate the GRM with the VanRaden method
Gmat_VanRaden <- Gmatrix(SNPmatrix=mat, 
                         missingValue=NA, 
                         method="VanRaden")

# We save the G matrix
saveRDS(Gmat_VanRaden, here("data/ScotsPine/GenomicData/data_update_Fev2025/formatted_data/Gmat_VanRaden.rds"))
```

We extract the genetic relatedness coefficients for each pair of individuals. 

In the tutorial [GWAS 5: Relatedness and population structure by 
Matti Pirine (University of Helsinki)](https://www.mv.helsinki.fi/home/mjxpirin/GWAS_course/material/GWAS5.html), the authors refer to cutting points to assign pairs of individuals to a certain degree of relatedness based on their $r_{ij}$. These thresholds are "a commonly-used inference procedure of the KING software" and are based on the powers of $1/2$. We will use these thresholds to assign the pairs of individuals to a given degree of relatedness (i.e., identical, 1st degree, 2nd degree, 3rd degree or unrelated). 


```{r ExtractGeneticRelatednessCoeffs, fig.height=6, fig.width=8, warning=FALSE}
# we load the GRM
Gmat_VanRaden <- readRDS(here("data/ScotsPine/GenomicData/data_update_Fev2025/formatted_data/Gmat_VanRaden.rds"))

# viz the matrix
Gmat_VanRaden[1:10,1:10] %>% 
  as.data.frame() %>% 
  kable_mydf(boldfirstcolumn = TRUE, round_number = 4)

# Function to extract the genetic relatedness coefficients in a vector
extract_relatedness_coeffs_from_GRM <- function(GRM){
# Extract upper triangle of the matrix
upper_triangle <- upper.tri(GRM, diag=TRUE)

# Set lower triangle elements to NA
GRM[!upper_triangle] <- NA

# Convert the resulting matrix to a dataframe
df <- as.data.frame(as.table(GRM))

# Remove rows with NA values (optional if you want to remove NA values)
df <- na.omit(df)

# Rename the columns
names(df) <- c("Ind1", "Ind2", "r")

df
}


# Run the function on our GRM
df_r <- extract_relatedness_coeffs_from_GRM(GRM=Gmat_VanRaden) %>% arrange(r)

# Define the thresholds to separate the degrees of relatedness
KING_thresholds <- c(0.088, 0.177, 0.354, 0.707)

# We count the number of individual pairs within each degree of relatedness 
r_counts <- df_r %>%
  filter(!Ind1==Ind2) %>% # we remove the genetic relatedness coeffs calculated by comparing the same individuals
  summarize(
    identical = sum(r > KING_thresholds[[4]]),
    first_degree = sum(r <= KING_thresholds[[4]] & r > KING_thresholds[[3]]),
    second_degree = sum(r <= KING_thresholds[[3]] & r > KING_thresholds[[2]]),
    third_degree = sum(r <= KING_thresholds[[2]] & r > KING_thresholds[[1]]),
    unrelated = sum(r <= KING_thresholds[[1]])
  )

# we prepare the dataset for the plot 
dfplot <- df_r %>% 
  filter(!Ind1==Ind2) %>% # 
  filter(r>0.05) %>% # to aid visualization and speed up plotting (it was long if I keep all the pairs)
  mutate(Index=1:nrow(.))

# Define the breaks and labels for the y-scale
y_breaks <- c(0, 0.25, 0.5, 0.75, 1, KING_thresholds)
y_labels_colors <- ifelse(y_breaks %in% KING_thresholds, "orange", "black")

# Plot
ggplot(dfplot, aes(x = Index, 
                   y = r, 
                   color = cut(r, breaks = c(-Inf, KING_thresholds, Inf),
                               labels = c(paste0("Unrelated (n = ",r_counts$unrelated,")"), 
                                          paste0("3rd degree (n = ",r_counts$third_degree,")"), 
                                          paste0("2nd degree (n = ",r_counts$second_degree,")"), 
                                          paste0("1st degree (n = ",r_counts$first_degree,")"), 
                                          paste0("Identical (n = ",r_counts$identical,")"))))) +
  geom_point() +
  geom_hline(yintercept = KING_thresholds, linetype = "dashed", color = "orange") +
  labs(x = "Pairs of individuals", y = "Genetic relatedness coefficient") +
  theme_bw() +
  scale_color_manual(values = c("blue", "green", "deeppink", "purple", "yellow")) +
   scale_y_continuous(breaks = y_breaks) +
  theme(axis.text.y = element_text(angle = 0, hjust = 1, colour = y_labels_colors),
        legend.position = c(0.2,0.85),
        legend.title = element_blank())
```

In our dataset, we expect trees to be unrelated and that trees from the same family are half-siblings (with $0.177 > r_{ij} > 0.354$).

We look at the field data of the trees with higher $r_{ij}$ than expected.

```{r LoadFieldData}
# We load data with field and nursery information (that we merge with the updated family assignement)
field_data <- readRDS(file=here("data/ScotsPine/formatted_CG_data.rds")) %>% 
  dplyr::select(PopulationCode, FieldCode, FieldSite, Block, Nursery) %>% 
  mutate(FieldCode = as.character(FieldCode)) %>% 
  right_join(field_data, by=c("FieldCode", "PopulationCode"))
```

```{r FunctionShowFieldData}
# function to merge field data of the two individuals of each pair
show_fielddata <- function(r_threshold_min,r_threshold_max){
  
  subdf <-  dfplot %>% 
    filter(r>r_threshold_min) %>% 
    filter(r<r_threshold_max)
    
  lapply(c("Ind1","Ind2"), function(x) {
    subdf %>% 
      dplyr::select(any_of(x)) %>% 
      dplyr::rename(FieldCode=any_of(x)) %>% 
      left_join(field_data, by ="FieldCode") %>% 
      set_colnames(paste0(colnames(.),"_",x))
}) %>% 
    bind_cols() %>% 
    mutate(RelatednessCoeff = subdf$r)
}
```


We first look at pairs of individuals considered to be identical (=clones), i.e. with $r_{ij}>0.707$.

```{r ShowPairsIdenticalIndividuals}
show_fielddata(r_threshold_min = KING_thresholds[[4]],
               r_threshold_max = 1.1) %>% 
    kable_mydf
```

For the eight pairs of individuals with $r_{ij}>0.707$ (i.e., individuals considered as identical = clones), individuals come from the same families but not always the same field site and nurseries.

We then look at the `r r_counts$first_degree` pairs of individuals considered to be full-siblings (i.e., with $0.354>r_{ij}>0.707$). We do not plot all the pairs but only those for which the two individuals do not belong to the same family. 

```{r ShowPairsFirstDegreeRelatedIndividuals}
show_fielddata(r_threshold_min = KING_thresholds[[3]],
               r_threshold_max = KING_thresholds[[4]]) %>% 
  filter(!Family_Ind1 == Family_Ind2) %>% 
  kable_mydf()
```


<!-- We export the pairs of individuals with $r_{ij}>0.088$.  -->

<!-- ```{r ExportingRelatedIndividuals} -->
<!-- lapply(c("Ind1","Ind2"), function(x) { -->
<!--   dfplot %>%  -->
<!--     filter(r>KING_thresholds[[1]]) %>%  -->
<!--     dplyr::select(any_of(x)) %>%  -->
<!--     dplyr::rename(FieldCode=any_of(x)) %>%  -->
<!--     left_join(field_data, by ="FieldCode") %>%  -->
<!--     set_colnames(paste0(colnames(.),"_",x)) -->
<!-- }) %>%  -->
<!--   bind_cols() %>%  -->
<!--   mutate(RelatednessCoeff = dfplot %>% filter(r>KING_thresholds[[1]]) %>% .$r) %>%  -->
<!--   write.csv(here("data/ScotsPine/RelatedIndividuals.csv"), row.names = FALSE) -->
<!-- ``` -->


# SNPs positions on a reference genome

The files 'snpInfo420k_PR_010623.txt' and 'snpInfo_PR_010623.txt' contains genetic map locations in the *P. tabuliformis* genome for the PiSy420k and PiSy50k arrays, respectively. This is from Tanja Pyhajarvi’s group at the University of Helsinki, and mostly the work of Pasi Rastas.

We show the first rows of 'snpInfo420k_PR_010623.txt' below.

```{r SNPpositionFile, message=F, warning=F}
map_snp_420k <- read_delim(here("data/ScotsPine/GenomicData/SNPpositions/snpInfo420k_PR_010623.txt"), 
                           delim = "\t", 
                           escape_double = FALSE, 
                           trim_ws = TRUE,
                           show_col_types = F)
map_snp_50k <- read_delim(here("data/ScotsPine/GenomicData/SNPpositions/snpInfo_PR_010623.txt"), 
                          delim = "\t", 
                          escape_double = FALSE, 
                          trim_ws = TRUE,
                          show_col_types = F)

# sapply(map_snp_420k, function(x) sum(is.na(x)))

map_snp_420k[1:10,] %>% kable_mydf(boldfirstcolumn = T, font_size = 10)
```

We review the number of SNPs genotyped in Scots pine and selected for further analyses in the previous sections, which can be found in these two files:

  - `r map_snp_420k %>% dplyr::filter(SNP %in% colnames(list_geno$list_withmaf[,-c(1:3)])) %>% nrow()` SNPs from the dataset *with* MAF were found in 'snpInfo420k_PR_010623.txt'.
   
  - `r map_snp_420k %>% dplyr::filter(SNP %in% colnames(list_geno$list_withoutmaf[,-c(1:3)])) %>% nrow()` SNPs from the dataset *without* MAF were found in 'snpInfo420k_PR_010623.txt'.
    
  - `r map_snp_50k %>% dplyr::filter(SNP %in% colnames(list_geno$list_withmaf[,-c(1:3)])) %>% nrow()` SNPs from the dataset *with* MAF were found in 'snpInfo50k_PR_010623.txt'.  
    
  - `r map_snp_50k %>% dplyr::filter(SNP %in% colnames(list_geno$list_withoutmaf[,-c(1:3)])) %>% nrow()` SNPs from the dataset *without* MAF were found in 'snpInfo50k_PR_010623.txt'.


`r setdiff(colnames(list_geno$list_withmaf[,-c(1:3)]), map_snp_420k$SNP) %>% length()` SNPs (from the dataset with MAF) are not in the file 'snpInfo420k_PR_010623.txt', and `r setdiff(colnames(list_geno$list_withmaf[,-c(1:3)]), map_snp_50k$SNP) %>% length()` are not in the file 'snpInfo50k_PR_010623.txt'.

Below we use the 'snpInfo420k_PR_010623.txt' file.

We  look at the number of SNPs in each linkage group (`LG` column) among the SNPs genotyped in Scots pine (the SNPs selected in the previous sections).

```{r NumberSNPsLG}
lapply(list_geno,function(x){
  
  map_snp_420k %>% 
    filter(SNP %in% colnames(x[,-c(1:3)])) %>% 
    count(LG) 
  
}) %>% 
  setNames(c("Dataset with MAF","Dataset without MAF")) %>% 
  bind_rows(.id="Dataset") %>% 
  pivot_wider(names_from = "Dataset", values_from = "n") %>% 
  kable_mydf()
```

For each linkage group, we look at the number of SNPs assigned to each chromosome (`TAB_CONTIG` column). For most linkage groups, SNPs are mostly assigned to one chromosome, except for the linkage groups 2 and 3 for which the SNPs are almost equally well assigned to chromosome 1 and 3.

```{r NumberSNPsChromosome}
lapply(list_geno,function(x){
  
 map_snp_420k %>% 
    filter(SNP %in% colnames(x[,-c(1:3)])) %>% 
    count(LG, TAB_CONTIG)
  
}) %>% 
  setNames(c("Dataset with MAF","Dataset without MAF")) %>% 
  bind_rows(.id="Dataset") %>% 
  pivot_wider(names_from = "Dataset", values_from = "n") %>% 
  kable_mydf()
```

To identify SNP located nearby on the genome, we first identify SNPs on the same chromosome (based on the `TAB_CONTIG` column) and then determine their proximity using the `TAB_POS` column.

Below we plot the genome positions of the SNPs without MAF.

```{r SNPmappingFigure, warning=F}
# Formatting gen_coord for the figure
gen_coord <- map_snp_420k %>%
  filter(SNP %in% colnames(list_geno$list_withmaf[,-c(1:3)])) %>% 
  filter(TAB_CONTIG %in% paste0("chr", 1:12))
# sapply(gen_coord, function(x) sum(is.na(x))) # no NAs in the TAB_CONTIG and TAB_POS columns


gen_coord %>%
  ggplot() +
  # All SNPs of the reference genome 
  geom_segment(aes(x = TAB_POS, xend = TAB_POS, y = 0, yend = 0.2), col = "blue", size = 0.06, alpha=0.3) + 
  ylim(0,0.5) +
  facet_wrap(~TAB_CONTIG, scales = "free_x", ncol=3) +
  xlab("SNP positions on the reference genome") +
  ylab("") +
  theme_bw(base_size = 11) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "right",
    legend.background = element_blank(),
    panel.grid = element_blank(),
    legend.box.background = element_blank(),
    plot.background = element_blank(),
    panel.background = element_blank(),
    legend.text = element_text(size = rel(0.8)),
    strip.text = element_text(size = 11)
  )
```

